---
title: "indoor_environment"
author: "Souilla Luc"
date: "2026-03-04"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
encoding: UTF-8
---
<!--
================================================================================
SCRIPT:  PRG/Environmental/indoor_environment_hourly.Rmd
PURPOSE: Data preparation pipeline for indoor environmental data (temperature and relative humidity) collected at minute resolution via environmental nodes during the Montreal heatwave 2024. This script:
           1. Assigns participant IDs using baseline_nodes lookup
           2. Assigns heatwave status and filters to study window
           3. Aggregates minute-level data to hourly means
           4. Flags and removes implausible values (visual then definitive)
           5. Applies participant-specific correction for node malfunction
           6. Re-aggregates on cleaned data
           7. Summarizes data completeness per participant
           8. Exports the clean hourly dataset

INPUT FILES (DAT/):
  - environmental_minute.csv        -> raw minute-level indoor sensor data
  - baseline_nodes.csv              -> participant-nodeID-date lookup (from Data_preparation.Rmd)

OUTPUT FILES (DAT/Processed/):
  - indoor_environment_hourly_processed.csv -> cleaned hourly indoor data

SENSOR NOTE:
  Indoor temperature and relative humidity are derived primarily from the BME680 sensor, which was present in all nodes. A subset of nodes also included either a SHT41 or SHT85 sensor. Due to data quality issues encountered with the SHT41 and SHT85 sensors across participants, the BME680 was retained as the primary sensor for all analyses. SHT41/SHT85 readings are preserved in the dataset as secondary variables for reference.

IMPLAUSIBLE VALUE THRESHOLDS:
  Visual flagging (for inspection only, not removal):
    temp_mean < 10 or > 34 deg C  -- used to identify anomalous episodes rh_mean   < 5  or > 90 %

  Definitive removal (applied to raw minute data before re-aggregation):
    sensors.BME680.temp < 10 or > 50 deg C
    sensors.BME680.rh   < 5  or > 90 %

The two-threshold approach is intentional: the tighter visual flag (> 34) was used to identify P19's node malfunction episode (see P19 NOTE below), while the definitive removal threshold (> 50) is the physiologically implausible bound applied uniformly across all participants.

P19 NODE MALFUNCTION NOTE:
Participant P19's environmental node experienced a malfunction episode during which relative humidity readings exceeded 90% (abnormally high), accompanied by simultaneous anomalous temperature readings fixed at ~34.6. Visual inspection confirmed these were sensor artefacts, not real environmental conditions. Both temp and rh values are set to NA for all minutes where P19's rh >= 90%, removing the entire malfunction episode.

HEATWAVE PERIODS (consistent with assign_heatwave()):
  Before heatwave : 2024-06-11 to 2024-06-17
  During heatwave : 2024-06-18 to 2024-06-20
  After heatwave  : 2024-06-21 to 2024-06-27
================================================================================
-->

```{r setup, include=FALSE}
# Global knitr options: show code, suppress messages and warnings
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE
)
```

# Overview

This script prepares the **indoor environment** dataset from minute-level temperature and relative humidity measurements collected via environmental nodes placed in participants' homes during the Montreal Summer 2024 heatwave study.

Data from the BME680 sensor is used as the primary source for all participants.
Minute-level data is first aggregated to hourly means for visual quality inspection, then cleaned at the raw minute level before final re-aggregation.

---

# Environment setup

## Load packages
```{r load-packages, message=FALSE, warning=FALSE, include=FALSE}
required_packages <- c(
  "here", "dplyr", "tidyr", "lubridate",
  "ggplot2", "patchwork", "viridis", "openxlsx"
)

install_if_missing <- function(packages) {
  missing <- packages[!(packages %in% installed.packages()[, "Package"])]
  if (length(missing) > 0) install.packages(missing)
}

install_if_missing(required_packages)
invisible(lapply(required_packages, library, character.only = TRUE))
```

## Set project root and directory paths

```{r set-paths}
here::i_am("PRG/Environmental/indoor_environment_hourly.Rmd")

path_dat <- here("DAT")           # Raw data folder
path_prg <- here("PRG")           # Scripts folder
path_res <- here("DAT/Processed") # Output folder for processed data
```

## Source custom functions

```{r source-functions}
source(here("PRG/Functions/functions_participant_period.R"))
```

---

# Step 1 - Load raw data and assign participant IDs {#step1}

## Load raw minute-level indoor environment data

```{r IndEnvH-upload}
# Load raw minute-level indoor sensor data
# local_time is created by assign_study_period() from the unix column
indoor_environment_raw <- read.csv(
  file.path(path_dat, "environmental_minute.csv"),
  header = TRUE, sep = ",", quote = "\"'", dec = "."
)
```

## Load node-participant lookup table
```{r load-baseline-nodes}
# baseline_nodes maps each node macID + date range to a participant_id
# Uses environmental nodes (not smartwatch bangles)
# Created by Data_preparation.Rmd -- must be run first
baseline_nodes <- read.csv(
  file.path(path_dat, "baseline_nodes.csv"),
  header = TRUE, sep = ",", quote = "\"'", dec = "."
)
```
## Assign study period and participant ID
```{r IndEnvH-assign-participant}
# Step 1a: assign_study_period() creates local_time and period columns
# from the unix timestamp in the raw data
indoor_environment_participant <- assign_study_period(
  indoor_environment_raw, c("unix")
)

# Step 1b: Match each row to a participant_id using node macID + local_time
# Note: baseline_nodes is used here (not baseline_long) because indoor
# nodes have separate macIDs from smartwatches
indoor_environment_participant <- add_participant_id_watch(
  indoor_environment_participant,
  baseline_nodes,
  mac_col  = "macID",
  time_col = "local_time"
)
```

## Assign heatwave status and filter to study window

```{r IndEnvH-assign-heatwave}
# Assign heatwave status using standard function
indoor_environment_participant <- assign_heatwave(
  indoor_environment_participant
)

indoor_environment_participant <- indoor_environment_participant %>%
  filter(heatwave_status %in% c("BEF_HW_June", "HW_June", "AFT_HW_June"))

message("Minute-level records after heatwave filter: ",
        nrow(indoor_environment_participant))
```

---

# Step 2 - Initial hourly aggregation for QC {#step2}

Raw minute-level data is aggregated to hourly means for visual quality inspection. This is a preliminary aggregation on uncleaned data — a second aggregation on cleaned data is performed in Step 5.

```{r IndEnvH-aggregate-hourly-raw}
# Create hour-floor timestamp for grouping
indoor_environment_participant <- indoor_environment_participant %>%
  mutate(
    hour_floor = floor_date(as.POSIXct(local_time), unit = "hour")
  )

# Aggregate to hourly means using BME680 
indoor_environment_hourly_raw <- indoor_environment_participant %>%
  group_by(participant_id, hour_floor, date_converted, period, heatwave_status) %>%
  summarise(
    temp_mean    = mean(sensors.BME680.temp, na.rm = TRUE),
    temp_min     = min(sensors.BME680.temp,  na.rm = TRUE),
    temp_max     = max(sensors.BME680.temp,  na.rm = TRUE),
    temp_sd      = sd(sensors.BME680.temp,   na.rm = TRUE),
    rh_mean      = mean(sensors.BME680.rh,   na.rm = TRUE),
    rh_min       = min(sensors.BME680.rh,    na.rm = TRUE),
    rh_max       = max(sensors.BME680.rh,    na.rm = TRUE),
    rh_sd        = sd(sensors.BME680.rh,     na.rm = TRUE),
    n_obs        = n(),
    n_temp_valid = sum(!is.na(sensors.BME680.temp)),
    .groups = "drop"
  ) %>%
  # Replace Inf values produced by min/max on all-NA groups with NA
  mutate(
    across(c(temp_min, temp_max, rh_min, rh_max),
           ~ ifelse(is.infinite(.), NA, .))
  )

message("Hourly records (raw): ", nrow(indoor_environment_hourly_raw))
```

---

# Step 3 - Coverage heatmap {#step3}

Visualize the number of valid hourly measurements per participant per day to identify missing days or node failures.

```{r IndEnvH-coverage-heatmap, echo=FALSE}
all_participants <- sort(unique(indoor_environment_hourly_raw$participant_id))
n_per_page       <- 4
n_pages          <- ceiling(length(all_participants) / n_per_page)

for (i in seq_len(n_pages)) {
  start_idx         <- (i - 1) * n_per_page + 1
  end_idx           <- min(i * n_per_page, length(all_participants))
  participants_page <- all_participants[start_idx:end_idx]

  hourly_subset <- indoor_environment_hourly_raw %>%
    filter(participant_id %in% participants_page)

  indoor_env_summary <- hourly_subset %>%
    group_by(participant_id, date_converted) %>%
    summarise(
      n_hours_valid = sum(!is.na(temp_mean)),
      .groups = "drop"
    )

  heatmap_indiv_day <- ggplot(
    indoor_env_summary,
    aes(x = date_converted, y = "", fill = n_hours_valid)
  ) +
    geom_tile(color = "grey") +
    geom_text(aes(label = ifelse(n_hours_valid == 0, "X", n_hours_valid)),
              size = 3, color = "white") +
    scale_fill_gradient(
      low    = "lightblue", high = "darkblue",
      limits = c(0, 24), breaks = c(0, 6, 12, 18, 24),
      name   = "Valid hours\nper day"
    ) +
    facet_wrap(~ participant_id, nrow = 2, ncol = 2) +
    labs(title = "Indoor environment -- valid hourly measurements per day",
         x = NULL, y = NULL) +
    theme_minimal() +
    theme(legend.position = "bottom",
          axis.text.x     = element_text(angle = 45, hjust = 1))
  print(heatmap_indiv_day)
}
```

---

# Step 4 - Visual flagging and QC plots {#step4}

## Flag implausible values for visual inspection

Two flagging thresholds are applied for visual inspection only:

- `temp_mean` < 10 or > 34 deg C
- `rh_mean` < 5 or > 90 %

The tighter upper temperature bound (34 deg C) was used specifically to identify P19's node malfunction episode during visual inspection. The definitive removal threshold applied in Step 5 is the physiologically implausible bound of 50 deg C — see header block for full rationale.


```{r IndEnvH-flag-visual}
indoor_env_data_subset_raw <- indoor_environment_hourly_raw %>%
  mutate(
    measurement_type = case_when(
      !is.na(temp_mean) & (temp_mean < 10 | temp_mean > 34) ~ "temp_flag",
      !is.na(rh_mean)   & (rh_mean   < 5  | rh_mean   > 90) ~ "rh_flag",
      TRUE                                                    ~ "Valid"
    ),
    measurement_type = factor(measurement_type,
                               levels = c("Valid", "temp_flag", "rh_flag"))
  )

flag_colors <- c("Valid"     = "#2E86AB",
                 "temp_flag" = "#F6AE2D",
                 "rh_flag"   = "#E63946")
flag_shapes <- c("Valid"     = 16,
                 "temp_flag" = 17,
                 "rh_flag"   = 4)
```

## Diagnostic summary of flagged values per participant

```{r IndEnvH-flag-summary}
# Add numeric flags for diagnostic summary
indoor_environment_hourly_raw <- indoor_environment_hourly_raw %>%
  mutate(
    temp_flag = case_when(
      is.na(temp_mean)                     ~ NA_real_,
      temp_mean < 10 | temp_mean > 34      ~ 1,
      TRUE                                 ~ 0
    ),
    rh_flag = case_when(
      is.na(rh_mean)                       ~ NA_real_,
      rh_mean < 5 | rh_mean > 90          ~ 1,
      TRUE                                 ~ 0
    )
  )

# Per-participant summary of flagged hours
indoor_environment_hourly_raw %>%
  group_by(participant_id) %>%
  summarise(
    n_temp_flagged = sum(temp_flag == 1, na.rm = TRUE),
    n_rh_flagged   = sum(rh_flag   == 1, na.rm = TRUE),
    n_both_flagged = sum(temp_flag == 1 & rh_flag == 1, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  filter(n_temp_flagged > 0 | n_rh_flagged > 0) %>%
  arrange(desc(n_both_flagged))
```


## Before filtering plots

```{r IndEnvH-plot-before, echo=FALSE}
line_temp_pooled_flagged <- ggplot(
  indoor_env_data_subset_raw,
  aes(x = hour_floor, y = temp_mean)
) +
  geom_point(aes(color = measurement_type, shape = measurement_type),
             size = 1, alpha = 0.85) +
  scale_color_manual(values = flag_colors, name = "Type") +
  scale_shape_manual(values = flag_shapes, name = "Type") +
  labs(title = "Before filtering",
       x = "", y = "Average indoor temperature (deg C)") +
  theme_minimal() +
  theme(legend.position = "right",
        axis.text.x     = element_text(angle = 45, hjust = 1, size = 10))

line_rh_pooled_flagged <- ggplot(
  indoor_env_data_subset_raw,
  aes(x = hour_floor, y = rh_mean)
) +
  geom_point(aes(color = measurement_type, shape = measurement_type),
             size = 1, alpha = 0.85) +
  scale_color_manual(values = flag_colors, name = "Type") +
  scale_shape_manual(values = flag_shapes, name = "Type") +
  labs(title = "Before filtering",
       x = "", y = "Average indoor relative humidity (%)") +
  theme_minimal() +
  theme(legend.position = "right",
        axis.text.x     = element_text(angle = 45, hjust = 1, size = 10))

print(line_temp_pooled_flagged)
print(line_rh_pooled_flagged)
```

---

# Step 5 - Clean raw minute data and re-aggregate {#step5}

Following visual inspection, implausible values are removed at the **raw minute level** before re-aggregation. This approach is more conservative than removing at the hourly level because it prevents a single invalid minute from contaminating an otherwise valid hourly mean.

Definitive removal thresholds:

| Variable | Lower | Upper |
|---|---|---|
| sensors.BME680.temp | < 10 deg C | > 50 deg C |
| sensors.BME680.rh   | < 5 %      | > 90 %     |

Values are set to NA at the variable level (not row deletion) to preserve valid concurrent measurements from the same node.

**P19 node malfunction:** Visual inspection revealed that P19's node experienced a malfunction episode characterized by rh >= 90% accompanied by simultaneous anomalous temperature readings. Both variables are set to NA for all minutes where P19's rh >= 90% to remove the entire malfunction episode. This participant-specific correction is justified by device malfunction confirmed during visual inspection.

```{r IndEnvH-clean-raw}
indoor_environment_participant <- indoor_environment_participant %>%
  mutate(
    # Definitive removal: physiologically implausible bounds
    sensors.BME680.temp = case_when(
      sensors.BME680.temp < 10 ~ NA_real_,  # Below 10 deg C -- implausible indoors
      sensors.BME680.temp > 50 ~ NA_real_,  # Above 50 deg C -- sensor error
      TRUE                     ~ sensors.BME680.temp
    ),
    sensors.BME680.rh = case_when(
      sensors.BME680.rh < 5   ~ NA_real_,   
      sensors.BME680.rh > 90  ~ NA_real_,  
      TRUE                    ~ sensors.BME680.rh
    ),
    # P19 node malfunction correction:
    # When rh >= 90 for P19, both temp and rh are set to NA
    # because the malfunction affects both variables simultaneously
    sensors.BME680.temp = ifelse(
      participant_id == "P19" & sensors.BME680.rh >= 90,
      NA_real_, sensors.BME680.temp
    ),
    sensors.BME680.rh = ifelse(
      participant_id == "P19" & sensors.BME680.rh >= 90,
      NA_real_, sensors.BME680.rh
    )
  )
```

## Re-aggregate to hourly on cleaned data

```{r IndEnvH-aggregate-hourly-clean}
# Re-aggregate using the same grouping as Step 2 but on cleaned data
indoor_environment_hourly <- indoor_environment_participant %>%
  group_by(participant_id, hour_floor, date_converted, period, heatwave_status) %>%
  summarise(
    temp_mean    = mean(sensors.BME680.temp, na.rm = TRUE),
    temp_min     = min(sensors.BME680.temp,  na.rm = TRUE),
    temp_max     = max(sensors.BME680.temp,  na.rm = TRUE),
    temp_sd      = sd(sensors.BME680.temp,   na.rm = TRUE),
    rh_mean      = mean(sensors.BME680.rh,   na.rm = TRUE),
    rh_min       = min(sensors.BME680.rh,    na.rm = TRUE),
    rh_max       = max(sensors.BME680.rh,    na.rm = TRUE),
    rh_sd        = sd(sensors.BME680.rh,     na.rm = TRUE),
    n_obs        = n(),
    n_temp_valid = sum(!is.na(sensors.BME680.temp)),
    .groups = "drop"
  ) %>%
  mutate(
    across(c(temp_min, temp_max, rh_min, rh_max),
           ~ ifelse(is.infinite(.), NA, .))
  )

message("Hourly records (cleaned): ", nrow(indoor_environment_hourly))
```

---

# Step 6 - After filtering plots {#step6}
```{r IndEnvH-plot-after, echo=FALSE}
line_temp_pooled_after <- ggplot(
  indoor_environment_hourly,
  aes(x = hour_floor, y = temp_mean)
) +
  geom_point(color = "#2E86AB", size = 1, alpha = 0.85) +
  labs(title = "After filtering",
       x = "", y = "Average indoor temperature (deg C)") +
  theme_minimal() +
  theme(legend.position = "right",
        axis.text.x     = element_text(angle = 45, hjust = 1, size = 10))

line_rh_pooled_after <- ggplot(
  indoor_environment_hourly,
  aes(x = hour_floor, y = rh_mean)
) +
  geom_point(color = "#2E86AB", size = 1, alpha = 0.85) +
  labs(title = "After filtering",
       x = "", y = "Average indoor relative humidity (%)") +
  theme_minimal() +
  theme(legend.position = "right",
        axis.text.x     = element_text(angle = 45, hjust = 1, size = 10))

line_temp_pooled_flagged / line_temp_pooled_after +
  plot_annotation(title = "Effect of data cleaning on average indoor temperature")

line_rh_pooled_flagged / line_rh_pooled_after +
  plot_annotation(title = "Effect of data cleaning on average indoor relative humidity")
```

---

# Step 7 - Data completeness summary {#step7}

```{r IndEnvH-completeness-summary}
# Per-participant completeness summary
summary_per_participant <- indoor_environment_hourly %>%
  group_by(participant_id) %>%
  summarise(
    n_hours_total  = n(),
    n_temp_valid   = sum(!is.na(temp_mean)),
    pct_temp_valid = round(100 * mean(!is.na(temp_mean)), 1),
    n_rh_valid     = sum(!is.na(rh_mean)),
    pct_rh_valid   = round(100 * mean(!is.na(rh_mean)), 1),
    .groups = "drop"
  )
print(summary_per_participant, n = Inf)

# Overall completeness
summary_overall <- indoor_environment_hourly %>%
  summarise(
    n_hours_total  = n(),
    n_temp_valid   = sum(!is.na(temp_mean)),
    pct_temp_valid = round(100 * mean(!is.na(temp_mean)), 1),
    n_rh_valid     = sum(!is.na(rh_mean)),
    pct_rh_valid   = round(100 * mean(!is.na(rh_mean)), 1)
  )
message("Overall: ", summary_overall$n_hours_total, " hours | ",
        "Temp valid: ", summary_overall$pct_temp_valid, "% | ",
        "RH valid: ", summary_overall$pct_rh_valid, "%")
```


## Distribution of cleaned variables

```{r IndEnvH-distribution-boxplot, echo=FALSE}
indoor_environment_hourly %>%
  select(participant_id, temp_mean, temp_min, temp_max,
         rh_mean, rh_min, rh_max) %>%
  pivot_longer(
    cols      = -participant_id,
    names_to  = "variable",
    values_to = "value"
  ) %>%
  mutate(
    type     = ifelse(grepl("temp", variable),
                      "Temperature (deg C)", "Relative humidity (%)"),
    variable = factor(variable,
                      levels = c("temp_min", "temp_mean", "temp_max",
                                 "rh_min",   "rh_mean",   "rh_max"))
  ) %>%
  ggplot(aes(x = variable, y = value, fill = variable)) +
  geom_boxplot(outlier.colour = "red", outlier.size = 0.5, alpha = 0.5) +
  facet_wrap(~ type, scales = "free_y") +
  scale_fill_viridis_d() +
  labs(title = "Distribution of indoor temperature and humidity after cleaning",
       x = NULL, y = NULL) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "none",
        axis.text.x     = element_text(angle = 45, hjust = 1))
```

---

# Step 8 - Final dataset and export {#step8}

```{r IndEnvH-clean-final}
# Remove intermediate QC columns not needed for analysis
final_indoor_environment_hourly <- indoor_environment_hourly %>%
  select(-n_obs, -n_temp_valid)
```

```{r IndEnvH-export}
output_file <- file.path(path_res, "indoor_environment_hourly_processed.csv")
write.csv(final_indoor_environment_hourly, output_file, row.names = FALSE)
message("Exported: indoor_environment_hourly_processed.csv (",
        nrow(final_indoor_environment_hourly), " rows x ",
        ncol(final_indoor_environment_hourly), " columns)")
```

---

# Session information

```{r session-info}
sessionInfo()
```
