---
title: "Indoor Environment Daily HeatSuite 2024"
author: "Souilla Luc"
date: "2026-03-04"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
encoding: UTF-8
---

<!--
================================================================================
SCRIPT:  PRG/Environmental/indoor_environment_daily.Rmd
PURPOSE: Data preparation pipeline for indoor environmental data (temperature and relative humidity) collected via environmental nodes during the Montreal heatwave 2024 . This script:
           1. Assigns participant IDs to daily indoor environment records
           2. Flags and removes implausible temperature and humidity values
           3. Exports the clean final dataset and QC plots

INPUT FILES (DAT/):
  - environmental_daily.csv         -> daily indoor temp and RH per node
  - baseline_long.csv               -> participant-macID-date lookup
                                       (from Data_preparation.Rmd)

OUTPUT FILES (DAT/Processed/):
  - indoor_environment_processed.csv -> cleaned daily indoor environment data

IMPLAUSIBLE VALUE THRESHOLDS:
  Temperature : temp_avg < 4 or > 50 deg C
  Relative humidity : rh_avg < 1 or > 100 %

HEATWAVE PERIODS (consistent with assign_heatwave()):
  Before heatwave : 2024-06-11 to 2024-06-17
  During heatwave : 2024-06-18 to 2024-06-20
  After heatwave  : 2024-06-21 to 2024-06-27

================================================================================
-->

```{r setup, include=FALSE}
# Global knitr options: show code, suppress messages and warnings
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE
)
```

# Overview

This script prepares the **indoor environment** dataset from daily temperature and relative humidity measurements collected via environmental nodes placed in participants' homes during the Montreal Summer 2024 heatwave study.

***

# Environment setup

## Load packages

```{r load-packages, message=FALSE, warning=FALSE, include=FALSE}
required_packages <- c(
  "here", "dplyr", "tidyr", "lubridate",
  "ggplot2", "patchwork"
)

install_if_missing <- function(packages) {
  missing <- packages[!(packages %in% installed.packages()[, "Package"])]
  if (length(missing) > 0) install.packages(missing)
}

install_if_missing(required_packages)
invisible(lapply(required_packages, library, character.only = TRUE))
```


## Set project root and directory paths

```{r set-paths}
here::i_am("PRG/Environmental/indoor_environment_daily.Rmd")

path_dat <- here("DAT")           # Raw data folder
path_prg <- here("PRG")           # Scripts folder
path_res <- here("DAT/Processed") # Output folder for processed data
```

## Source custom functions

```{r source-functions}
source(here("PRG/Functions/functions_participant_period.R"))
```


***

# Step 1 - Load raw data and assign participant IDs {#step1}

## Load raw indoor environment data
```{r IndEnv-upload}
# Load daily indoor environment data from environmental nodes
indoor_environment_raw <- read.csv(
  file.path(path_dat, "environmental_daily.csv"),
  header = TRUE, sep = ",", quote = "\"'", dec = "."
)
```

## Load participant-macID lookup table

```{r load-baseline-long}
baseline_long <- read.csv(
  file.path(path_dat, "baseline_long.csv"),
  header = TRUE, sep = ",", quote = "\"'", dec = "."
)
```

## Assign participant ID and heatwave status

```{r IndEnv-assign-participant}
# Step 1a: assign_study_period() called with "temp_min_unix" because this
# dataset does not have a standard "unix" column.
# The function creates "temp_min_unix_time_converted" which is renamed below.
indoor_environment_participant <- assign_study_period(
  indoor_environment_raw, c("temp_min_unix")
)

# Step 1b: Create local_time from sampling_date for participant ID matching
indoor_environment_participant <- indoor_environment_participant %>%
  mutate(
    local_time = as.POSIXct(sampling_date, format = "%d/%m/%Y", tz = "UTC")
  )

# Step 1c: Match each row to a participant_id using macID + local_time
indoor_environment_participant <- add_participant_id_watch(
  indoor_environment_participant,
  baseline_long,
  mac_col  = "macID",
  time_col = "local_time"
)

# Step 1d: Rename time column for compatibility with assign_heatwave()
indoor_environment_participant <- indoor_environment_participant %>%
  rename(unix_time_converted = temp_min_unix_time_converted)

# Step 1e: Assign heatwave status and restrict to monitoring window
indoor_environment_participant <- assign_heatwave(indoor_environment_participant)

indoor_environment_participant <- indoor_environment_participant %>%
  filter(heatwave_status %in% c("BEF_HW_June", "HW_June", "AFT_HW_June"))
```


***

# Step 2 - Check measurement coverage {#step2}

Visualize the number of measurements per participant per day to identify
missing days or unexpected duplicates.



 
```{r IndEnv-coverage-heatmap, echo=FALSE}
all_participants <- sort(unique(indoor_environment_participant$participant_id))
n_per_page       <- 4
n_pages          <- ceiling(length(all_participants) / n_per_page)

for (i in seq_len(n_pages)) {
  start_idx         <- (i - 1) * n_per_page + 1
  end_idx           <- min(i * n_per_page, length(all_participants))
  participants_page <- all_participants[start_idx:end_idx]

  subset_page <- indoor_environment_participant %>%
    filter(participant_id %in% participants_page)

  indoor_env_summary <- subset_page %>%
    group_by(participant_id, date_converted) %>%
    summarise(n_measurements = n(), .groups = "drop")

  heatmap_indiv_day <- ggplot(
    indoor_env_summary,
    aes(x = date_converted, y = "", fill = n_measurements)
  ) +
    geom_tile(color = "grey") +
    scale_fill_gradient(
      low    = "lightblue",
      high   = "darkblue",
      limits = c(1, max(indoor_env_summary$n_measurements)),
      breaks = function(x) unique(floor(pretty(x))),
      name   = "Number of\nmeasurements"
    ) +
    facet_wrap(~ participant_id, nrow = 2, ncol = 2) +
    labs(title = "Indoor environment measurements per day",
         x = "", y = "Days of measurement") +
    theme_minimal() +
    theme(legend.position = "bottom")
  print(heatmap_indiv_day)
}
```

***

# Step 3 - Flag implausible values {#step3}

Implausible values are flagged based on physiologically and physically
plausible indoor ranges:

| Variable | Flag condition |
|---|---|
| temp_avg | < 4 deg C (below freezing, implausible indoors) or > 50 deg C (sensor error) |
| rh_avg | < 1% or > 100% (physically impossible) |



```{r IndEnv-flag-implausible}
indoor_environment_participant <- indoor_environment_participant %>%
  mutate(
    temp_flag = case_when(
      is.na(temp_avg)  ~ NA_real_,
      temp_avg < 4     ~ 1,   
      temp_avg > 50    ~ 1,   
      TRUE             ~ 0
    ),
    rh_flag = case_when(
      is.na(rh_avg)    ~ NA_real_,
      rh_avg < 1       ~ 1,   
      rh_avg > 100     ~ 1,   
      TRUE             ~ 0
    )
  )

# Summary of flagged values
message("=== Temperature implausible values ===")
message("N implausible: ",
        sum(indoor_environment_participant$temp_flag == 1, na.rm = TRUE))
message("% implausible: ",
        round(mean(indoor_environment_participant$temp_flag == 1,
                   na.rm = TRUE) * 100, 2), "%")

message("=== Relative humidity implausible values ===")
message("N implausible: ",
        sum(indoor_environment_participant$rh_flag == 1, na.rm = TRUE))
message("% implausible: ",
        round(mean(indoor_environment_participant$rh_flag == 1,
                   na.rm = TRUE) * 100, 2), "%")
```


## Visualize flagged values before removal

```{r IndEnv-plot-before, echo=FALSE}
indoor_env_data_subset <- indoor_environment_participant %>%
  mutate(
    measurement_type = case_when(
      temp_flag == 1 ~ "temp_flag",
      rh_flag   == 1 ~ "rh_flag",
      TRUE           ~ "Valid"
    ),
    measurement_type = factor(measurement_type,
                               levels = c("Valid", "temp_flag", "rh_flag"))
  ) %>%
  group_by(participant_id, date_converted) %>%
  mutate(n_measurements = n()) %>%
  ungroup()

flag_colors <- c("Valid"     = "#2E86AB",
                 "temp_flag" = "#F6AE2D",
                 "rh_flag"   = "#E63946")
flag_shapes <- c("Valid"     = 16,
                 "temp_flag" = 17,
                 "rh_flag"   = 4)

line_temp_pooled_flagged <- ggplot(
  indoor_env_data_subset,
  aes(x = date_converted, y = temp_avg)
) +
  geom_point(aes(color = measurement_type, shape = measurement_type),
             size = 2.5, alpha = 0.85) +
  scale_color_manual(values = flag_colors, name = "Type") +
  scale_shape_manual(values = flag_shapes, name = "Type") +
  labs(title = "Before filtering",
       x = "", y = "Average indoor temperature (deg C)") +
  theme_minimal() +
  theme(legend.position  = "right",
        axis.text.x      = element_text(angle = 45, hjust = 1, size = 10))
print(line_temp_pooled_flagged)

line_rh_pooled_flagged <- ggplot(
  indoor_env_data_subset,
  aes(x = date_converted, y = rh_avg)
) +
  geom_point(aes(color = measurement_type, shape = measurement_type),
             size = 2.5, alpha = 0.85) +
  scale_color_manual(values = flag_colors, name = "Type") +
  scale_shape_manual(values = flag_shapes, name = "Type") +
  labs(title = "Before filtering",
       x = "", y = "Average indoor relative humidity (%)") +
  theme_minimal() +
  theme(legend.position  = "right",
        axis.text.x      = element_text(angle = 45, hjust = 1, size = 10))
print(line_rh_pooled_flagged)
```


## Remove flagged values

```{r IndEnv-remove-flagged}
cleaned_indoor_env <- indoor_env_data_subset %>%
  filter(temp_flag == 0 & rh_flag == 0)

n_removed <- nrow(indoor_env_data_subset) - nrow(cleaned_indoor_env)
message("Rows removed (implausible temperature or humidity): ", n_removed)
```


## Visualize after removal

```{r IndEnv-plot-after, echo=FALSE}
line_temp_pooled_after <- ggplot(
  cleaned_indoor_env,
  aes(x = date_converted, y = temp_avg)
) +
  geom_point(color = "#2E86AB", size = 2.5, alpha = 0.85) +
  labs(title = "After filtering",
       x = "", y = "Average indoor temperature (deg C)") +
  theme_minimal() +
  theme(legend.position  = "right",
        axis.text.x      = element_text(angle = 45, hjust = 1, size = 10))

line_rh_pooled_after <- ggplot(
  cleaned_indoor_env,
  aes(x = date_converted, y = rh_avg)
) +
  geom_point(color = "#2E86AB", size = 2.5, alpha = 0.85) +
  labs(title = "After filtering",
       x = "", y = "Average indoor relative humidity (%)") +
  theme_minimal() +
  theme(legend.position  = "right",
        axis.text.x      = element_text(angle = 45, hjust = 1, size = 10))

line_temp_pooled_flagged / line_temp_pooled_after +
  plot_annotation(title = "Effect of data cleaning on average indoor temperature")

line_rh_pooled_flagged / line_rh_pooled_after +
  plot_annotation(title = "Effect of data cleaning on average indoor relative humidity")
```

***

# Step 4 - Final dataset and export {#step4}

## Remove intermediate processing columns

```{r IndEnv-clean-final}
final_indoor_environment_data <- cleaned_indoor_env %>%
  select(
    -temp_min_unix, -temp_max_unix, -rh_min_unix, -rh_max_unix,
    -local_time, -local_date,
    -temp_flag, -rh_flag,
    -measurement_type, -n_measurements, -sampling_date, -id, -period, -unix_time_converted
  )
```

## Export processed dataset

```{r IndEnv-export}
output_file <- file.path(path_res, "indoor_environment_processed.csv")
write.csv(final_indoor_environment_data, output_file, row.names = FALSE)
message("Exported: indoor_environment_processed.csv (",
        nrow(final_indoor_environment_data), " rows x ",
        ncol(final_indoor_environment_data), " columns)")
```

***

# Session information

```{r session-info}
sessionInfo()
```
