---
title: "Main Analysis Environment HeatSuite 2024"
author: "Souilla Luc"
date: "2026-03-05"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
encoding: UTF-8
---

<!--
================================================================================
SCRIPT:  PRG/Environmental/main_analysis_environment.Rmd
PURPOSE: Descriptive and inferential analysis of indoor and outdoor
         environmental data (temperature, relative humidity) collected
         during the Montreal Summer 2024 heatwave study. This script:
           1. Loads all processed environmental datasets
           2. Produces descriptive statistics tables by heatwave period
           3. Computes threshold exceedance summaries (26 and 31 deg C)
           4. Produces time series plots (pooled and by AC status)
           5. Runs linear mixed models on indoor environment
           6. Visualizes indoor temperature by building characteristics

INPUT FILES (DAT/Processed/):
  - indoor_environment_processed.csv        -> daily indoor temp and RH
  - outdoor_environment_processed.csv       -> daily outdoor temp and RH
  - indoor_environment_hourly_processed.csv -> hourly indoor temp and RH
  - outdoor_environment_hourly.csv          -> hourly outdoor temp and RH
  - characteristics_participants.csv        -> baseline participant data

OUTPUT FILES (RES/Environmental/):
  - Table/table_pooled_env_heatwave.xlsx    -> descriptive stats tables
  - Graph/building_characteristics_temperature.png
  - Graph/building_characteristics_temp_max.png
  - Graph/ (additional plots saved by custom functions)

STATISTICAL APPROACH:
  Mixed Model Repeated Measures (MMRM) with AR(1) covariance structure.
  Fixed effects tested with car::Anova() Type II (Kenward-Roger df).
  Post-hoc comparisons with emmeans() using mvt adjustment.
  Rationale: see narrative in Statistical Methods section below.

AC STATUS NOTE:
  Air conditioning status is derived from participant self-report at
  baseline. The list below was verified against characteristics_participants.csv and is hardcoded for reproducibility. If participants are added or removed, this list must be updated manually.

HEATWAVE PERIODS:
  Before heatwave : 2024-06-11 to 2024-06-17
  During heatwave : 2024-06-18 to 2024-06-20
  After heatwave  : 2024-06-21 to 2024-06-27

================================================================================
-->

```{r setup, include=FALSE}
# Global knitr options: show code, suppress messages and warnings
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE
)
```

# Overview

This script performs descriptive and inferential analyses of indoor and
outdoor environmental conditions during the Montreal Summer 2024 heatwave.
Analyses include descriptive statistics by heatwave period, threshold
exceedance summaries, time series visualizations, linear mixed models,
and exploratory plots of building characteristics.

---

# Environment setup

## Load packages

```{r load-packages, message=FALSE, warning=FALSE, include=FALSE}
required_packages <- c(
  # Data wrangling
  "here", "dplyr", "tidyr", "lubridate", "zoo", "stringr", "purrr",
  # Reporting
  "knitr", "kableExtra", "openxlsx",
  # Visualization
  "ggplot2", "patchwork", "viridis", "ggeffects",
  # Statistical modelling
  "rstatix", "carData", "car", "emmeans", "mmrm",
  "mlogit", "ordinal", "glmmTMB"
)

install_if_missing <- function(packages) {
  missing <- packages[!(packages %in% installed.packages()[, "Package"])]
  if (length(missing) > 0) install.packages(missing)
}

install_if_missing(required_packages)
invisible(lapply(required_packages, library, character.only = TRUE))
```

## Set project root and directory paths

```{r set-paths}
here::i_am("PRG/Environmental/main_analysis_environment.Rmd")

path_dat <- here("DAT/Processed")       # Processed data inputs
path_prg <- here("PRG")                 # Scripts folder
path_res <- here("RES/Environmental")   # Results output folder
```


## Set global color palette

```{r set-color-palette, echo=FALSE, warning=FALSE}
# Override ggplot2 discrete scales globally with viridis (color-blind friendly)
scale_color_discrete <- function(...) scale_color_viridis_d(...)
scale_fill_discrete  <- function(...) scale_fill_viridis_d(...)
```

## Source custom analysis functions

```{r source-functions}
# All custom functions used in this script
# See each .R file in PRG/Functions/ for full documentation
source(here("PRG/Functions/functions_cleaning_data.R"))
source(here("PRG/Functions/functions_descriptive_stats.R"))  # descriptive_stats(), display_table()
source(here("PRG/Functions/functions_descriptive_plot.R"))   # time_series_ functions
source(here("PRG/Functions/functions_linear_models.R"))      # lmm_one_way(), lmm_one_way_AC(), lmm_recovery()
source(here("PRG/Functions/residual_check_mmrm.R"))
source(here("PRG/Functions/functions_check_clmm.R"))
```

---

# Load processed data {#load-data}

```{r load-data}
# Daily indoor environment (temperature and relative humidity)
final_indoor_environment_data <- read.csv(
  file.path(path_dat, "indoor_environment_processed.csv"),
  header = TRUE, sep = ",", quote = "\"'", dec = "."
)

# Daily outdoor environment (temperature and relative humidity)
final_outdoor_environment_data <- read.csv(
  file.path(path_dat, "outdoor_environment_processed.csv"),
  header = TRUE, sep = ",", quote = "\"'", dec = "."
)

# Hourly indoor environment
final_indoor_environment_hourly <- read.csv(
  file.path(path_dat, "indoor_environment_hourly_processed.csv"),
  header = TRUE, sep = ",", quote = "\"'", dec = "."
)

# Hourly outdoor environment
final_outdoor_environment_hourly <- read.csv(
  file.path(path_dat, "outdoor_environment_hourly.csv"),
  header = TRUE, sep = ",", quote = "\"'", dec = "."
)

# Baseline participant characteristics (for building characteristics section)
final_baseline <- read.csv(
  file.path(path_dat, "characteristics_participants.csv"),
  header = TRUE, sep = ",", quote = "\"'", dec = "."
)
```

## Convert hourly timestamps to POSIXct

```{r convert-hourly-timestamps}

final_indoor_environment_hourly <- final_indoor_environment_hourly %>%
  mutate(hour_floor = lubridate::ymd_hms(hour_floor, truncated = 3))

final_outdoor_environment_hourly <- final_outdoor_environment_hourly %>%
  mutate(hour_floor = lubridate::ymd_hms(hour_floor, truncated = 3))
```

---

# Define analysis variable lists {#variable-lists}

## Indoor environment variables (daily, participant-level)

```{r define-indoor-list}

data_list_num_environmental <- list(
  indoor_environment = final_indoor_environment_data %>%
    mutate(period = "AM")
)

# Each entry specifies: display label, source dataframe, variable name, date column
outcome_vars_num_environmental <- list(
  "Indoor Average Temperature (°C)" = list(
    data = "indoor_environment", var = "temp_avg", date = "date_converted"),
  "Indoor Minimal Temperature (°C)" = list(
    data = "indoor_environment", var = "temp_min", date = "date_converted"),
  "Indoor Maximal Temperature (°C)" = list(
    data = "indoor_environment", var = "temp_max", date = "date_converted"),
  "Indoor Average Relative Humidity (%)" = list(
    data = "indoor_environment", var = "rh_avg",   date = "date_converted")
)
```


## Outdoor environment variables (daily, station-level)

```{r define-outdoor-list}
# Outdoor data is station-level (not participant-level)
# participant_id = "Outdoor" is set as a placeholder to satisfy
# the descriptive_stats() function signature which requires this column
final_outdoor_environment_data <- final_outdoor_environment_data %>%
  mutate(participant_id = "Outdoor")

data_list_num_no_participant <- list(
  outdoor_environment = final_outdoor_environment_data
)

outcome_vars_num_no_participant <- list(
  "Outdoor Average Temperature (deg C)" = list(
    data = "outdoor_environment", var = "avg_tp", date = "date_converted"),
  "Outdoor Minimal Temperature (deg C)" = list(
    data = "outdoor_environment", var = "min_tp", date = "date_converted"),
  "Outdoor Maximal Temperature (deg C)" = list(
    data = "outdoor_environment", var = "max_tp", date = "date_converted"),
  "Outdoor Average Relative Humidity (%)" = list(
    data = "outdoor_environment", var = "avg_rh", date = "date_converted")
)
```


## Hourly environment variables (indoor and outdoor)

```{r define-hourly-list}
data_list_num_environmental_hourly <- list(
  indoor_environment_hourly  = final_indoor_environment_hourly,
  outdoor_environment_hourly = final_outdoor_environment_hourly
)

outcome_vars_num_environmental_hourly <- list(
  "Indoor Average Temperature (deg C)" = list(
    data = "indoor_environment_hourly",  var = "temp_mean", date = "hour_floor"),
  "Indoor Minimal Temperature (deg C)" = list(
    data = "indoor_environment_hourly",  var = "temp_min",  date = "hour_floor"),
  "Indoor Maximal Temperature (deg C)" = list(
    data = "indoor_environment_hourly",  var = "temp_max",  date = "hour_floor"),
  "Indoor Average Relative Humidity (%)" = list(
    data = "indoor_environment_hourly",  var = "rh_mean",   date = "hour_floor"),
  "Outdoor Average Temperature (deg C)" = list(
    data = "outdoor_environment_hourly", var = "temp_mean", date = "hour_floor"),
  "Outdoor Average Relative Humidity (%)" = list(
    data = "outdoor_environment_hourly", var = "rh_mean",   date = "hour_floor")
)
```

---

# Descriptive statistics {#descriptive-stats}

We use the custom `descriptive_stats()` function to compute mean, SD, median, and quantiles for each outcome variable grouped by heatwave period.
See `PRG/Functions/functions_descriptive_stats.R` for full documentation.

## Indoor environment by heatwave period


```{r StatDesc-indoor-heatwave}
outcome_table_heatwave_indoor <- descriptive_stats(
  data_list_num_environmental,
  outcome_vars_num_environmental,
  group_by_participant = TRUE,
  group_by             = "heatwave",
  categorical_grouping = FALSE,
  numeric_variable     = TRUE
)
display_table(outcome_table_heatwave_indoor)
```

## Outdoor environment by heatwave period

```{r StatDesc-outdoor-heatwave}
outcome_table_heatwave_outdoor <- descriptive_stats(
  data_list_num_no_participant,
  outcome_vars_num_no_participant,
  group_by_participant = FALSE,
  group_by             = "heatwave",
  categorical_grouping = FALSE,
  numeric_variable     = TRUE
)
display_table(outcome_table_heatwave_outdoor)
```

---

# Temperature threshold exceedance {#thresholds}

We compute the number and proportion of daily observations exceeding two
indoor temperature thresholds recommended in the heat-health literature:

- **26 deg C** — commonly used threshold for indoor thermal discomfort
- **31 deg C** — threshold associated with susceptible to increased health risk during heatwaves

Results are grouped by participant and heatwave period.

## Threshold: 26 deg C

```{r threshold-26}
threshold_TP_26 <- final_indoor_environment_data %>%
  filter(heatwave_status %in% c("HW_June", "BEF_HW_June", "AFT_HW_June")) %>%
  mutate(
    heatwave_status = factor(
      heatwave_status,
      levels = c("BEF_HW_June", "HW_June", "AFT_HW_June")
    )
  ) %>%
  group_by(heatwave_status, participant_id) %>%
  summarize(
    # Average temperature summary
    mean_temp_avg    = round(mean(temp_avg, na.rm = TRUE), 2),
    n_above_26_avg   = sum(temp_avg > 26,  na.rm = TRUE),
    n_below_26_avg   = sum(temp_avg <= 26, na.rm = TRUE),
    pct_above_26_avg = round(mean(temp_avg > 26, na.rm = TRUE) * 100, 1),
    total_avg        = sum(!is.na(temp_avg)),
    # Maximal temperature summary
    mean_temp_max    = round(mean(temp_max, na.rm = TRUE), 2),
    n_above_26_max   = sum(temp_max > 26,  na.rm = TRUE),
    n_below_26_max   = sum(temp_max <= 26, na.rm = TRUE),
    pct_above_26_max = round(mean(temp_max > 26, na.rm = TRUE) * 100, 1),
    total_max        = sum(!is.na(temp_max)),
    .groups = "drop"
  )

display_table(threshold_TP_26)
```

## Threshold: 31 deg C

```{r threshold-31}
threshold_TP_31 <- final_indoor_environment_data %>%
  filter(heatwave_status %in% c("HW_June", "BEF_HW_June", "AFT_HW_June")) %>%
  mutate(
    heatwave_status = factor(
      heatwave_status,
      levels = c("BEF_HW_June", "HW_June", "AFT_HW_June")
    )
  ) %>%
  group_by(heatwave_status, participant_id) %>%
  summarize(
    # Average temperature summary
    mean_temp_avg    = round(mean(temp_avg, na.rm = TRUE), 2),
    n_above_31_avg   = sum(temp_avg > 31,  na.rm = TRUE),
    n_below_31_avg   = sum(temp_avg <= 31, na.rm = TRUE),
    pct_above_31_avg = round(mean(temp_avg > 31, na.rm = TRUE) * 100, 1),
    total_avg        = sum(!is.na(temp_avg)),
    # Maximal temperature summary
    mean_temp_max    = round(mean(temp_max, na.rm = TRUE), 2),
    n_above_31_max   = sum(temp_max > 31,  na.rm = TRUE),
    n_below_31_max   = sum(temp_max <= 31, na.rm = TRUE),
    pct_above_31_max = round(mean(temp_max > 31, na.rm = TRUE) * 100, 1),
    total_max        = sum(!is.na(temp_max)),
    .groups = "drop"
  )

display_table(threshold_TP_31)
```

---

# Save descriptive tables to Excel {#save-tables}

```{r save-tables}
# Rename to avoid overwriting objects used in plots
table_heatwave_indoor  <- outcome_table_heatwave_indoor
table_heatwave_outdoor <- outcome_table_heatwave_outdoor

# Export all tables to a single Excel workbook with separate sheets
wb <- createWorkbook()
addWorksheet(wb, "Indoor_heatwave")
writeData(wb, "Indoor_heatwave",  table_heatwave_indoor$stats)

addWorksheet(wb, "Outdoor_heatwave")
writeData(wb, "Outdoor_heatwave", table_heatwave_outdoor$stats)

addWorksheet(wb, "Threshold_over_31")
writeData(wb, "Threshold_over_31", threshold_TP_31)

saveWorkbook(
  wb,
  file      = here("RES/Environmental/Table/table_pooled_env_heatwave.xlsx"),
  overwrite = TRUE
)
message("Exported: table_pooled_env_heatwave.xlsx (3 sheets)")
```

---

# Descriptive plots {#descriptive-plots}

## Time series plots

The following time series plots show each outcome variable per heatwave
period (BEF/HW/AFT). Vertical dashed lines mark the start and end of the
heatwave. Each point represents the daily mean for one participant. Thin
grey lines connect measurements from the same participant across days.
Shaded areas represent 95% confidence intervals of the group mean.

Plots are generated by the custom `time_series_*` functions and saved
automatically to `RES/Environmental/Graph/`. See `PRG/Functions/functions_descriptive_plot.R` for full documentation.

```{r plot-timeseries-heatwave, warning=FALSE}
# Pooled time series by heatwave period and period of day
# Note: eval=FALSE version (time_series_heatwave_period) omitted here as
# the combined plot below provides equivalent information in a cleaner format
time_series_combine_plot_environment(
  data_list_num_environmental,
  outcome_vars_num_environmental,
  output_dir = here("RES/Environmental/Graph")
)
```

```{r plot-timeseries-individual, warning=FALSE}
# Individual participant trajectories overlaid on group mean
time_series_combine_plot_environment_individual(
  data_list_num_environmental,
  outcome_vars_num_environmental,
  output_dir = here("RES/Environmental/Graph")
)
```

## Hourly time series by heatwave phase

```{r plot-hourly-timeseries, warning=FALSE}
# Hourly resolution plots for indoor and outdoor variables
# Shows diurnal patterns within each heatwave phase
time_series_environment_by_phase(
  data_list_num_environmental_hourly,
  outcome_vars_num_environmental_hourly,
  output_dir = here("RES/Environmental/Graph")
)
```


---

# Air conditioning subgroup analysis {#ac-analysis}

We investigate the effect of air conditioning (AC) access on indoor
temperature. AC status is derived from participant self-report at baseline.

## Define AC status

```{r define-ac-status}
# Participants with access to air conditioning
# Verified against characteristics_participants.csv (ApartmentAC column)
# Hardcoded for reproducibility -- update if study population changes
participants_with_AC <- c(
  "P01", "P02", "P03", "P04", "P06", "P08", "P09", "P10",
  "P11", "P13", "P16", "P17", "P19", "P21", "P24", "P25", "P26"
)

# Add AC binary indicator to indoor environment data
final_indoor_environment_data <- final_indoor_environment_data %>%
  mutate(
    AC = ifelse(participant_id %in% participants_with_AC, 1, 0)
  )

# Update the data list to use the AC-enriched dataframe
data_list_num_environmental$indoor_environment <- final_indoor_environment_data
```

## Time series plots by AC status

```{r plot-ac-timeseries, warning=FALSE}
# Compare indoor temperature trajectories between AC and non-AC participants
time_series_heatwave_AC(
  data_list_num_environmental,
  outcome_vars_num_environmental,
  group_by   = "heatwave",
  output_dir = here("RES/Environmental/Graph")
)
```



---

# Statistical analysis: Linear mixed models {#lmm}


## Main effect of heatwave period

```{r lmm-main-effect}
# One-way MMRM: effect of heatwave period (BEF / HW / AFT) on indoor
# environment variables. posthoc = TRUE runs emmeans pairwise comparisons.
lmm_one_way(
  data_list_num_environmental,
  outcome_vars_num_environmental,
  posthoc = TRUE
)
```

## Effect of air conditioning

```{r lmm-ac-effect}
# MMRM including AC status as a moderator variable
lmm_one_way_AC(
  data_list_num_environmental,
  outcome_vars_num_environmental,
  posthoc = TRUE
)
```

## Recovery analysis

```{r lmm-recovery}
# MMRM focused on the post-heatwave recovery period
# Tests whether indoor environment returns to pre-heatwave levels
lmm_recovery(
  data_list_num_environmental,
  outcome_vars_num_environmental
)
```

---

# Building characteristics and indoor temperature {#building-characteristics}

Given the small sample size, formal statistical testing of building
characteristic effects has insufficient power. We therefore use visual
inspection to explore distributional patterns across building features.

Six characteristics are examined: AC ownership, fan ownership, shading,
number of rooms, thermal insulation (self-reported), and frequency of
feeling too hot inside.

> Note: P03, P14, P18, P23 are excluded from these plots as they have no indoor sensor data.

## Prepare building characteristics dataset

```{r prepare-building-data}
# Join baseline building characteristics with daily indoor temperature
plot_data <- final_baseline %>%
  select(
    participant_id,
    ApartmentAC, ApartmentFansYesNo,
    ApartmentShadeYesNo, ApartmentRooms,
    ApartmentThermalInsulation,
    ApartmentTooHotInside
  ) %>%
  filter(
    !is.na(ApartmentAC),
    !is.na(ApartmentFansYesNo),
    !is.na(ApartmentShadeYesNo),
    !is.na(ApartmentRooms)
  ) %>%
  inner_join(
    final_indoor_environment_data %>%
      filter(heatwave_status %in% c("BEF_HW_June", "HW_June", "AFT_HW_June")) %>%
      select(participant_id, date_converted, heatwave_status,
             temp_avg, temp_min, temp_max, rh_avg),
    by = "participant_id"
  ) %>%
  mutate(
    ApartmentAC                = factor(ApartmentAC,
                                        labels = c("No AC", "Owned AC")),
    ApartmentFansYesNo         = factor(ApartmentFansYesNo,
                                        labels = c("No fan", "Owned fan")),
    ApartmentShadeYesNo        = factor(ApartmentShadeYesNo,
                                        labels = c("No shade", "Shade")),
    ApartmentThermalInsulation = factor(ApartmentThermalInsulation,
                                        levels = c("bad", "adequately", "well")),
    ApartmentTooHotInside      = factor(ApartmentTooHotInside,
                                        levels = c("never", "monthly",
                                                   "weekly", "daily")),
    heatwave_status = factor(
      heatwave_status,
      levels = c("BEF_HW_June", "HW_June", "AFT_HW_June"),
      labels = c("Pre-heatwave", "During heatwave", "Post-heatwave")
    )
  )
```

## Average indoor temperature by building characteristics

```{r plot-building-avg, fig.height=30, fig.width=22}
p_ac <- ggplot(plot_data,
               aes(x = ApartmentAC, y = temp_avg, fill = ApartmentAC)) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +
  geom_jitter(width = 0.1, size = 1, alpha = 0.4) +
  scale_fill_viridis_d() +
  facet_wrap(~ heatwave_status, nrow = 1) +
  labs(x = NULL, y = "Average indoor temperature (deg C)",
       title = "Air conditioning") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none", strip.text = element_text(size = 11))

p_fan <- ggplot(plot_data,
                aes(x = ApartmentFansYesNo, y = temp_avg,
                    fill = ApartmentFansYesNo)) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +
  geom_jitter(width = 0.1, size = 1, alpha = 0.4) +
  scale_fill_viridis_d() +
  facet_wrap(~ heatwave_status, nrow = 1) +
  labs(x = NULL, y = "Average indoor temperature (deg C)",
       title = "Fans") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none", strip.text = element_text(size = 11))

p_shade <- ggplot(plot_data,
                  aes(x = ApartmentShadeYesNo, y = temp_avg,
                      fill = ApartmentShadeYesNo)) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +
  geom_jitter(width = 0.1, size = 1, alpha = 0.4) +
  scale_fill_viridis_d() +
  facet_wrap(~ heatwave_status, nrow = 1) +
  labs(x = NULL, y = "Average indoor temperature (deg C)",
       title = "Apartment in the shade during the day") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none", strip.text = element_text(size = 11))

p_rooms <- ggplot(plot_data, aes(x = ApartmentRooms, y = temp_avg)) +
  geom_point(size = 1.5, alpha = 0.4) +
  geom_smooth(method = "lm", se = TRUE,
              linetype = "dashed", linewidth = 0.8) +
  facet_wrap(~ heatwave_status, nrow = 1) +
  labs(x = "Number of rooms", y = "Average indoor temperature (deg C)",
       title = "Number of rooms") +
  theme_minimal(base_size = 13) +
  theme(strip.text = element_text(size = 11))

p_insulation <- ggplot(
  plot_data %>% filter(!is.na(ApartmentThermalInsulation)),
  aes(x = ApartmentThermalInsulation, y = temp_avg,
      fill = ApartmentThermalInsulation)
) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +
  geom_jitter(width = 0.1, size = 1, alpha = 0.4) +
  scale_fill_viridis_d() +
  facet_wrap(~ heatwave_status, nrow = 1) +
  labs(x = NULL, y = "Average indoor temperature (deg C)",
       title = "Thermal insulation (self-reported)") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none", strip.text = element_text(size = 11))

p_toohot <- ggplot(
  plot_data %>% filter(!is.na(ApartmentTooHotInside)),
  aes(x = ApartmentTooHotInside, y = temp_avg,
      fill = ApartmentTooHotInside)
) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +
  geom_jitter(width = 0.1, size = 1, alpha = 0.4) +
  scale_fill_viridis_d() +
  facet_wrap(~ heatwave_status, nrow = 1) +
  labs(x = NULL, y = "Average indoor temperature (deg C)",
       title = "Frequency of feeling too hot inside") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none", strip.text = element_text(size = 11))

combined_avg <- (p_ac / p_fan / p_shade / p_insulation / p_toohot / p_rooms) +
  plot_annotation(
    title    = "Building characteristics and average indoor temperature across heatwave phases",
    subtitle = "Each point represents one daily observation. P03, P14, P18, P23 excluded (no indoor sensor)."
  )

print(combined_avg)

ggsave(
  filename = here("RES/Environmental/Graph/building_characteristics_temperature.png"),
  plot     = combined_avg,
  width    = 22, height = 30, dpi = 600, device = "png"
)
```

## Maximal indoor temperature by building characteristics

```{r plot-building-max, fig.height=30, fig.width=22}
p_ac_max <- ggplot(plot_data,
                   aes(x = ApartmentAC, y = temp_max, fill = ApartmentAC)) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +
  geom_jitter(width = 0.1, size = 1, alpha = 0.4) +
  scale_fill_viridis_d() +
  facet_wrap(~ heatwave_status, nrow = 1) +
  labs(x = NULL, y = "Maximal indoor temperature (deg C)",
       title = "Air conditioning") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none", strip.text = element_text(size = 11))

p_fan_max <- ggplot(plot_data,
                    aes(x = ApartmentFansYesNo, y = temp_max,
                        fill = ApartmentFansYesNo)) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +
  geom_jitter(width = 0.1, size = 1, alpha = 0.4) +
  scale_fill_viridis_d() +
  facet_wrap(~ heatwave_status, nrow = 1) +
  labs(x = NULL, y = "Maximal indoor temperature (deg C)",
       title = "Fans") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none", strip.text = element_text(size = 11))

p_shade_max <- ggplot(plot_data,
                      aes(x = ApartmentShadeYesNo, y = temp_max,
                          fill = ApartmentShadeYesNo)) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +
  geom_jitter(width = 0.1, size = 1, alpha = 0.4) +
  scale_fill_viridis_d() +
  facet_wrap(~ heatwave_status, nrow = 1) +
  labs(x = NULL, y = "Maximal indoor temperature (deg C)",
       title = "Apartment in the shade during the day") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none", strip.text = element_text(size = 11))

p_insulation_max <- ggplot(
  plot_data %>% filter(!is.na(ApartmentThermalInsulation)),
  aes(x = ApartmentThermalInsulation, y = temp_max,
      fill = ApartmentThermalInsulation)
) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +
  geom_jitter(width = 0.1, size = 1, alpha = 0.4) +
  scale_fill_viridis_d() +
  facet_wrap(~ heatwave_status, nrow = 1) +
  labs(x = NULL, y = "Maximal indoor temperature (deg C)",
       title = "Thermal insulation (self-reported)") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none", strip.text = element_text(size = 11))

p_toohot_max <- ggplot(
  plot_data %>% filter(!is.na(ApartmentTooHotInside)),
  aes(x = ApartmentTooHotInside, y = temp_max,
      fill = ApartmentTooHotInside)
) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +
  geom_jitter(width = 0.1, size = 1, alpha = 0.4) +
  scale_fill_viridis_d() +
  facet_wrap(~ heatwave_status, nrow = 1) +
  labs(x = NULL, y = "Maximal indoor temperature (deg C)",
       title = "Frequency of feeling too hot inside") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none", strip.text = element_text(size = 11))

p_rooms_max <- ggplot(plot_data, aes(x = ApartmentRooms, y = temp_max)) +
  geom_point(size = 1.5, alpha = 0.4) +
  geom_smooth(method = "lm", se = TRUE,
              linetype = "dashed", linewidth = 0.8) +
  facet_wrap(~ heatwave_status, nrow = 1) +
  labs(x = "Number of rooms", y = "Maximal indoor temperature (deg C)",
       title = "Number of rooms") +
  theme_minimal(base_size = 13) +
  theme(strip.text = element_text(size = 11))

combined_max <- (p_ac_max / p_fan_max / p_shade_max /
                 p_insulation_max / p_toohot_max / p_rooms_max) +
  plot_annotation(
    title    = "Building characteristics and maximal indoor temperature across heatwave phases",
    subtitle = "Each point represents one daily observation. P03, P14, P18, P23 excluded (no indoor sensor)."
  )

ggsave(
  filename = here("RES/Environmental/Graph/building_characteristics_temp_max.png"),
  plot     = combined_max,
  width    = 22, height = 30, dpi = 600, device = "png"
)
```

---

# Session information

```{r session-info}
sessionInfo()
```

