---
title: "Outdoor Environment HeatSuite 2024"
author: "Souilla Luc"
date: "2026-03-04"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
encoding: UTF-8
---
<!--
================================================================================
SCRIPT:  PRG/Environmental/outdoor_environment.Rmd
PURPOSE: Data preparation pipeline for outdoor environmental data (temperature and relative humidity) from the Montreal meteorological station during the Montreal Heatwave 2024. This script:
           1. Loads hourly outdoor data and assigns heatwave status
           2. Exports hourly dataset for time-series analyses
           3. Aggregates to daily summaries (mean, min, max)
           4. Flags and removes implausible values
           5. Exports daily processed dataset and QC plots

INPUT FILES (DAT/):
  - env_outdoor_hourly.csv          -> hourly outdoor temperature and humidity from Montreal meteorological station.  Note: column names are in French(Annee, Mois, Jour, Heure,Temp, Humidite.relative)

OUTPUT FILES (DAT/Processed/):
  - outdoor_environment_hourly.csv  -> hourly outdoor data with heatwave status
  - outdoor_environment_processed.csv -> daily aggregated outdoor data

IMPLAUSIBLE VALUE THRESHOLDS:
  Temperature : avg_tp < 1 or > 45 deg C
  Relative humidity : avg_rh < 1 or > 100 %


NOTE ON COLUMN NAMES:
  The source CSV uses French column names from Environment Canada:
    Annee            -> year
    Mois             -> month
    Jour             -> day
    Heure            -> hour (integer, e.g. 14 for 14:00)
    Temp             -> temperature (deg C)
    Humidite.relative -> relative humidity (%)
  These are used directly -- do not rename the CSV headers.

HEATWAVE PERIODS:
  Before heatwave : 2024-06-11 to 2024-06-17
  During heatwave : 2024-06-18 to 2024-06-20
  After heatwave  : 2024-06-21 to 2024-06-27
================================================================================
-->

```{r setup, include=FALSE}
# Global knitr options: show code, suppress messages and warnings
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE
)
```

# Overview

This script prepares the **outdoor environment** dataset from hourly
meteorological station data for Montreal during the Summer 2024 heatwave study.
Two outputs are produced: an hourly dataset for time-series analyses and a daily aggregated dataset for statistical modelling.

Unlike indoor data, outdoor data comes from a fixed station (not participant devices), so no participant ID assignment is needed.

***

# Environment setup

## Load packages

```{r load-packages, message=FALSE, warning=FALSE, include=FALSE}
required_packages <- c(
  "here", "dplyr", "tidyr", "lubridate",
  "ggplot2", "patchwork"
)

install_if_missing <- function(packages) {
  missing <- packages[!(packages %in% installed.packages()[, "Package"])]
  if (length(missing) > 0) install.packages(missing)
}

install_if_missing(required_packages)
invisible(lapply(required_packages, library, character.only = TRUE))
```

## Set project root and directory paths

```{r set-paths}
here::i_am("PRG/Environmental/outdoor_environment.Rmd")

path_dat <- here("DAT")           # Raw data folder
path_prg <- here("PRG")           # Scripts folder
path_res <- here("DAT/Processed") # Output folder for processed data
```

***

# Step 1 - Load raw data {#step1}

```{r OutEnv-upload}
# Load hourly outdoor meteorological data from Environment Canada
# Column names are in French -- see header block for mapping
outdoor_environment_raw <- read.csv(
  file.path(path_dat, "env_outdoor_hourly.csv"),
  header = TRUE, sep = ",", quote = "\"'", dec = "."
)
```

***
# Step 2 - Create hourly dataset {#step2}

The hourly dataset preserves full temporal resolution for time-series analyses and is exported directly after heatwave status assignment.

```{r OutEnv-create-hourly}
# Parse date and hour from French column names
# Note: Heure contains integer hour (e.g. 14), not "14:00"
# Format "%Y-%m-%d-%H" handles integer hour correctly
outdoor_hourly <- outdoor_environment_raw %>%
  mutate(
    hour_floor = as.POSIXct(
      paste(Année, Mois, Jour, Heure, sep = "-"),
      format = "%Y-%m-%d-%H"   # %H handles integer hour (0-23)
    ),
    date_converted = as.Date(paste(Année, Mois, Jour, sep = "-")),
    temp_mean      = as.numeric(Temp),
    rh_mean        = as.numeric(Humidite.relative)
  ) %>%
  # Remove rows where datetime parsing failed
  filter(!is.na(hour_floor)) %>%
  mutate(
    heatwave_status = case_when(
      date_converted >= as.Date("2024-06-18") & date_converted <= as.Date("2024-06-20") ~ "HW_June",
      date_converted >= as.Date("2024-06-11") & date_converted <  as.Date("2024-06-18") ~ "BEF_HW_June",
      date_converted >  as.Date("2024-06-20") & date_converted <= as.Date("2024-06-27") ~ "AFT_HW_June",
      TRUE ~ "NoHW"
    )
  ) %>%
  filter(heatwave_status != "NoHW") %>%
  select(hour_floor, date_converted, heatwave_status, temp_mean, rh_mean)

message("Hourly records retained: ", nrow(outdoor_hourly))
```

## Export hourly dataset

```{r OutEnv-export-hourly}
write.csv(
  outdoor_hourly,
  file      = file.path(path_res, "outdoor_environment_hourly.csv"),
  row.names = FALSE
)
message("Exported: outdoor_environment_hourly.csv (",
        nrow(outdoor_hourly), " rows)")
```

***

# Step 3 - Create daily aggregated dataset {#step3}

Hourly data is aggregated to daily summaries (mean, min, max) for each
meteorological variable.



```{r OutEnv-create-daily}
outdoor_daily <- outdoor_environment_raw %>%
  mutate(
    date_converted = as.Date(paste(Année, Mois, Jour, sep = "-")),
    temp           = Temp,
    rh             = Humidite.relative
  ) %>%
  group_by(date_converted) %>%
  summarise(
    # Temperature summaries
    avg_tp    = round(mean(temp, na.rm = TRUE), 2),
    min_tp    = round(min(temp,  na.rm = TRUE), 2),
    max_tp    = round(max(temp,  na.rm = TRUE), 2),
    sample_tp = sum(!is.na(temp)),   # Number of valid hourly records
    # Relative humidity summaries
    avg_rh    = round(mean(rh, na.rm = TRUE), 2),
    min_rh    = round(min(rh,  na.rm = TRUE), 2),
    max_rh    = round(max(rh,  na.rm = TRUE), 2),
    sample_rh = sum(!is.na(rh)),
    .groups   = "drop"
  ) %>%
  mutate(
    heatwave_status = case_when(
      date_converted >= as.Date("2024-06-18") & date_converted <= as.Date("2024-06-20") ~ "HW_June",
      date_converted >= as.Date("2024-06-11") & date_converted <  as.Date("2024-06-18") ~ "BEF_HW_June",
      date_converted >  as.Date("2024-06-20") & date_converted <= as.Date("2024-06-27") ~ "AFT_HW_June",
      TRUE ~ "NoHW"
    ),
    heatwave_status = factor(
      heatwave_status,
      levels = c("BEF_HW_June", "HW_June", "AFT_HW_June", "NoHW")
    )
  )
```



```{r OutEnv-inspect-daily}
# Flag implausible values BEFORE removing them for visual inspection
outdoor_daily <- outdoor_daily %>%
  mutate(
    temp_flag = case_when(
      is.na(avg_tp) ~ NA_real_,
      avg_tp < 1    ~ 1,
      avg_tp > 45   ~ 1,
      TRUE          ~ 0
    ),
    rh_flag = case_when(
      is.na(avg_rh) ~ NA_real_,
      avg_rh < 1    ~ 1,
      avg_rh > 100  ~ 1,
      TRUE          ~ 0
    ),
    measurement_type = case_when(
      temp_flag == 1 ~ "temp_flag",
      rh_flag   == 1 ~ "rh_flag",
      TRUE           ~ "Valid"
    ),
    measurement_type = factor(measurement_type,
                               levels = c("Valid", "temp_flag", "rh_flag"))
  )

flag_colors <- c("Valid"     = "#2E86AB",
                 "temp_flag" = "#F6AE2D",
                 "rh_flag"   = "#E63946")
flag_shapes <- c("Valid"     = 16,
                 "temp_flag" = 17,
                 "rh_flag"   = 4)

outdoor_daily_filtered <- outdoor_daily %>%
  filter(heatwave_status %in% c("BEF_HW_June", "HW_June", "AFT_HW_June"))

line_temp_bef <- ggplot(
  outdoor_daily_filtered,
  aes(x = date_converted, y = avg_tp)
) +
  geom_line(color = "grey", linewidth = 1, alpha = 0.6) +
  geom_point(aes(color = measurement_type, shape = measurement_type),
             size = 2.5, alpha = 0.85) +
  scale_color_manual(values = flag_colors, name = "Type") +
  scale_shape_manual(values = flag_shapes, name = "Type") +
  labs(title = "Before filtering",
       x = "", y = "Average outdoor temperature (deg C)") +
  theme_minimal() +
  theme(legend.position = "right",
        axis.text.x     = element_text(angle = 45, hjust = 1, size = 10))

line_rh_bef <- ggplot(
  outdoor_daily_filtered,
  aes(x = date_converted, y = avg_rh)
) +
  geom_line(color = "grey", linewidth = 1, alpha = 0.6) +
  geom_point(aes(color = measurement_type, shape = measurement_type),
             size = 2.5, alpha = 0.85) +
  scale_color_manual(values = flag_colors, name = "Type") +
  scale_shape_manual(values = flag_shapes, name = "Type") +
  labs(title = "Before filtering",
       x = "", y = "Average outdoor relative humidity (%)") +
  theme_minimal() +
  theme(legend.position = "right",
        axis.text.x     = element_text(angle = 45, hjust = 1, size = 10))
```

## Restrict to study period

```{r OutEnv-filter-heatwave}
final_outdoor_environment <- outdoor_daily %>%
  filter(heatwave_status %in% c("BEF_HW_June", "HW_June", "AFT_HW_June"))

message("Daily records retained: ", nrow(final_outdoor_environment))
```

***

# Step 4 - Flag and remove implausible values {#step4}

Implausible values are flagged based on meteorologically plausible ranges for Montreal in June:

| Variable | Flag condition |
|---|---|
| avg_tp | < 1 deg C or > 45 deg C |
| avg_rh | < 1% or > 100% |
```{r OutEnv-flag-implausible}
final_outdoor_environment <- final_outdoor_environment %>%
  mutate(
    temp_flag = case_when(
      is.na(avg_tp)  ~ NA_real_,
      avg_tp < 1     ~ 1,
      avg_tp > 45    ~ 1,   
      TRUE           ~ 0
    ),
    rh_flag = case_when(
      is.na(avg_rh)  ~ NA_real_,
      avg_rh < 1     ~ 1, 
      avg_rh > 100   ~ 1,   
      TRUE           ~ 0
    )
  )

message("=== Temperature implausible values ===")
message("N implausible: ",
        sum(final_outdoor_environment$temp_flag == 1, na.rm = TRUE))
message("% implausible: ",
        round(mean(final_outdoor_environment$temp_flag == 1,
                   na.rm = TRUE) * 100, 2), "%")

message("=== Relative humidity implausible values ===")
message("N implausible: ",
        sum(final_outdoor_environment$rh_flag == 1, na.rm = TRUE))
message("% implausible: ",
        round(mean(final_outdoor_environment$rh_flag == 1,
                   na.rm = TRUE) * 100, 2), "%")

# Remove flagged rows
n_before <- nrow(final_outdoor_environment)
final_outdoor_environment <- final_outdoor_environment %>%
  filter(temp_flag == 0 & rh_flag == 0) %>%
  select(-temp_flag, -rh_flag)

message("Rows removed (implausible values): ",
        n_before - nrow(final_outdoor_environment))
```

***
# Step 5 - QC plots {#step5}

## Outdoor temperature over time

```{r OutEnv-plot-temperature, echo=FALSE}
line_temp_pooled_after <- ggplot(
  final_outdoor_environment,
  aes(x = date_converted, y = avg_tp)
) +
  geom_line(color = "grey", linewidth = 1, alpha = 0.85) +
  geom_point(color = "#2E86AB", size = 2.5, alpha = 0.85) +
  labs(title = "Outdoor temperature after filtering",
       x = "", y = "Average outdoor temperature (deg C)") +
  theme_minimal() +
  theme(legend.position  = "right",
        axis.text.x      = element_text(angle = 45, hjust = 1, size = 10))

```

## Outdoor relative humidity over time

```{r OutEnv-plot-humidity, echo=FALSE}
line_rh_pooled_after <- ggplot(
  final_outdoor_environment,
  aes(x = date_converted, y = avg_rh)
) +
  geom_line(color = "grey", linewidth = 1, alpha = 0.85) +
  geom_point(color = "#2E86AB", size = 2.5, alpha = 0.85) +
  labs(title = "Outdoor relative humidity after filtering",
       x = "", y = "Average outdoor relative humidity (%)") +
  theme_minimal() +
  theme(legend.position  = "right",
        axis.text.x      = element_text(angle = 45, hjust = 1, size = 10))

```

```{r plot-bef-after}
line_temp_bef / line_temp_pooled_after +
  plot_annotation(title = "Effect of data cleaning on outdoor temperature")

line_rh_bef / line_rh_pooled_after +
  plot_annotation(title = "Effect of data cleaning on outdoor relative humidity")

```

***

# Step 6 - Export daily processed dataset {#step6}

```{r OutEnv-export}
output_file <- file.path(path_res, "outdoor_environment_processed.csv")
write.csv(final_outdoor_environment, output_file, row.names = FALSE)
message("Exported: outdoor_environment_processed.csv (",
        nrow(final_outdoor_environment), " rows x ",
        ncol(final_outdoor_environment), " columns)")
```

***

# Session information

```{r session-info}
sessionInfo()
```