---
title: "Main Analysis Perceptuals HeatSuite 2024"
author: "Souilla Luc"
date: "2026-03-05"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
encoding: UTF-8
---
<!--
================================================================================
SCRIPT:  PRG/Perceptuals/main_analysis_perceptuals.Rmd
PURPOSE: Descriptive and inferential analysis of perceptual survey responses collected during the Montreal Summer 2024 heatwave study. This script:
           1. Loads the cleaned survey dataset
           2. Assigns AC status to participants
           3. Produces descriptive statistics tables by heatwave period
           4. Produces violin plots, bar charts, and heatmap plots
           5. Runs Cumulative Link Mixed Models (CLMM) for ordinal outcomes
           6. Runs Generalized Linear Mixed Models (glmmTMB) for binary outcomes

INPUT FILES (DAT/Processed/):
  - survey_processed.csv   -> cleaned perceptual survey responses (long format)

OUTPUT FILES (RES/Perceptuals/):
  - Table/survey_descriptive_tables.xlsx
  - Graph/violin_ordinal_period.png
  - Graph/violin_ordinal_heatwave.png
  - Graph/chart_ordinal_period.png
  - Graph/chart_ordinal_heatwave.png
  - Graph/ (additional plots saved by custom functions)

SURVEY KEYS ANALYSED:
  Ordinal    : sleep
  Categorical: preference, sensation, thirst, sleep
  Binary     : comfort, at-home, sweating

STATISTICAL APPROACH:
  Ordinal outcomes  -> Cumulative Link Mixed Models (CLMM, ordinal package)
  Binary outcomes   -> Generalized Linear Mixed Models (glmmTMB package)
    Post-hoc        -> marginaleffects (population-averaged risk ratios)

AC STATUS NOTE:
  Air conditioning status derived from participant self-report at baseline. Hardcoded list verified against characteristics_participants.csv.

HEATWAVE PERIODS:
  Before heatwave : 2024-06-11 to 2024-06-17
  During heatwave : 2024-06-18 to 2024-06-20
  After heatwave  : 2024-06-21 to 2024-06-27
================================================================================
-->

```{r setup, include=FALSE}
# Global knitr options: show code, suppress messages and warnings
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE
)
```

# Overview

This script performs descriptive and inferential analyses of perceptual survey responses collected during the Montreal Summer 2024 heatwave study. Survey outcomes are divided into ordinal (sleep, thirst), categorical (preference, sensation), and binary (comfort, at-home, sweating) variables, each analysed with an appropriate statistical model.

---

# Environment setup

## Load packages

```{r load-packages, message=FALSE, warning=FALSE, include=FALSE}
required_packages <- c(
  # Data wrangling
  "here", "dplyr", "tidyr", "lubridate", "stringr", "purrr",
  # Reporting
  "knitr", "kableExtra", "openxlsx",
  # Visualization
  "ggplot2", "patchwork", "viridis", "ggeffects",
  # Statistical modelling
  "rstatix", "carData", "car", "emmeans", "mmrm",
  "mlogit", "ordinal", "glmmTMB"
)

install_if_missing <- function(packages) {
  missing_packages <- packages[!(packages %in% installed.packages()[, "Package"])]
  if (length(missing_packages) > 0) {
    install.packages(missing_packages)
  }
}

load_packages <- function(packages) {
  lapply(packages, library, character.only = TRUE)
}

install_if_missing(required_packages)
load_packages(required_packages)
```

## Set project root and directory paths

```{r set-paths, message=FALSE, warning=FALSE, include=FALSE}
here::i_am("PRG/Perceptuals/main_analysis_perceptuals.Rmd")
```

```{r define-paths}
path_dat <- here("DAT/Processed")      # Processed data inputs
path_prg <- here("PRG")                # Scripts folder
path_res <- here("RES/Perceptuals")    # Results output folder
```

## Set global color palette

```{r set-color-palette, echo=FALSE, warning=FALSE}
scale_color_discrete <- function(...) scale_color_viridis_d(...)
scale_fill_discrete  <- function(...) scale_fill_viridis_d(...)
```

## Source custom analysis functions

```{r source-functions}
# All custom functions used in this script
# See each .R file in PRG/Functions/ for full documentation
source(here("PRG/Functions/functions_cleaning_data.R"))
source(here("PRG/Functions/functions_descriptive_stats.R"))  # descriptive_stats(), display_table(), create_survey_datasets(), get_variable_summary()
source(here("PRG/Functions/functions_descriptive_plot.R"))   # violin_plot_ordinal(), chart_categorical(), heatmap_plot_long(), heatmap_plot_facets_fixed()
source(here("PRG/Functions/functions_linear_models.R"))      # ordinal_mm(), glmmTMB_model(), clmm_recovery(), glmmTMB_recovery()
source(here("PRG/Functions/residual_check_mmrm.R"))
source(here("PRG/Functions/functions_check_clmm.R"))         # ordinal_mm_assumptions()
```

---

# Load processed data {#load-data}

```{r load-data}
# Load cleaned survey responses (long format: one row per key per participant per response)
final_survey_data <- read.csv(
  file.path(path_dat, "survey_processed.csv"),
  header = TRUE, sep = ",", quote = "\"'", dec = "."
)
```

## Assign AC status

```{r define-ac-status}
# Participants with air conditioning access at home
# Verified against characteristics_participants.csv (ApartmentAC column)
# Hardcoded for reproducibility -- update if study population changes
participants_with_AC <- c(
  "P01", "P02", "P03", "P04", "P06", "P08", "P09", "P10",
  "P11", "P13", "P16", "P17", "P19", "P21", "P24", "P25", "P26"
)

# Add binary AC indicator (1 = AC access, 0 = no AC)
final_survey_data$AC <- ifelse(
  final_survey_data$participant_id %in% participants_with_AC, 1, 0
)
```

---

# Descriptive statistics {#descriptive-stats}

## Prepare survey datasets by key

```{r survey-concatenate-survey-data, echo=FALSE, message=FALSE, warning=FALSE}
# Define the seven survey keys of interest
survey_keys <- c("sleep", "comfort", "sensation", "preference",
                 "sweating", "at-home", "thirst")

# Split the long-format survey data into one dataframe per key
# using the custom create_survey_datasets() function
survey_datasets <- create_survey_datasets(final_survey_data, survey_keys)

# Remove any datasets missing one or more heatwave periods
# (e.g., keys with no responses during HW_June would be excluded)
required_statuses <- c("BEF_HW_June", "HW_June", "AFT_HW_June")
survey_datasets <- survey_datasets[
  sapply(survey_datasets, function(df) all(required_statuses %in% unique(df$heatwave_status)))
]

# Print a summary of available responses per key per period
survey_summary <- get_variable_summary(survey_datasets)
print(survey_summary)
```

## Define analysis variable lists

```{r survey-create-data-list-outcome, echo=FALSE}
# data_list_survey: named list mapping keys to processed survey dataframes
# Sleep is filtered to AM only -- per protocol, the sleep questionnaire was completed in the morning (once per day), not 3x daily
data_list_survey <- list(
  sleep      = survey_datasets$survey_sleep %>% filter(period == "AM"),
  comfort    = survey_datasets$survey_comfort,
  preference = survey_datasets$survey_preference,
  sensation  = survey_datasets$survey_sensation,
  at_home    = survey_datasets$survey_at_home,
  sweating   = survey_datasets$survey_sweating,
  thirst     = survey_datasets$survey_thirst
)

# outcome_variables_survey: named list specifying each survey outcome
# Fields: data = key in data_list_survey, var = numeric value column,
#         resp = character response label column
outcome_variables_survey <- list(
  "Sleep"      = list(data = "sleep",      var = "value", resp = "resp"),
  "Comfort"    = list(data = "comfort",    var = "value", resp = "resp"),  
  "Preference" = list(data = "preference", var = "value", resp = "resp"),
  "Sensation"  = list(data = "sensation",  var = "value", resp = "resp"),
  "At_home"    = list(data = "at_home",    var = "value", resp = "resp"),
  "Sweating"   = list(data = "sweating",   var = "value", resp = "resp"),
  "Thirst"     = list(data = "thirst",     var = "value", resp = "resp")
)
```

## Compute descriptive statistics tables

Three grouping options are computed:

- **Option 1** (`Participant_HW_Period`): participant-weighted, grouped by
  heatwave period × period of day (AM/PM/NI) -- recommended for publication
- **Option 2** (`Participant_Period`): participant-weighted, grouped by
  heatwave period only -- for outcomes without period-of-day granularity
- **Option 3** (`Pooled_Heatwave`): pooled across participants, grouped by
  heatwave period -- for sensitivity comparison

```{r survey-descriptive-tables}
# Option 1: participant-weighted, grouped by heatwave x period of day
outcome_table_participant_survey <- descriptive_stats(
  data_list_survey,
  outcome_variables_survey,
  group_by_participant = TRUE,
  group_by             = "heatwave_period",
  categorical_grouping = FALSE,
  numeric_variable     = FALSE
)
display_table(outcome_table_participant_survey)

# Option 2: participant-weighted, grouped by heatwave period only

outcome_table_heatwave_survey <- descriptive_stats(
  data_list_survey,
  outcome_variables_survey,
  group_by_participant = TRUE,
  group_by             = "heatwave",
  categorical_grouping = FALSE,
  numeric_variable     = FALSE
)
display_table(outcome_table_heatwave_survey)

# Option 3: pooled across participants, grouped by heatwave period (categorical)
outcome_table_pooled_heatwave_survey <- descriptive_stats(
  data_list_survey,
  outcome_variables_survey,
  group_by_participant = FALSE,
  group_by             = "heatwave",
  categorical_grouping = TRUE,
  numeric_variable     = FALSE
)
display_table(outcome_table_pooled_heatwave_survey)
```

## Save descriptive tables to Excel

```{r save-survey-tables}
wb <- createWorkbook()

# Sheet 1: participant-weighted, by heatwave x period of day
addWorksheet(wb, "Participant_HW_Period")
writeData(wb, "Participant_HW_Period", outcome_table_participant_survey$stats)

# Sheet 2: participant-weighted, by heatwave period only
addWorksheet(wb, "Participant_HW")
writeData(wb, "Participant_HW", outcome_table_heatwave_survey$stats)

# Sheet 3: pooled across participants, by heatwave period
addWorksheet(wb, "Pooled_Heatwave")
writeData(wb, "Pooled_Heatwave", outcome_table_pooled_heatwave_survey$freq)

saveWorkbook(
  wb,
  file      = here("RES/Perceptuals/Table/survey_descriptive_tables.xlsx"),
  overwrite = TRUE
)
# FIX: path in message corrected from "RES/Survey/Table/" to "RES/Perceptuals/Table/"
message("Survey tables saved to RES/Perceptuals/Table/survey_descriptive_tables.xlsx")
```


---

# Descriptive plots {#descriptive-plots}

## Ordinal variables: violin plots

Ordinal variables (sleep and thirst) are visualized using violin plots showing the distribution of responses across periods of the day and heatwave periods.

```{r survey-plot-violin, echo=TRUE, fig.height=13, fig.width=19, message=FALSE, warning=FALSE}
# Ordinal variables: sleep (AM only) and thirst
data_list_survey_ordinal <- list(
  sleep  = survey_datasets$survey_sleep %>% filter(period == "AM"),
  thirst = survey_datasets$survey_thirst
)

outcome_variables_survey_ordinal <- list(
  "Sleep"  = list(data = "sleep",  var = "value", resp = "resp"),
  "Thirst" = list(data = "thirst", var = "value", resp = "resp")
)

# Violin plot by period of day
violin_period_ordinal <- violin_plot_ordinal(
  data_list_survey_ordinal,
  outcome_variables_survey_ordinal,
  plot_type = "period"
)
print(violin_period_ordinal)

# Violin plot by heatwave period
violin_heatwave_ordinal <- violin_plot_ordinal(
  data_list_survey_ordinal,
  outcome_variables_survey_ordinal,
  plot_type = "heatwave"
)
print(violin_heatwave_ordinal)
```

```{r save-violin-plots}
ggsave(
  filename = here("RES/Perceptuals/Graph/violin_ordinal_period.png"),
  plot     = violin_period_ordinal,
  width = 10, height = 6, dpi = 300, bg = "white"
)

ggsave(
  filename = here("RES/Perceptuals/Graph/violin_ordinal_heatwave.png"),
  plot     = violin_heatwave_ordinal,
  width = 10, height = 6, dpi = 300, bg = "white"
)
```

## Categorical variables: bar charts

Categorical variables (preference, sensation, thirst) are visualized as
proportional bar charts. The y-axis represents the proportion of responses (1 = 100%).

> Note: these charts do not account for multiple measurements per participant within the same period or heatwave status.

```{r survey-create-dataset-categorical, echo=TRUE, fig.height=16, fig.width=18, message=FALSE, warning=FALSE}
# Categorical variables: preference (-1/0/1), sensation (-1/0/1), thirst (0/1/2)
# Sleep commented out -- if analysing sleep with CLMM, modify clmm_model() accordingly
data_list_survey_categorical <- list(
  preference = survey_datasets$survey_preference,
  sensation  = survey_datasets$survey_sensation,
  thirst     = survey_datasets$survey_thirst,
  sleep      = survey_datasets$survey_sleep %>% filter(period == "AM")
)

outcome_variables_survey_categorical <- list(
  "Preference" = list(data = "preference", var = "value", levels = c(-1, 0, 1),   resp = "resp"),
  "Sensation"  = list(data = "sensation",  var = "value", levels = c(-1, 0, 1),   resp = "resp"),
  "Thirst"     = list(data = "thirst",     var = "value", levels = c(0, 1, 2),    resp = "resp")
  # "Sleep"    = list(data = "sleep", var = "value", levels = c(1,2,3,4,5), resp="resp")
  # Uncomment above to include sleep in CLMM -- modify clmm_model() function accordingly
)
```

```{r survey-categorical-barchart, echo=FALSE, fig.height=10, fig.width=12, message=FALSE, warning=FALSE}
# Bar chart by period of day
p_period <- chart_categorical(
  data_list_survey_categorical,
  outcome_variables_survey_categorical,
  chart_type = "period"
)

# Bar chart by heatwave period
p_hw <- chart_categorical(
  data_list_survey_categorical,
  outcome_variables_survey_categorical,
  chart_type = "heatwave"
)

ggsave(
  filename = here("RES/Perceptuals/Graph/chart_ordinal_period.png"),
  plot     = p_period,
  width = 10, height = 6, dpi = 300, bg = "white"
)

ggsave(
  filename = here("RES/Perceptuals/Graph/chart_ordinal_heatwave.png"),
  plot     = p_hw,
  width = 10, height = 6, dpi = 300, bg = "white"
)
```

## Binary variables: heatmap plots

Binary variables (comfort, at-home, sweating) are visualized as heatmaps
showing the mean response score per day per period. The color gradient
represents the mean score within each cell.

> Legend:
> - Comfort/sweating: 0 = no, 1 = yes
> - Sensation/preference: -1 = cold/cooler, 0 = neutral/no change, 1 = hot/hotter
> - Thirst: 0 = no, 1 = a little, 2 = very thirsty
> - At-home: 1 = yes, 0 = no

```{r survey-define-binary-list, echo=FALSE}
# Binary variables: comfort (0/1), at-home (1-4), sweating (0/1)
data_list_survey_binary <- list(
  comfort = survey_datasets$survey_comfort,
  at_home = survey_datasets$survey_at_home,
  sweating = survey_datasets$survey_sweating
)

outcome_variables_survey_binary <- list(
  "Comfort"  = list(data = "comfort",  var = "value", resp = "resp"),
  "At_home"  = list(data = "at_home",  var = "value", resp = "resp"),
  "Sweating" = list(data = "sweating", var = "value", resp = "resp")
)
```

```{r survey-binary-heatmap-plot, message=FALSE, warning=FALSE}
# Heatmap by heatwave period x period of day
heatmap_plot_long(
  data_list_survey_binary,
  outcome_variables_survey_binary,
  output_dir = here("RES/Perceptuals/Graph"),
  plot_type  = "heatwave_period"
)
```

```{r survey-binary-heatmap-plot-AC, message=FALSE, warning=FALSE}
# Heatmap split by AC status
heatmap_plot_long_AC(
  data_list_survey_binary,
  outcome_variables_survey_binary,
  output_dir = here("RES/Perceptuals/Graph")
)
```

```{r survey-categorical-heatmap}
# Faceted heatmap for categorical variables
heatmap_plot_facets_fixed(
  data_list_survey_categorical,
  outcome_variables_survey_categorical,
  output_dir = here("RES/Perceptuals/Graph")
)
```

```{r survey-categorical-heatmap-AC}
# Faceted heatmap for categorical variables split by AC status
heatmap_plot_facets_fixed_AC(
  data_list_survey_categorical,
  outcome_variables_survey_categorical,
  output_dir = here("RES/Perceptuals/Graph")
)
```

---

# Statistical analysis {#statistical-analysis}

## Cumulative Link Mixed Models (CLMM) for ordinal outcomes

CLMMs are designed for ordinal outcomes by linking a latent continuous
variable to observed ordinal responses via a nonlinear link function (logit,
probit, or complementary log-log). Random effects account for within-subject
variability across repeated measurements.

See `PRG/Functions/functions_linear_models.R` for model specification details.

```{r survey-clmm, message=FALSE, warning=FALSE}
# Main effect of heatwave period on ordinal outcomes (preference, sensation, thirst)
ordinal_mm(data_list_survey_categorical, outcome_variables_survey_categorical)
```

### CLMM assumption checks

```{r survey-clmm-assumptions, eval=FALSE, include=FALSE}
# Verify proportional odds assumption and other CLMM diagnostics
ordinal_mm_assumptions(data_list_survey_categorical, outcome_variables_survey_categorical)
```

### Effect of air conditioning on ordinal outcomes

```{r survey-clmm-AC}
# CLMM including AC status as a moderator variable
ordinal_mm_AC(data_list_survey_categorical, outcome_variables_survey_categorical)
```


### Recovery analysis for ordinal outcomes

```{r survey-clmm-recovery}
# CLMM focused on post-heatwave recovery (AFT vs BEF comparison)
clmm_recovery(data_list_survey_categorical, outcome_variables_survey_categorical)
```

---

## Generalized Linear Mixed Models (glmmTMB) for binary outcomes

`glmmTMB` fits generalized linear mixed models using the Template Model
Builder (TMB) backend for speed and flexibility. It is used here for binary
outcomes (comfort, at-home, sweating).

**Post-hoc: population-averaged risk ratios via `marginaleffects`**

For nonlinear models (e.g., logistic regression), differences on the log-odds
scale do not translate directly into risk ratios. `marginaleffects` computes
population-averaged contrasts on the response scale (probabilities / risk
ratios), averaging over the random effects distribution. This provides more
interpretable estimates than `emmeans` for binary outcomes.

```{r survey-glmmTMB-main, message=FALSE, warning=FALSE}
# Main effect of heatwave period on binary outcomes (comfort, at-home, sweating)
# Results stored but displayed within the function
glmmTMB_model(data_list_survey_binary, outcome_variables_survey_binary)
```

### Effect of air conditioning on binary outcomes

```{r survey-glmmTMB-AC, message=FALSE, warning=FALSE}
# glmmTMB including AC status as a moderator
glmmTMB_model_AC(data_list_survey_binary, outcome_variables_survey_binary)
```

### Recovery analysis for binary outcomes

```{r survey-glmmTMB-recovery, message=FALSE, warning=FALSE}
# glmmTMB focused on post-heatwave recovery period
glmmTMB_recovery(data_list_survey_binary, outcome_variables_survey_binary)
```

---

# Session information

```{r session-info}
sessionInfo()
```