---
title: "Survey Responses HeatSuite 2024"
author: "Souilla Luc"
date: "2026-03-04"
output: 
 html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
encoding: UTF-8
---

<!--
================================================================================
SCRIPT:  PRG/Perceptuals/surveys_responses.Rmd
PURPOSE: Data preparation pipeline for perceptual survey responses collected via smartwatch during the Montreal heatwave 2024 .
         This script:
           1. Assigns participant IDs and heatwave status to raw survey data
           2. Filters to the seven perceptual survey keys of interest
           3. Removes duplicate responses within a 10-minute window
           4. Merges with sleep data to assign reference times per period
           5. Selects one response per participant per key per period per day
              using a custom approach adapted to the multi-key survey structure
           6. Exports the clean final dataset

INPUT FILES (DAT/):
  - survey.csv                  -> raw survey responses from smartwatch
  - baseline_long.csv           -> participant-macID-date lookup
                                   (from Data_preparation.Rmd)
  - sleep_analysis_daily.csv    -> daily sleep start/end times per participant

OUTPUT FILES (DAT/Processed/):
  - survey_processed.csv        -> cleaned survey responses (long format)

SURVEY KEYS RETAINED:
  sleep, at-home, comfort, sensation, preference, sweating, thirst

DUPLICATE HANDLING -- CUSTOM APPROACH:
  Standard physiological deduplication (get_surrounding_indices /
  remove_farthest_measure) cannot be used for survey data because adjacent rows in the dataset belong to DIFFERENT keys of the same questionnaire instance, not to duplicate responses of the same key.

  Instead, a two-step approach is used:
    Step 1: Remove responses taken within 10 minutes of a prior response
            for the same participant/date/period/key (keep the last one).
            Rationale: the questionnaire takes at most 150 seconds -- a
            second response within 10 minutes is a correction attempt,
            not a genuinely independent measurement.
    Step 2: For any remaining duplicates (> 10 min apart), retain the
            response closest to the period reference time:
              AM  -> awakening time (from sleep analysis)
              PM  -> 15:00
              NI  -> bedtime (adjusted for post-midnight measurements)


HEATWAVE PERIODS:
  Before heatwave : 2024-06-11 to 2024-06-17
  During heatwave : 2024-06-18 to 2024-06-20
  After heatwave  : 2024-06-21 to 2024-06-27
================================================================================
-->

```{r setup, include=FALSE}
# Global knitr options: show code, suppress messages and warnings
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE
)
```

# Overview

This script prepares the **perceptual survey** dataset collected via smartwatch during the Montreal heatwave 2024. Survey responses cover seven perceptual dimensions: sleep quality, presence at home, thermal comfort, thermal sensation, thermal preference, sweating, and thirst.

The deduplication approach is adapted specifically for the multi-key survey structure (see the header block for full rationale).

---

# Environment setup

## Load packages

```{r load-packages, message=FALSE, warning=FALSE, include=FALSE}
required_packages <- c(
  "here", "dplyr", "tidyr", "lubridate",
  "ggplot2", "patchwork", "purrr"
)

install_if_missing <- function(packages) {
  missing <- packages[!(packages %in% installed.packages()[, "Package"])]
  if (length(missing) > 0) install.packages(missing)
}

install_if_missing(required_packages)
invisible(lapply(required_packages, library, character.only = TRUE))
```

## Set project root and directory paths

```{r set-paths}
here::i_am("PRG/Perceptuals/surveys_responses.Rmd")

path_dat <- here("DAT")           # Raw data folder
path_prg <- here("PRG")           # Scripts folder
path_res <- here("DAT/Processed") # Output folder for processed data
```

## Source custom functions

```{r source-functions}
# Participant ID assignment, period labelling, heatwave status
source(here("PRG/Functions/functions_participant_period.R"))
# Reference-time assignment per period (AM/PM/NI)
source(here("PRG/Functions/functions_cleaning_data.R"))
# Night-period reference time adjustment for post-midnight measurements
source(here("PRG/Functions/functions_adjusted_ref_time.R"))
```

---

# Step 1 - Load raw data and assign participant IDs {#step1}

## Load raw survey data

```{r Survey-upload}
# Load raw survey responses exported from smartwatch platform
# Data is in long format: one row per survey key per participant per response

survey_raw <- read.csv(
  file.path(path_dat, "survey.csv"),
  header = TRUE, sep = ",", quote = "\"'", dec = "."
)

# Load participant-macID-date lookup table
# Created by Data_preparation.Rmd -- must be run first
baseline_long <- read.csv(
  file.path(path_dat, "baseline_long.csv"),
  header = TRUE, sep = ",", quote = "\"'", dec = "."
)
```


## Assign study period, participant ID, and heatwave status

```{r Survey-assign-participant}
# Step 1a: Convert unix timestamp and assign period of day (AM / PM / NI)
survey_participant <- assign_study_period(survey_raw, c("unix"))

# Step 1b: Match each row to a participant_id using macID + date_converted
# Note: survey_raw already contains a local_time column. date_converted is
# used here for matching because it is a stable date-level key consistent
# with the survey data structure. Verify class(survey_raw$local_time) --
# if it is POSIXct, switching to time_col = "local_time" may improve precision.
survey_participant <- add_participant_id_watch(
  survey_participant,
  baseline_long,
  mac_col  = "macID",
  time_col = "date_converted"
)

# Step 1c: Assign heatwave status and restrict to the monitoring window
survey_participant <- assign_heatwave(survey_participant)

survey_participant <- survey_participant %>%
  filter(heatwave_status %in% c("BEF_HW_June", "HW_June", "AFT_HW_June"))
```

## QC: inspect available survey keys

```{r Survey-inspect-keys}
# Display all unique survey keys found in the data
# This is a QC step to confirm the expected keys are present before filtering
survey_participant_class <- unique(survey_participant$key)
message("Survey keys available: ", paste(sort(survey_participant_class), collapse = ", "))
```

## Filter to perceptual survey keys of interest

```{r filter-survey-keys}
# Retain only the seven perceptual dimensions used in this study
# All other keys (device diagnostics, additional lifestyle items) are excluded
survey_participant <- survey_participant %>%
  filter(key %in% c("sleep", "at-home", "comfort", "sensation",
                    "preference", "sweating", "thirst"))

message("Rows retained after key filter: ", nrow(survey_participant))
```


---

# Step 2 - Duplicate detection and removal {#step2}

Survey responses may be duplicated for two reasons:

1. **Rapid re-entry (< 10 min):** A participant submitted a response and
   immediately corrected it by completing the survey again. Since each
   questionnaire takes at most 150 seconds, a second response within 10
   minutes is treated as a correction = only the last response is retained.

2. **Multiple completions (> 10 min apart):** A participant genuinely
   completed the survey more than once during the same period. These are
   resolved in Step 4 using reference-time selection.

## Step 2a - Remove rapid re-entries (10-minute window)

```{r Survey-remove-rapid-duplicates}
survey_duplicate <- survey_participant %>%
  mutate(
    # Convert unix_time_converted to POSIXct for time difference calculation
    survey_time = as.POSIXct(unix_time_converted, origin = "1970-01-01", tz = "UTC")
  ) %>%
  arrange(participant_id, date_converted, period, key, survey_time) %>%
  group_by(participant_id, date_converted, period, key) %>%
  mutate(
    # Time elapsed since the previous response for the same key/period/day (seconds)
    time_diff_key = as.numeric(difftime(survey_time, lag(survey_time), units = "secs")),
    n_responses   = n(),
    response_rank = row_number(),

    # Classify each response:
    #   Valid                        -> single response, or responses > 10 min apart
    #   Remove - Duplicate (<10min) -> earlier response within a 10-min window (removed)
    #   Keep - Last response (<10min) -> most recent response within a 10-min window (kept)
    duplicate_flag = case_when(
      n_responses == 1                                                               ~ "Valid",
      !is.na(time_diff_key) & time_diff_key <= 600 & response_rank < n_responses   ~ "Remove - Duplicate (<10min)",
      !is.na(time_diff_key) & time_diff_key <= 600 & response_rank == n_responses  ~ "Keep - Last response (<10min)",
      TRUE                                                                           ~ "Valid"
    )
  ) %>%
  ungroup()

# Summary of duplicate flags
message("=== Duplicate flags (10-minute window) ===")
print(table(survey_duplicate$duplicate_flag))

message("\n=== Duplicates removed by survey key ===")
survey_duplicate %>%
  filter(duplicate_flag == "Remove - Duplicate (<10min)") %>%
  count(key) %>%
  print()

message("\n=== Duplicates removed by participant ===")
survey_duplicate %>%
  filter(duplicate_flag == "Remove - Duplicate (<10min)") %>%
  count(participant_id) %>%
  print()

# Remove flagged rapid duplicates -- keep all Valid and Keep-Last rows
survey_data_cleaned <- survey_duplicate %>%
  filter(duplicate_flag != "Remove - Duplicate (<10min)")

message("\nRows before cleaning : ", nrow(survey_duplicate))
message("Rows after cleaning  : ", nrow(survey_data_cleaned))
message("Rows removed         : ", nrow(survey_duplicate) - nrow(survey_data_cleaned))
```

## Step 2b - Check remaining duplicates (> 10 min apart)

```{r Survey-check-remaining-duplicates}
# Identify cases where a participant still has more than one response
# per key per period per day after the 10-minute window cleaning
remaining_duplicates <- survey_data_cleaned %>%
  group_by(participant_id, date_converted, period, key) %>%
  summarise(
    n_responses = n(),
    .groups = "drop"
  ) %>%
  filter(n_responses > 1)

message("=== Remaining duplicates after 10-min cleaning ===")
message("N cases with > 1 response per key/period/day/participant: ",
        nrow(remaining_duplicates))

message("\n=== By survey key ===")
remaining_duplicates %>% count(key) %>% print()

message("\n=== By participant ===")
remaining_duplicates %>% count(participant_id) %>% print()

message("\n=== Details ===")
print(remaining_duplicates)
```


---

# Step 3 - Merge with sleep data for reference-time assignment {#step3}

Remaining duplicates (> 10 min apart) are resolved by selecting the response closest to a physiologically meaningful reference time per period:

- **AM:** closest to awakening time (from sleep analysis)
- **PM:** closest to 15:00
- **NI:** closest to bedtime (adjusted for post-midnight measurements)

## Load and prepare sleep data

```{r Survey-sleep-upload}
# Load daily sleep analysis data for awakening and bedtime reference times
sleep_analysis_raw <- read.csv(
  file.path(path_dat, "sleep_analysis_daily.csv"),
  header = TRUE, sep = ",", quote = "\"'", dec = "."
)
```

```{r Survey-sleep-assign-participant}
# Derive local_time from sleepEnd for participant ID matching
# EDT offset: -14400 seconds (UTC-4, valid for June 2024 Montreal)
sleep_analysis_raw <- sleep_analysis_raw %>%
  mutate(
    local_time = as.Date(
      as.POSIXct(sleepEnd, origin = "1970-01-01", tz = "UTC") - 14400
    )
  )

sleep_analysis_participant <- add_participant_id_watch(
  sleep_analysis_raw,
  baseline_long,
  mac_col  = "macID",
  time_col = "local_time"
)
```

```{r Survey-sleep-filter, message=FALSE, warning=FALSE}
# Convert Unix timestamps to local Montreal time and extract date components
sleep_analysis_participant <- sleep_analysis_participant %>%
  mutate(
    # Apply EDT offset (-14400 s) to convert UTC unix to local Montreal time
    sleepStart_time = as.POSIXct(sleepStart, origin = "1970-01-01", tz = "UTC") - 14400,
    sleepEnd_time   = as.POSIXct(sleepEnd,   origin = "1970-01-01", tz = "UTC") - 14400,
    date_bedtime    = as.Date(sleepStart_time),
    date_awakening  = as.Date(sleepEnd_time)
  ) %>%
  # Restrict to heatwave monitoring window (June 9-29, 2024)
  filter(
    (sleepStart_time >= as.POSIXct("2024-06-09 00:00:00") &
     sleepStart_time <= as.POSIXct("2024-06-29 23:59:59")) |
    (sleepEnd_time   >= as.POSIXct("2024-06-09 00:00:00") &
     sleepEnd_time   <= as.POSIXct("2024-06-29 23:59:59"))
  )

# Retain only physiologically plausible sleep episodes
sleep_analysis_filter <- sleep_analysis_participant %>%
  select(participant_id, macID, sleepStart, sleepEnd,
         sleepStart_time, sleepEnd_time, date_bedtime, date_awakening) %>%
  filter(
    format(sleepEnd_time,    "%H:%M") >= "03:00" &
    format(sleepEnd_time,    "%H:%M") <= "11:59" &
    (format(sleepStart_time, "%H:%M") >= "18:00" |
     format(sleepStart_time, "%H:%M") <= "02:59")
  )

n_sleep_removed <- nrow(sleep_analysis_participant) - nrow(sleep_analysis_filter)
message("Sleep records retained: ", nrow(sleep_analysis_filter),
        " | Removed (implausible windows): ", n_sleep_removed)
```

## Merge survey and sleep data

```{r Survey-merge-sleep}
# Use date_awakening as the merge key (matches date_converted in survey data)
sleep_analysis_filter <- sleep_analysis_filter %>%
  mutate(date_converted = date_awakening)

# Left join: every survey record is kept; sleep times added where available
survey_data_cleaned <- survey_data_cleaned %>%
  left_join(
    sleep_analysis_filter,
    by = c("macID", "participant_id", "date_converted")
  )
```

## Assign reference times

```{r Survey-reference-time}
# Assign reference_time per period:
#   AM -> awakening time (sleepEnd_time)
#   PM -> 15:00 on date_converted
#   NI -> bedtime (sleepStart_time)
# See PRG/Functions/functions_cleaning_data.R for full details
survey_sleep_merged <- set_reference_time(survey_data_cleaned)
```

```{r Survey-adjust-night-reference}
# Adjust reference_time for NI measurements recorded after midnight
# See PRG/Functions/functions_adjusted_ref_time.R for full details
survey_sleep_merged <- set_adjusted_night_time(survey_sleep_merged)
```

## Verify night reference time adjustment

```{r Survey-verify-night-reference}
# Compute time difference between NI responses and adjusted bedtime
survey_night_verif <- survey_sleep_merged %>%
  filter(period == "NI") %>%
  mutate(
    time_diff_minutes = as.numeric(
      difftime(unix_time_converted, adjusted_reference_time, units = "mins")
    )
  )

time_diff_summary <- survey_night_verif %>%
  summarise(
    mean_diff = mean(time_diff_minutes, na.rm = TRUE),
    sd_diff   = sd(time_diff_minutes,   na.rm = TRUE),
    min_diff  = min(time_diff_minutes,  na.rm = TRUE),
    max_diff  = max(time_diff_minutes,  na.rm = TRUE)
  )

message(
  "NI period -- time between survey response and adjusted bedtime (minutes):",
  "\n  Mean = ", round(time_diff_summary$mean_diff, 2),
  " | SD = ",   round(time_diff_summary$sd_diff,   2),
  " | Min = ",  round(time_diff_summary$min_diff,  2),
  " | Max = ",  round(time_diff_summary$max_diff,  2)
)
```
---

# Step 4 - Select one response per participant per key per period {#step4}

## Why standard deduplication functions cannot be used here

The standard functions `get_surrounding_indices()` and `remove_farthest_measure()` were designed for physiological data where adjacent rows in the dataset are duplicate measurements of the **same variable**. For survey data, adjacent rows belong to **different keys** of the same questionnaire instance (e.g., `sensation` at 08:09 is followed by `comfort` at 08:09, not by a duplicate `sensation`).

Applying the standard approach would incorrectly group `sensation` (09:01) with `comfort` (09:01) instead of comparing the two `sensation` responses (08:09 vs 09:01).

**Solution:** group explicitly by `participant_id`, `date_converted`, `period`, and `key`, then retain the response with the smallest absolute time difference from the reference time. A `slice(1)` tiebreaker handles the rare case of two responses equidistant from the reference time.

```{r Survey-select-reference-response}
# Ungroup before applying the custom selection logic
survey_sleep_merged <- survey_sleep_merged %>%
  ungroup()

final_survey_data <- survey_sleep_merged %>%
  arrange(participant_id, date_converted, period, key, unix) %>%
  group_by(participant_id, date_converted, period, key) %>%
  mutate(
    # Use adjusted_reference_time for NI period (accounts for post-midnight shift)
    # Use reference_time for AM and PM
    ref_time_to_use = if_else(
      period == "NI" & !is.na(adjusted_reference_time),
      adjusted_reference_time,
      reference_time
    ),
    # Compute absolute time difference between response and reference time (seconds)
    abs_time_diff = abs(as.numeric(
      difftime(unix_time_converted, ref_time_to_use, units = "secs")
    ))
  ) %>%
  # Keep only the response closest to the reference time
  filter(abs_time_diff == min(abs_time_diff, na.rm = TRUE)) %>%
  # Tiebreaker: if two responses are equidistant, keep the first (earliest)
  dplyr::slice(1) %>%
  ungroup() %>%
  # Remove intermediate helper columns
  select(-abs_time_diff, -ref_time_to_use)
```

## Final duplicate verification

```{r Survey-final-duplicate-check}
# Confirm that no duplicates remain after reference-time selection
remaining_duplicates_final <- final_survey_data %>%
  group_by(participant_id, date_converted, period, key) %>%
  summarise(n_responses = n(), .groups = "drop") %>%
  filter(n_responses > 1)

message("=== Final duplicate check ===")
message("N cases with > 1 response per key/period/day/participant: ",
        nrow(remaining_duplicates_final))

if (nrow(remaining_duplicates_final) > 0) {
  message("WARNING: Duplicates remain -- investigate before export")
  remaining_duplicates_final %>% count(key) %>% print()
  remaining_duplicates_final %>% count(participant_id) %>% print()
} else {
  message("Passed: one response per participant per key per period per day.")
}
```

---

# Step 5 - QC heatmap: duplicate survey keys per period {#step5}

Visualize the number of duplicate survey keys per participant per period per day before and after cleaning, to confirm that deduplication was effective.

```{r Survey-heatmap, echo=FALSE}
all_participants <- sort(unique(final_survey_data$participant_id))
n_per_page       <- 4
n_pages          <- ceiling(length(all_participants) / n_per_page)

# Store original data before loop to prevent accidental overwrite
original_data <- final_survey_data %>%
  arrange(participant_id, date_converted, period) %>%
  mutate(
    heatwave_status = factor(heatwave_status,
                             levels = c("BEF_HW_June", "HW_June", "AFT_HW_June")),
    period          = factor(period, levels = c("AM", "PM", "NI"))
  )

for (i in seq_len(n_pages)) {
  start_idx         <- (i - 1) * n_per_page + 1
  end_idx           <- min(i * n_per_page, length(all_participants))
  participants_page <- all_participants[start_idx:end_idx]

  # Use a temporary subset -- never overwrite original_data inside the loop
  heatmap_data <- original_data %>%
    filter(participant_id %in% participants_page) %>%
    group_by(participant_id, date_converted, period, key) %>%
    summarise(
      n_responses = n(),   # count responses per key/period/day/participant
      .groups = "drop"
    ) %>%
    group_by(participant_id, date_converted, period) %>%
    summarise(
      # Number of keys that still have more than one response (should be 0 after cleaning)
      n_duplicates = sum(n_responses > 1),
      .groups = "drop"
    )

  heatmap_indiv <- ggplot(
    heatmap_data,
    aes(x = date_converted, y = period, fill = n_duplicates)
  ) +
    geom_tile(color = "grey") +
    scale_fill_gradient(
      low  = "white",    # white = no duplicates remaining
      high = "#E63946",  # red = duplicates still present
      name = "N duplicate\nkeys"
    ) +
    facet_wrap(~ participant_id, nrow = 2, ncol = 2) +
    labs(title = paste("Duplicate survey keys after cleaning -- page", i),
         x     = "",
         y     = "Period of the day") +
    theme_minimal() +
    theme(legend.position = "bottom")

  print(heatmap_indiv)
}
```


---

# Step 6 - Final dataset and export {#step6}

## Remove intermediate processing columns

```{r Survey-clean-final}
# Remove columns used only for QC and intermediate processing
# Note: time_diff_minutes was computed on a filtered NI subset only
#       and is NOT a column in final_survey_data -- do not attempt to drop it
final_survey_data <- final_survey_data %>%
  select(
    -survey_time, -unix, -id,
    -sleepStart_time, -sleepEnd_time,
    -date_bedtime, -date_awakening,
    -sleepStart, -sleepEnd,
    -hours_discrepancy, -adjusted_reference_time,
    -duplicate_flag, -reference_time,
    -n_responses, -response_rank, -time_diff_key, -local_time,-local_date,-time_diff_minutes
  )
```

## Export processed dataset

```{r Survey-export}
# Export clean survey dataset to DAT/Processed/ for use in perceptual analyses
output_file <- file.path(path_res, "survey_processed.csv")
write.csv(final_survey_data, output_file, row.names = FALSE)
message("Exported: survey_processed.csv (",
        nrow(final_survey_data), " rows x ",
        ncol(final_survey_data), " columns)")
```

---

# Session information

```{r session-info}
sessionInfo()
```


