---
title: "Blood Pressure — HeatSuite 2024"
author: "Souilla Luc"
date: "2026-03-04"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
encoding: UTF-8
---

<!--
================================================================================
SCRIPT:  PRG/Physiological/blood_pressure.Rmd
PURPOSE: Full data preparation pipeline for  blood pressure and heart rate collected via blood pressure monitors by participant during the Montreal heatwave in 2024. This script:
           1. Assigns participant IDs and study periods to raw BP data
           2. Flags and removes error and implausible measurements
           3. Removes duplicate measurements within 5-minute windows
           4. Merges BP data with sleep analysis (awakening / bedtime)
           5. Selects one reference measurement per participant per period (AM/PM/NI)
           6. Computes mean arterial pressure (MAP)
           7. Exports the clean final dataset and QC plots

INPUT FILES (DAT/):
  - blood_pressure.csv              → raw smartwatch BP measurements
  - baseline_long.csv               → participant–macID–date lookup (from Data_preparation.Rmd)
  - sleep_analysis_daily.csv        → daily sleep start/end times per participant

OUTPUT FILES (RES/Physiological/):
  - blood_pressure_processed.csv    → cleaned, one-measurement-per-period BP dataset

DEPENDENCIES:
  - PRG/Functions/functions_participant_period.R   (assign_study_period,
                            add_participant_id_watch,assign_heatwave)
  - PRG/Functions/functions_cleaning_data.R        (set_reference_time)
  - PRG/Functions/functions_adjusted_ref_time. (set_adjusted_night_time)

IMPLAUSIBLE VALUE THRESHOLDS (clinical guidelines):
  - SBP:  70 – 240 mmHg  (flagging); 75 – 200 mmHg (removal)
  - DBP:  40 – 150 mmHg  (flagging); 44 – 135 mmHg (removal)
  - HR:   30 – 220 bpm   (flagging); 40 – 140 bpm  (removal)

TIMEZONE NOTE:
  All Unix timestamps are stored in UTC. Montreal operates on EDT (UTC-4) during the study period (June 2024).

================================================================================
-->

```{r setup, include=FALSE}
# Global knitr options: show code, suppress package messages and warnings
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE
)
```

# Overview

This script prepares the **blood pressure (BP)** and **heart rate (HR)** dataset collected during the Montreal heatwave 2024.
It follows the three-step pipeline defined in `Data_preparation.Rmd`:

1. **`blood_pressure_raw`** — raw upload and participant ID assignment
2. **`blood_pressure_participant`** — error flagging, implausible value removal, deduplication
3. **`final_blood_pressure_data`** — reference-time selection, MAP computation, export processed data in DAT/ Processed folder

---

# Environment setup

## Load packages

> **Note:** This script can be run standalone. All required packages are loaded here explicitly. If running after `Data_preparation.Rmd` in the same R session,packages are already loaded and this step is skipped automatically.

```{r load-packages, message=FALSE, warning=FALSE, include=FALSE}
required_packages <- c(
  "here", "dplyr", "tidyr", "lubridate",
  "ggplot2", "patchwork", "purrr"
)

install_if_missing <- function(packages) {
  missing <- packages[!(packages %in% installed.packages()[, "Package"])]
  if (length(missing) > 0) install.packages(missing)
}

install_if_missing(required_packages)
invisible(lapply(required_packages, library, character.only = TRUE))
```


## Set project root and directory paths

```{r set-paths}
# Anchor here() to the project root from this script's location
here::i_am("PRG/Physiological/blood_pressure.Rmd")

# Define paths using 'here' to get project-relative paths
path_dat <- here("DAT")                  # Raw data folder
path_prg <- here("PRG")                  # Scripts folder
path_res <- here("DAT/Processed")       # Output folder for data processed
```

## Source custom functions

All custom functions are loaded upfront for clarity and to allow standalone execution of this script. See each `.R` file in `PRG/Functions/` for full documentation and parameter descriptions.

```{r source-functions}
source(here("PRG/Functions/functions_participant_period.R"))  # assign_study_period, add_participant_id_watch, assign_heatwave
source(here("PRG/Functions/functions_cleaning_data.R"))       # set_reference_time
source(here("PRG/Functions/functions_adjusted_ref_time.R"))   # set_adjusted_night_time
```

---

# Step 1 — Load raw data and assign participant IDs {#step1}

## Load raw blood pressure data

```{r load-bp-raw}
# Load raw blood pressure CSV exported from smartwatch platform
# sep = ";" because the device software exports with semicolon delimiter
blood_pressure_raw <- read.csv(
  file.path(path_dat, "blood_pressure.csv"),
  header = TRUE, sep = ";", quote = "\"'", dec = "."
)
```

## Load participant–macID lookup table

```{r load-baseline-long}
# baseline_long was created by Data_preparation.Rmd
# It maps each macID + date range to a participant_id
baseline_long <- read.csv(
  file.path(path_dat, "baseline_long.csv"),
  header = TRUE, sep = ",", quote = "\"'", dec = "."
)
```


## Assign study period, participant ID, and heatwave status


We need to add period, participant ID and dates of heatwave to the blood pressure data table. We will use the function `functions_participant_period.R` script to do so. 

```{r assign-participant-period}
# Step 1a: Convert unix timestamp and assign period of day (AM / PM / NI)
blood_pressure_participant <- assign_study_period(blood_pressure_raw, c("unix"))

# Step 1b: Match each row to a participant_id using macID + local_time window
blood_pressure_participant <- add_participant_id_watch(
  blood_pressure_participant,
  baseline_long,
  mac_col  = "macID",
  time_col = "local_time"
)

# Step 1c: Assign heatwave status (BEF_HW_June / HW_June / AFT_HW_June)
# based on Montreal Public Health definitions (see Data_preparation.Rmd)
blood_pressure_participant <- assign_heatwave(blood_pressure_participant)

# Keep only rows within the heatwave monitoring window
blood_pressure_participant <- blood_pressure_participant %>%
  filter(heatwave_status %in% c("BEF_HW_June", "HW_June", "AFT_HW_June"))
```


---

# Step 2 — Quality control {#step2}

## Flag measurement errors

Device-reported error flags are combined into a single `error_measurement`indicator. The four source flags are: `cuffLoose`, `irregularPulse`, `moved`, and `improperMeasure`.

```{r flag-error-measurement}
# Create composite error flag: 1 if any device error flag is active
blood_pressure_participant <- blood_pressure_participant %>%
  mutate(
    error_measurement = ifelse(
      cuffLoose == 1 | irregularPulse == 1 | moved == 1 | improperMeasure == 1,
      1, 0
    )
  )

# Compute and report error counts
total_n               <- nrow(blood_pressure_participant)
total_n_error         <- sum(blood_pressure_participant$error_measurement)
pct_error             <- round((total_n_error / total_n) * 100, 2)
total_n_cuffLoose     <- sum(blood_pressure_participant$cuffLoose)
total_n_irregPulse    <- sum(blood_pressure_participant$irregularPulse)
total_n_moved         <- sum(blood_pressure_participant$moved)
total_n_improper      <- sum(blood_pressure_participant$improperMeasure)

message("Error measurements: ", total_n_error, " / ", total_n, " (", pct_error, "%)")
message("  cuffLoose=", total_n_cuffLoose,
        " | irregularPulse=", total_n_irregPulse,
        " | moved=", total_n_moved,
        " | improperMeasure=", total_n_improper)
```

## Flag implausible values

Implausible values are flagged based on clinical thresholds.
Note: a fixed value of 255 mmHg appears in SBP, DBP, and HR when the monitor first connects to the smartwatch — this is a known device artefact.

Flagging thresholds:

| Variable | Lower | Upper |
|---|---|---|
| SBP | < 70 mmHg | > 240 mmHg |
| DBP | < 40 mmHg | > 150 mmHg |
| HR  | < 30 bpm  | > 220 bpm  |


```{r flag-implausible-values}
blood_pressure_participant <- blood_pressure_participant %>%
  mutate(
    implausible_systolic    = ifelse(sbp < 70  | sbp > 240, 1, 0),
    implausible_diastolic   = ifelse(dbp < 40  | dbp > 150, 1, 0),
    implausible_hr          = ifelse(hr  < 30  | hr  > 220, 1, 0),
    implausible_measurement = ifelse(
      implausible_systolic == 1 | implausible_diastolic == 1 | implausible_hr == 1,
      1, 0
    )
  )

total_n_implausible <- sum(blood_pressure_participant$implausible_measurement)
pct_implausible     <- round((total_n_implausible / total_n) * 100, 2)
message("Implausible measurements: ", total_n_implausible, " / ", total_n,
        " (", pct_implausible, "%)")
```

## Plot distributions before filtering

Visualize SBP, DBP, and HR over time, color-coded by measurement type
(Valid / Implausible / Error), to visually confirm the flagging before removal.


```{r plot-before-filtering}
# Prepare plotting dataset with combined measurement type factor
bp_data_before <- blood_pressure_participant %>%
  mutate(
    measurement_type = case_when(
      error_measurement       == 1 ~ "Error",
      implausible_measurement == 1 ~ "Implausible",
      TRUE                         ~ "Valid"
    ),
    measurement_type = factor(measurement_type, levels = c("Valid", "Implausible", "Error")),
    hour_of_day = hour(unix_time_converted)
  )

# Color and shape mapping for measurement quality
flag_colors <- c("Valid" = "#2E86AB", "Implausible" = "#F6AE2D", "Error" = "#E63946")
flag_shapes <- c("Valid" = 16,        "Implausible" = 17,        "Error" = 4)

# SBP over time
line_sbp_complete_bef <- ggplot(bp_data_before, aes(x = date_converted, y = sbp)) +
  geom_jitter(aes(color = measurement_type, shape = measurement_type),
              size = 0.8, alpha = 0.85, width = 0.2, height = 0) +
  scale_color_manual(values = flag_colors, name = "Type") +
  scale_shape_manual(values = flag_shapes, name = "Type") +
  labs(title = "Before filtering", x = "Date of measurement", y = "Systolic BP (mmHg)") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5), legend.position = "right",
        axis.text.x = element_text(angle = 45, hjust = 1, size = 10))
print(line_sbp_complete_bef)

# DBP over time
line_dbp_complete_bef <- ggplot(bp_data_before, aes(x = date_converted, y = dbp)) +
  geom_jitter(aes(color = measurement_type, shape = measurement_type),
              size = 0.8, alpha = 0.85, width = 0.2, height = 0) +
  scale_color_manual(values = flag_colors, name = "Type") +
  scale_shape_manual(values = flag_shapes, name = "Type") +
  labs(title = "Before filtering", x = "Date of measurement", y = "Diastolic BP (mmHg)") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5), legend.position = "right",
        axis.text.x = element_text(angle = 45, hjust = 1, size = 10))
print(line_dbp_complete_bef)

# HR over time
line_hr_complete_bef <- ggplot(bp_data_before, aes(x = date_converted, y = hr)) +
  geom_jitter(aes(color = measurement_type, shape = measurement_type),
              size = 0.8, alpha = 0.85, width = 0.2, height = 0) +
  scale_color_manual(values = flag_colors, name = "Type") +
  scale_shape_manual(values = flag_shapes, name = "Type") +
  labs(title = "Before filtering", x = "Date of measurement", y = "Heart rate (bpm)") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5), legend.position = "right",
        axis.text.x = element_text(angle = 45, hjust = 1, size = 10))
print(line_hr_complete_bef)
```

## Remove flagged values

Visual inspection confirms that most irregular pulse flags fall within plausible ranges and are retained. 

| Variable | Removal threshold |
|---|---|
| SBP | < 75 or > 200 mmHg |
| DBP | < 44 or > 135 mmHg |
| HR  | < 40 or > 140 bpm  |

```{r remove-outliers}
cleaned_bp_data <- blood_pressure_participant %>%
  arrange(macID, period, date_converted, unix) %>%
  group_by(macID, period, date_converted) %>%
  filter(
    sbp >= 75  & sbp <= 200,   # Refined SBP removal bounds
    dbp >= 44  & dbp <= 135,   # Refined DBP removal bounds
    hr  >= 40  & hr  <= 140    # Refined HR removal bounds
  ) %>%
  ungroup()

n_removed_outliers <- nrow(blood_pressure_participant) - nrow(cleaned_bp_data)
message("Rows removed (outliers / error measurements): ", n_removed_outliers)
```

## Remove duplicate measurements within 5-minute windows

Participants could press the measurement button multiple times in quick succession.
We group consecutive measurements within a 300-second (5-minute) window and keep only the **last** measurement in each group, as it best reflects the participant's settled resting state after device placement.
```{r remove-rapid-duplicates}
 cleaned_bp_data <- cleaned_bp_data %>%
  arrange(macID, period, date_converted, unix) %>%
  group_by(macID, period, date_converted) %>%
  mutate(
    # Time elapsed since the previous measurement (seconds)
    time_diff = unix - lag(unix, default = first(unix)),
    # Increment group counter each time a gap >= 300 seconds occurs
    group_row = cumsum(is.na(time_diff) | time_diff >= 300)
  ) %>%
  group_by(macID, period, date_converted, group_row) %>%
  # Within each 5-minute window, keep only the last (most settled) measurement
  slice_tail(n = 1) %>%
  ungroup()

n_removed_dupl <- nrow(blood_pressure_participant) -
                  nrow(cleaned_bp_data) - n_removed_outliers
message("Additional rows removed (rapid duplicates within 5-min window): ",
        n_removed_dupl)
```

---

# Step 3 — Merge with sleep data for reference-time selection {#step3}

After deduplication, some participants still have multiple measurements per period per day (e.g., two morning readings on the same date). We resolve thisby selecting the measurement **closest to a physiologically meaningful reference time**:

- **AM (morning):** closest to awakening time (from sleep analysis)
- **PM (afternoon):** closest to 15:00 (3 PM)
- **NI (night):** closest to bedtime (from sleep analysis)

## Load and prepare sleep analysis data

```{r load-sleep}
# Load daily sleep analysis exported from smartwatch platform
sleep_analysis_raw <- read.csv(
  file.path(path_dat, "sleep_analysis_daily.csv"),
  header = TRUE, sep = ",", quote = "\"'", dec = "."
)
```

First, we will load the the file sleep analysis from folder "DAT" to get awakening time and bedtime.
```{r Sleep-upload}
file_path_baseline<-file.path(path_dat,"sleep_analysis_daily.csv") #Charging file in DAT folder, make sure the name of the file mathcing with this one
sleep_analysis_raw<- read.csv(file_path_baseline, header = TRUE, sep = ",", quote = "\"'", dec = ".")
```

## Assign participant ID to sleep data

The sleep data contains unix timestamps but no `local_time` column. We derived `local_time` from `sleepEnd` (awakening time) using the EDT offset (-14400 s = UTC-4).

```{r assign-participant-sleep}
# Create local_time from sleepEnd unix timestamp for participant ID matching
sleep_analysis_raw <- sleep_analysis_raw %>%
  mutate(
    local_time = as.Date(
      as.POSIXct(sleepEnd, origin = "1970-01-01", tz = "UTC") - 14400  # EDT offset
    )
  )

# Match each sleep record to a participant_id using macID + local_time
sleep_analysis_participant <- add_participant_id_watch(
  sleep_analysis_raw,
  baseline_long,
  mac_col  = "macID",
  time_col = "local_time"
)
```

## Filter sleep data to study period and valid sleep windows

```{r filter-sleep}
# Step 1: Convert Unix timestamps to POSIXct and extract date components
sleep_analysis_participant <- sleep_analysis_participant %>%
  mutate(
    # Apply EDT offset (-14400 s) to convert UTC unix to local Montreal time
    sleepStart_time = as.POSIXct(sleepStart, origin = "1970-01-01", tz = "UTC") - 14400,
    sleepEnd_time   = as.POSIXct(sleepEnd,   origin = "1970-01-01", tz = "UTC") - 14400,
    date_bedtime    = as.Date(sleepStart_time),
    date_awakening  = as.Date(sleepEnd_time)
  ) %>%
  # Restrict to the heatwave monitoring window (June 9–29, 2024)
  filter(
    (sleepStart_time >= as.POSIXct("2024-06-09 00:00:00") &
     sleepStart_time <= as.POSIXct("2024-06-29 23:59:59")) |
    (sleepEnd_time   >= as.POSIXct("2024-06-09 00:00:00") &
     sleepEnd_time   <= as.POSIXct("2024-06-29 23:59:59"))
  )

# Step 2: Retain only physiologically plausible sleep episodes:
#   - Awakening (sleepEnd) between 03:00 and 11:59 AM
#   - Bedtime (sleepStart) between 18:00 and 02:59 (same or next day)
sleep_analysis_filter <- sleep_analysis_participant %>%
  select(participant_id, macID, sleepStart, sleepEnd,
         sleepStart_time, sleepEnd_time, date_bedtime, date_awakening) %>%
  filter(
    format(sleepEnd_time,   "%H:%M") >= "03:00" &
    format(sleepEnd_time,   "%H:%M") <= "11:59" &
    (format(sleepStart_time, "%H:%M") >= "18:00" |
     format(sleepStart_time, "%H:%M") <= "02:59")
  )

n_sleep_removed <- nrow(sleep_analysis_participant) - nrow(sleep_analysis_filter)
message("Sleep records retained: ", nrow(sleep_analysis_filter),
        " | Removed (implausible sleep windows): ", n_sleep_removed)
```


## Merge blood pressure and sleep data

```{r merge-bp-sleep}
# Align date keys: use date_awakening as the merge date (matches date_converted in BP data)
sleep_analysis_filter <- sleep_analysis_filter %>%
  mutate(date_converted = date_awakening)

# Left join: every BP record is kept; sleep times added where available
bp_sleep_merged <- cleaned_bp_data %>%
  left_join(
    sleep_analysis_filter,
    by = c("macID", "participant_id", "date_converted")
  )

# Remove columns used only for QC flagging and intermediate processing
bp_sleep_merged <- bp_sleep_merged %>%
  select(
    -moved, -cuffLoose, -irregularPulse, -improperMeasure,
    -implausible_systolic, -implausible_diastolic, -implausible_hr,
    -implausible_measurement, -error_measurement,
    -tz, -X, -local_time
  )
```


---

# Step 4 — Assign reference times and select closest measurement {#step4}

## Set reference time per period
```{r set-reference-time}
# Assigns reference_time column based on period:
#   AM  → sleepEnd_time (awakening)
#   PM  → 15:00 on date_converted
#   NI  → sleepStart_time (bedtime)
# See PRG/Functions/functions_cleaning_data.R for full details
bp_sleep_merged <- set_reference_time(bp_sleep_merged)
```

## Adjust night reference time for post-midnight measurements

Some NI measurements are recorded after midnight (next calendar day).
`set_adjusted_night_time()` corrects the reference time for these cases.
```{r adjust-night-reference-time}
# Adjusts reference_time for NI period when measurement falls past midnight
# See PRG/Functions/functions_adjusted_ref_time.R for full details
bp_sleep_merged <- set_adjusted_night_time(bp_sleep_merged)
```

## Verify night reference time adjustment
```{r verify-night-reference-time}
# Compute time difference (in minutes) between NI measurements and adjusted bedtime
bp_night_verif <- bp_sleep_merged %>%
  filter(period == "NI") %>%
  mutate(
    time_diff_minutes = as.numeric(
      difftime(unix_time_converted, adjusted_reference_time, units = "mins")
    )
  )

time_diff_summary <- bp_night_verif %>%
  summarise(
    mean_diff = mean(time_diff_minutes, na.rm = TRUE),
    sd_diff   = sd(time_diff_minutes,   na.rm = TRUE),
    min_diff  = min(time_diff_minutes,  na.rm = TRUE),
    max_diff  = max(time_diff_minutes,  na.rm = TRUE)
  )

message(
  "NI period — time between measurement and adjusted bedtime (minutes):",
  "\n  Mean = ", round(time_diff_summary$mean_diff, 2),
  " | SD = ",   round(time_diff_summary$sd_diff,   2),
  " | Min = ",  round(time_diff_summary$min_diff,  2),
  " | Max = ",  round(time_diff_summary$max_diff,  2)
)
```

## Select one measurement per participant per period

When multiple measurements remain for the same participant, date, and period, we retain the one with the **smallest absolute time difference** from the reference time.

For example: three morning readings at 08:45, 09:15, and 10:15 with awakening
at 08:30 → the 08:45 reading is kept.


```{r select-reference-measurement, include=FALSE}
# Number each measurement per group in DESCENDING unix order
# row_num = 1 → most recent measurement in the group
#create row_num which numbered the measurement per period per participant 
bp_sleep_merged <- bp_sleep_merged %>%
  arrange(macID, period, date_converted, unix) %>%
  group_by(macID, period, date_converted, participant_id) %>%
  mutate(row_num = row_number(desc(unix))) %>%
  ungroup()

# Compute absolute time difference from reference_time for each measurement
bp_sleep_merged <- bp_sleep_merged %>%
  group_by(participant_id, date_converted, period) %>%
  mutate(
    abs_time_diff = abs(difftime(unix_time_converted, reference_time, units = "secs"))
  ) %>%
  ungroup()

# Identify indices of rows where more than one measurement exists per group
indices_row <- which(bp_sleep_merged$row_num > 1)
message("Groups with more than one measurement: ", length(indices_row), " rows")

# Retrieve surrounding indices and remove the farthest measurement per group
surrounding_indices_list <- get_surrounding_indices(indices_row, bp_sleep_merged)
final_blood_pressure_data <- remove_farthest_measure(bp_sleep_merged, surrounding_indices_list)
```

---

# Step 5 — Final dataset: compute MAP and export {#step5}

## Check for remaining duplicates

```{r check-duplicates,  message=FALSE, warning=FALSE}

# Verify that no participant has more than one measurement per period per date
duplicates_check <- final_blood_pressure_data %>%
  group_by(period, macID, date_converted) %>%
  summarise(duplicate_count = n(), .groups = "drop") %>%
  filter(duplicate_count > 1)

if (nrow(duplicates_check) > 0) {
  message("WARNING: Duplicates remain for the following groups:")
  print(duplicates_check)
} else {
  message("Duplicate check passed: one measurement per participant per period per date.")
}

```

## Compute mean arterial pressure and clean final columns

```{r finalize-dataset}
# Compute MAP using the standard clinical formula: MAP = 1/3 SBP + 2/3 DBP
# Remove intermediate processing columns not needed for analysis
final_blood_pressure_data <- final_blood_pressure_data %>%
  mutate(
    map = round((1/3 * sbp + 2/3 * dbp), 1)   # Mean arterial pressure (mmHg)
  ) %>%
  select(
    -row_num, -abs_time_diff,
    -unix, -id,
    -sleepStart_time, -sleepEnd_time,
    -date_bedtime, -date_awakening,
    -sleepStart, -sleepEnd,
    -hours_discrepancy, -adjusted_reference_time, -local_date, -time_diff, -time_diff_minutes, -group_row, -reference_time
  )
```

## Export processed dataset
```{r export-final}
# Export clean dataset to RES/Physiological/ for use in statistical analysis
output_file <- file.path(path_res, "blood_pressure_processed.csv")
write.csv(final_blood_pressure_data, output_file, row.names = FALSE)
```

---

# Step 6 — Quality control plots after filtering {#step6}

## Time series: SBP, DBP, HR by period of day


```{r plot-timeseries-after, echo=FALSE}
line_sbp_aft <- ggplot(final_blood_pressure_data, aes(x = date_converted, y = sbp, color = period)) +
  geom_jitter(alpha = 0.8, size = 0.8, width = 0.2) +
  labs(title = "After filtering", x = "", y = "Systolic BP (mmHg)", color = "Period of day") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5), legend.position = "bottom")
print(line_sbp_aft)

line_dbp_aft <- ggplot(final_blood_pressure_data, aes(x = date_converted, y = dbp, color = period)) +
  geom_jitter(alpha = 0.8, size = 0.8, width = 0.2) +
  labs(title = "After filtering", x = "", y = "Diastolic BP (mmHg)", color = "Period of day") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5), legend.position = "bottom")
print(line_dbp_aft)

line_hr_aft <- ggplot(final_blood_pressure_data, aes(x = date_converted, y = hr, color = period)) +
  geom_jitter(alpha = 0.8, size = 0.8, width = 0.2) +
  labs(title = "After filtering", x = "Date of measurement", y = "Heart rate (bpm)", color = "Period of day") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5), legend.position = "bottom")
print(line_hr_aft)
```

## Histograms: distributions after filtering

```{r plot-histograms-after, echo=FALSE}
histo_sbp_aft <- ggplot(final_blood_pressure_data, aes(x = sbp)) +
  geom_histogram(aes(y = after_stat(density)), fill = "lightblue", color = "grey", bins = 30) +
  labs(title = "After filtering", x = "Systolic blood pressure (mmHg)", y = "Density") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))
print(histo_sbp_aft)

histo_dbp_aft <- ggplot(final_blood_pressure_data, aes(x = dbp)) +
  geom_histogram(aes(y = after_stat(density)), fill = "lightblue", color = "grey", bins = 30) +
  labs(title = "After filtering", x = "Diastolic blood pressure (mmHg)", y = "Density") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))
print(histo_dbp_aft)

histo_hr_aft <- ggplot(final_blood_pressure_data, aes(x = hr)) +
  geom_histogram(aes(y = after_stat(density)), fill = "lightblue", color = "grey", bins = 30) +
  labs(title = "After filtering", x = "Heart rate (bpm)", y = "Density") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))
print(histo_hr_aft)
```

## Boxplots by period of day
```{r plot-boxplots-after, echo=FALSE}
bxp_sbp_aft <- ggplot(final_blood_pressure_data, aes(x = period, y = sbp, fill = period)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.8, width = 0.5) +
  geom_jitter(width = 0.2, alpha = 0.4, size = 1, color = "grey2") +
  coord_cartesian(ylim = c(60, 200)) +
  labs(title = "After filtering", x = "", y = "Systolic BP (mmHg)") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5), legend.position = "bottom")
print(bxp_sbp_aft)

bxp_dbp_aft <- ggplot(final_blood_pressure_data, aes(x = period, y = dbp, fill = period)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.8, width = 0.5) +
  geom_jitter(width = 0.2, alpha = 0.4, size = 1, color = "grey2") +
  coord_cartesian(ylim = c(30, 155)) +
  labs(title = "After filtering", x = "", y = "Diastolic BP (mmHg)") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5), legend.position = "bottom")
print(bxp_dbp_aft)

bxp_hr_aft <- ggplot(final_blood_pressure_data, aes(x = period, y = hr, fill = period)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.8, width = 0.5) +
  geom_jitter(width = 0.2, alpha = 0.4, size = 1, color = "grey2") +
  coord_cartesian(ylim = c(30, 200)) +
  labs(title = "After filtering", x = "", y = "Heart rate (bpm)") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5), legend.position = "bottom")
print(bxp_hr_aft)
```


## Heatmap: measurement count per participant per period per day

Used to verify that the deduplication pipeline achieved one measurement per
participant per period per date across all participants.
```{r plot-heatmap-individual, echo=FALSE}
all_participants <- sort(unique(final_blood_pressure_data$participant_id))
n_per_page       <- 4   # Display 4 participants per page (2×2 grid)
n_pages          <- ceiling(length(all_participants) / n_per_page)

# Store original data before loop to prevent accidental overwrite
original_data <- final_blood_pressure_data

for (i in seq_len(n_pages)) {
  start_idx          <- (i - 1) * n_per_page + 1
  end_idx            <- min(i * n_per_page, length(all_participants))
  participants_page  <- all_participants[start_idx:end_idx]

  # Use a temporary object — never overwrite the original data inside a loop
  plot_data <- original_data %>%
    filter(participant_id %in% participants_page) %>%
    group_by(participant_id, date_converted, period) %>%
    summarise(n_measurements = n(), .groups = "drop")

  heatmap_indiv <- ggplot(plot_data, aes(x = date_converted, y = period, fill = n_measurements)) +
    geom_tile(color = "grey") +
    scale_fill_gradient(low = "lightblue", high = "darkblue", name = "N measurements") +
    facet_wrap(~ participant_id, nrow = 2, ncol = 2) +
    labs(title = "BP measurements per period per day", x = "", y = "Period") +
    theme_minimal() +
    theme(legend.position = "bottom")
  print(heatmap_indiv)
}
```

## Before vs. after filtering comparison

```{r plot-before-after-comparison, echo=FALSE}
line_sbp_complete_bef / line_sbp_aft +
  plot_annotation(title = "Effect of data cleaning on systolic BP")

line_dbp_complete_bef / line_dbp_aft +
  plot_annotation(title = "Effect of data cleaning on diastolic BP")

line_hr_complete_bef / line_hr_aft +
  plot_annotation(title = "Effect of data cleaning on heart rate")
```

---

# Session information

```{r session-info}
sessionInfo()
```
