---
title: "Main Analysis Physiological HeatSuite 2024"
author: "Souilla Luc"
date: "2026-03-05"
output: 
 html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
encoding: UTF-8
---

<!--
================================================================================
SCRIPT:  PRG/Physiological/main_analysis_physio.Rmd
PURPOSE: Descriptive and inferential analysis of all physiological outcomes collected during the Montreal heatwave 2024. This script:
           1. Loads all processed physiological datasets
           2. Produces descriptive statistics tables (pooled and participant-level)
           3. Produces time series and boxplot visualizations
           4. Runs linear mixed models (MMRM) for two-factor and one-factor designs
           5. Explores the moderating effect of air conditioning (AC) access

INPUT FILES (DAT/Processed/):
  - blood_pressure_processed.csv           -> cleaned BP and HR (3x daily)
  - oral_temp_processed.csv                -> cleaned oral temperature (3x daily)
  - body_mass_processed.csv                -> cleaned body mass (daily, AM only)
  - urine_data_processed.csv               -> cleaned urine frequency (daily)
  - daily_physical_activity_processed.csv  -> cleaned mean daily acceleration (daily)

OUTPUT FILES (RES/Physiological/):
  - Table/table_pooled_two_ways_heatwave.xlsx -> pooled descriptive stats (two-factor)
  - Table/Table_two_ways_participant.xlsx     -> participant-weighted stats (two-factor)
  - Table/Table_one_way_participant.xlsx      -> participant-weighted stats (one-factor)
  - Graph/ (plots saved automatically by custom functions)

STUDY DESIGN:
  Two-factor outcomes (heatwave x period of day -- AM/PM/NI):
    Blood pressure (SBP, DBP, MAP, HR) and oral temperature -- measured 3x daily
  One-factor outcomes (heatwave only -- BEF/HW/AFT):
    Body mass, urine frequency, mean daily acceleration -- measured once daily

DESCRIPTIVE STATISTICS -- TWO APPROACHES:
  Option 1 (group_by_participant = FALSE): all observations pooled across participants before computing statistics. Participants with more measurements carry more weight. Used for population-level summaries.

  Option 2 (group_by_participant = TRUE): statistics computed per participant first, then averaged across participants. Each participant contributes equally regardless of number of measurements. Used for publication tables.

STATISTICAL APPROACH:
  Mixed Model Repeated Measures (MMRM) with AR(1) covariance structure.
  Fixed effects tested with car::Anova() Type II (Kenward-Roger df).
  Post-hoc comparisons with emmeans() using mvt adjustment.

AC STATUS NOTE:
  Air conditioning status derived from participant self-report at baseline. Hardcoded list verified against characteristics_participants.csv. Update manually if study population changes.

HEATWAVE PERIODS:
  Before heatwave : 2024-06-11 to 2024-06-17
  During heatwave : 2024-06-18 to 2024-06-20
  After heatwave  : 2024-06-21 to 2024-06-27

================================================================================
-->

```{r setup, include=FALSE}
# Global knitr options: show code, suppress messages and warnings
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE
)
```

# Overview

This script performs descriptive and inferential analyses of all physiological outcomes measured during the Montreal heatwave 2024.

Outcomes are organized into two analytical designs:

- **Two-factor design** (heatwave x period of day): blood pressure, heart rate, oral temperature -- measured three times daily (AM/PM/NI)
- **One-factor design** (heatwave only): body mass, urine frequency, physical activity -- measured once daily

---

# Environment setup

## Load packages

```{r load-packages, message=FALSE, warning=FALSE, include=FALSE}
# Required packages -- duplicates removed (lubridate, stringr)
required_packages <- c(
  # Data wrangling
  "here", "dplyr", "tidyr", "lubridate", "stringr", "purrr",
  # Reporting
  "knitr", "kableExtra", "openxlsx",
  # Visualization
  "ggplot2", "patchwork", "viridis", "ggeffects",
  # Statistical modelling
  "rstatix", "carData", "car", "emmeans", "mmrm",
  "mlogit", "ordinal", "glmmTMB"
)

# Install any missing packages automatically
install_if_missing <- function(packages) {
  missing_packages <- packages[!(packages %in% installed.packages()[, "Package"])]
  if (length(missing_packages) > 0) {
    install.packages(missing_packages)
  }
}

# Load all packages silently
load_packages <- function(packages) {
  lapply(packages, library, character.only = TRUE)
}

install_if_missing(required_packages)
load_packages(required_packages)
```

## Set project root and directory paths

```{r set-paths, message=FALSE, warning=FALSE, include=FALSE}
# Anchor here() to the project root from this script's location
here::i_am("PRG/Physiological/main_analysis_physio.Rmd")
```

```{r define-paths}
# FIX: corrected capitalisation from "DAT/processed" to "DAT/Processed"
# to match actual folder name (case-sensitive on Mac/Linux)
path_dat <- here("DAT/Processed")       # Processed data inputs
path_prg <- here("PRG")                 # Scripts folder
path_res <- here("RES/Physiological")   # Results output folder
```

## Set global color palette

We define colors for plotting using the `viridis` package, which provides a color-blind-friendly palette applied globally to all ggplot2 figures.

```{r set-color-palette, echo=FALSE, warning=FALSE}
scale_color_discrete <- function(...) scale_color_viridis_d(...)
scale_fill_discrete  <- function(...) scale_fill_viridis_d(...)
```

## Source custom analysis functions

```{r source-functions}
# All custom functions used in this script
# See each .R file in PRG/Functions/ for full documentation and parameters
source(here("PRG/Functions/functions_participant_period.R"))
source(here("PRG/Functions/functions_cleaning_data.R"))
source(here("PRG/Functions/functions_adjusted_ref_time.R"))
source(here("PRG/Functions/functions_descriptive_stats.R"))  # descriptive_stats(), display_table()
source(here("PRG/Functions/functions_descriptive_plot.R"))   # time_series_*, boxplot_numeric()
source(here("PRG/Functions/functions_linear_models.R"))      # lmm_two_way(), lmm_one_way(), lmm_*_AC()
source(here("PRG/Functions/residual_check_mmrm.R")) # add R code inside this file to check MMRM assumptions and model fit for continuous outcomes
source(here("PRG/Functions/functions_check_clmm.R")) # add R code inside this file to check clmm assumptions and model fit for ordinal outcomes 
```


---

# Load processed data {#load-data}

```{r load-data}
# Blood pressure and heart rate -- three measurements per day (AM, PM, NI)
final_blood_pressure_data <- read.csv(
  file.path(path_dat, "blood_pressure_processed.csv"),
  header = TRUE, sep = ",", quote = "\"'", dec = "."
)

# Oral temperature -- three measurements per day (AM, PM, NI)
final_core_temp_data <- read.csv(
  file.path(path_dat, "oral_temp_processed.csv"),
  header = TRUE, sep = ",", quote = "\"'", dec = "."
)

# Body mass -- one measurement per day (morning only per protocol)
final_body_mass_data <- read.csv(
  file.path(path_dat, "body_mass_processed.csv"),
  header = TRUE, sep = ",", quote = "\"'", dec = "."
)

# Urine frequency -- one self-reported entry per day
final_urine_data <- read.csv(
  file.path(path_dat, "urine_data_processed.csv"),
  header = TRUE, sep = ",", quote = "\"'", dec = "."
)

# Mean daily acceleration -- one aggregated value per day
final_activity_data_day <- read.csv(
  file.path(path_dat, "daily_physical_activity_processed.csv"),
  header = TRUE, sep = ",", quote = "\"'", dec = "."
)
```


---

# Physiological descriptive tables {#descriptive-tables}

For both descriptive tables and plots, outcomes are organized into two input
lists: `data_list` (the processed dataframes) and `outcome_vars` (the columns
of interest). Two analytical designs are used based on measurement frequency:

- **Two-factor** (heatwave x period of day): outcomes measured 3x daily
- **One-factor** (heatwave only): outcomes measured once daily

The `descriptive_stats()` function computes mean, SD, median, and quantiles.
To modify descriptive statistics features, edit `PRG/Functions/functions_descriptive_stats.R`.

## Two-factor outcomes: blood pressure and oral temperature {#two-factor}

### Define input lists

```{r StatDesc-2-ways-list, echo=FALSE, message=FALSE, warning=FALSE}
# data_list: named list mapping keys to processed dataframes
# Each key is referenced by name in outcome_vars below
data_list_num_two_way <- list(
  bp_data   = final_blood_pressure_data,
  temp_data = final_core_temp_data
)

# outcome_vars: named list specifying each outcome variable
# Fields: data = key in data_list, var = column name,
#         date = date column, day_time = time column for visit ordering
outcome_vars_num_two_way <- list(
  "Systolic Blood Pressure (mmHg)"      = list(data = "bp_data",   var = "sbp",            date = "date_converted", day_time = "unix_time_converted"),
  "Diastolic Blood Pressure (mmHg)"     = list(data = "bp_data",   var = "dbp",            date = "date_converted", day_time = "unix_time_converted"),
  "Mean Arterial Pressure (mmHg)"       = list(data = "bp_data",   var = "map",            date = "date_converted", day_time = "unix_time_converted"),
  "Heart Rate (bpm)"                    = list(data = "bp_data",   var = "hr",             date = "date_converted", day_time = "unix_time_converted"),
  "Oral Temperature (deg C)"            = list(data = "temp_data", var = "temp_corrected", date = "date_converted", day_time = "unix_time_converted"),
  "Daily change\nOral Temperature (deg C)" = list(data = "temp_data", var = "delta_oral_temp", date = "date_converted", day_time = "unix_time_converted")
)
```

### Option 1 -- Population-level summary (pooled across participants)

In this approach, all observations are pooled before computing statistics. Participants with more measurements contribute more to the group summary. Useful for population-level description.

```{r StatDesc-2-ways-calcul}
# Group by heatwave period AND period of day -- pooled across participants
outcome_table_period <- descriptive_stats(
  data_list_num_two_way,
  outcome_vars_num_two_way,
  group_by_participant = FALSE,
  group_by             = "heatwave_period",
  categorical_grouping = FALSE,
  numeric_variable     = TRUE
)
display_table(outcome_table_period)

# Group by heatwave period only -- pooled across participants and periods of day
outcome_table_heatwave <- descriptive_stats(
  data_list_num_two_way,
  outcome_vars_num_two_way,
  group_by_participant = FALSE,
  group_by             = "heatwave",
  categorical_grouping = FALSE,
  numeric_variable     = TRUE
)
display_table(outcome_table_heatwave)
```

### Save Option 1 tables

```{r save-two-way-pooled-tables}
# Rename to preserve objects before next table generation
table_period   <- outcome_table_period
table_heatwave <- outcome_table_heatwave

# Export both tables to a single Excel workbook with separate sheets
wb <- createWorkbook()
addWorksheet(wb, "By_period")
writeData(wb, "By_period",   table_period$stats)
addWorksheet(wb, "By_heatwave")
writeData(wb, "By_heatwave", table_heatwave$stats)
saveWorkbook(
  wb,
  file      = here("RES/Physiological/Table/table_pooled_two_ways_heatwave.xlsx"),
  overwrite = TRUE
)
message("Exported: table_pooled_two_ways_heatwave.xlsx")
```

### Option 2 -- Participant-weighted summary (recommended for publication)

Here, statistics are first computed per participant, then averaged across participants. Each participant contributes equally regardless of how many measurements they have. This is the recommended approach for repeated-measures physiological data as it avoids bias from unequal measurement counts.

```{r StatDesc-2-ways-calcul-2}
# Group by heatwave x period of day -- participant-weighted
outcome_table_participant_hw_period <- descriptive_stats(
  data_list_num_two_way,
  outcome_vars_num_two_way,
  group_by_participant = TRUE,
  group_by             = "heatwave_period",
  categorical_grouping = FALSE,
  numeric_variable     = TRUE
)
display_table(outcome_table_participant_hw_period)

# Group by heatwave period only -- participant-weighted
outcome_table_participant_hw <- descriptive_stats(
  data_list_num_two_way,
  outcome_vars_num_two_way,
  group_by_participant = TRUE,
  group_by             = "heatwave",
  categorical_grouping = FALSE,
  numeric_variable     = TRUE
)
display_table(outcome_table_participant_hw)
```

### Save Option 2 tables

```{r save-two-way-participant-tables}
# Rename to preserve objects
outcome_table_participant_heatwave_period <- outcome_table_participant_hw_period
outcome_table_participant_heatwave        <- outcome_table_participant_hw

wb <- createWorkbook()
addWorksheet(wb, "By_heatwave_period")
writeData(wb, "By_heatwave_period", outcome_table_participant_heatwave_period$stats)
addWorksheet(wb, "By_heatwave")
writeData(wb, "By_heatwave",        outcome_table_participant_heatwave$stats)
saveWorkbook(
  wb,
  file      = here("RES/Physiological/Table/Table_two_ways_participant.xlsx"),
  overwrite = TRUE
)
message("Exported: Table_two_ways_participant.xlsx")
```


---

## One-factor outcomes: body mass, physical activity, urine {#one-factor}

### Define input lists

```{r StatDesc-1-way-list, echo=FALSE, message=FALSE, warning=FALSE}
# data_list: named list mapping keys to processed dataframes
data_list_num_one_way <- list(
  mass_data          = final_body_mass_data,
  total_movement_day = final_activity_data_day,
  urine_frequency    = final_urine_data
)

# outcome_vars: one entry per outcome of interest
# No day_time field needed -- these are daily outcomes with no period of day
outcome_vars_num_one_way <- list(
  "Body Mass (kg)"                = list(data = "mass_data",          var = "mass",           date = "date_converted"),
  "Daily change\nBody Mass (g)"   = list(data = "mass_data",          var = "delta_bodymass", date = "date_converted"),
  "Mean Daily Acceleration (mg)"  = list(data = "total_movement_day", var = "mean_acc_day",   date = "date_converted"),
  "Urine Frequency per Day"       = list(data = "urine_frequency",    var = "urineFreq",      date = "date_converted")
)
```

### Option 1 -- Population-level summary (pooled)

```{r StatDesc-1-ways-calcul}
# Group by heatwave period -- pooled across participants
outcome_table_one_way_heatwave <- descriptive_stats(
  data_list_num_one_way,
  outcome_vars_num_one_way,
  group_by_participant = FALSE,
  group_by             = "heatwave",
  categorical_grouping = FALSE,
  numeric_variable     = TRUE
)
display_table(outcome_table_one_way_heatwave)
```

### Option 2 -- Participant-weighted summary (recommended for publication)

```{r StatDesc-1-ways-calcul-2}
# Group by heatwave period -- participant-weighted (recommended)
outcome_table_one_way_participant <- descriptive_stats(
  data_list_num_one_way,
  outcome_vars_num_one_way,
  group_by_participant = TRUE,
  group_by             = "heatwave",
  categorical_grouping = FALSE,
  numeric_variable     = TRUE
)
display_table(outcome_table_one_way_participant)
```

### Save one-factor tables

```{r save-one-way-tables}
# Rename to preserve objects
table_one_way_heatwave           <- outcome_table_one_way_heatwave
table_one_way_participant_heatwave <- outcome_table_one_way_participant

# Export to Excel: By_heatwave = pooled, By_heatwave_participant = weighted
wb <- createWorkbook()
addWorksheet(wb, "By_heatwave")
writeData(wb, "By_heatwave",             table_one_way_heatwave$stats)
addWorksheet(wb, "By_heatwave_participant")
writeData(wb, "By_heatwave_participant", table_one_way_participant_heatwave$stats)
saveWorkbook(
  wb,
  file      = here("RES/Physiological/Table/Table_one_way_participant.xlsx"),
  overwrite = TRUE
)
message("Exported: Table_one_way_participant.xlsx")
```


---

# Descriptive plots {#descriptive-plots}

Plots are generated by custom functions in `PRG/Functions/functions_descriptive_plot.R` and saved automatically to `RES/Physiological/Graph/`. See that file for full documentation of each plotting function.

## Boxplots by heatwave period

The box represents the interquartile range (IQR) with median shown as a
horizontal line. Jittered grey points show individual repeated measurements. Red points indicate statistical outliers beyond 1.5x IQR within each group.

```{r Plot-boxplot-num-hw, echo=FALSE, fig.height=25, fig.width=20, message=FALSE, warning=FALSE}
# Boxplots for two-factor outcomes (BP and temperature) grouped by heatwave period
boxplot_numeric(
  data_list_num_two_way,
  outcome_vars_num_two_way,
  plot_type  = "heatwave",
  output_dir = here("RES/Physiological/Graph")
)
```

## Time series plots

The following time series plots show mean values per day with 95% confidence intervals. Vertical dashed lines mark heatwave start and end. Thin grey lines connect daily values for the same participant. Bold colored lines show the group mean per heatwave period (BEF/HW/AFT).

> Note: The individual `time_series_heatwave_period()` calls below are set to `eval=FALSE` -- they produce intermediate exploratory plots. The combined publication-ready plots are produced by `time_series_combine_plot_*()` below.

```{r eval=FALSE, fig.height=8, fig.width=6, message=FALSE, warning=FALSE, include=FALSE}
# Exploratory: separate plots by period of day only (not saved to RES)
time_series_heatwave_period(
  data_list_num_two_way, outcome_vars_num_two_way,
  group_by  = "period",
  common_x  = FALSE,
  output_dir = here("RES/Physiological/Graph")
)
```

```{r eval=FALSE, fig.height=8, fig.width=8, message=FALSE, warning=FALSE, include=FALSE}
# Exploratory: separate plots by heatwave status only (not saved to RES)
time_series_heatwave_period(
  data_list_num_two_way, outcome_vars_num_two_way,
  group_by   = "heatwave_status",
  common_x   = FALSE,
  output_dir = here("RES/Physiological/Graph")
)
```

### Two-factor combined time series (publication figure)

```{r Plot-time-series-2way-combine, fig.height=16, fig.width=19, message=FALSE, warning=FALSE}
# Subset of variables selected for the publication figure
# MAP excluded here as SBP, DBP and HR are the primary outcomes reported
data_list_num_two_way_final <- list(
  bp_data   = final_blood_pressure_data,
  temp_data = final_core_temp_data
)

outcome_vars_num_two_way_final <- list(
  "Systolic Blood Pressure (mmHg)"         = list(data = "bp_data",   var = "sbp",            date = "date_converted", day_time = "unix_time_converted"),
  "Diastolic Blood Pressure (mmHg)"        = list(data = "bp_data",   var = "dbp",            date = "date_converted", day_time = "unix_time_converted"),
  "Heart Rate (bpm)"                       = list(data = "bp_data",   var = "hr",             date = "date_converted", day_time = "unix_time_converted"),
  "Daily change\nOral Temperature (deg C)" = list(data = "temp_data", var = "delta_oral_temp", date = "date_converted", day_time = "unix_time_converted")
)

# Combined plot: all outcomes in one figure, faceted by period of day
time_series_combine_plot_two_way(
  data_list_num_two_way_final,
  outcome_vars_num_two_way_final,
  output_dir = here("RES/Physiological/Graph")
)
```

```{r Plot-time-series-2way-individual, fig.height=16, fig.width=19, message=FALSE, warning=FALSE}
# Same combined figure with individual participant trajectories overlaid
# Thin grey lines = individual participants, thick colored line = group mean
time_series_combine_plot_two_way_individual(
  data_list_num_two_way_final,
  outcome_vars_num_two_way_final,
  output_dir = here("RES/Physiological/Graph")
)
```

### One-factor combined time series (publication figure)

```{r Plot-time-series-1way-combine, message=FALSE, warning=FALSE}
# Subset of one-factor variables for the publication figure
# Body mass absolute value excluded -- only delta shown
data_list_num_one_way_final <- list(
  mass_data          = final_body_mass_data,
  total_movement_day = final_activity_data_day,
  urine_frequency    = final_urine_data
)

outcome_vars_num_one_way_final <- list(
  "Daily change\nBody Mass (g)"  = list(data = "mass_data",          var = "delta_bodymass", date = "date_converted"),
  "Mean Daily Acceleration (mg)" = list(data = "total_movement_day", var = "mean_acc_day",   date = "date_converted"),
  "Urine Frequency per Day"      = list(data = "urine_frequency",    var = "urineFreq",      date = "date_converted")
)

# Combined plot: all one-factor outcomes in one figure
time_series_combine_plot_one_way(
  data_list_num_one_way_final,
  outcome_vars_num_one_way_final,
  output_dir = here("RES/Physiological/Graph")
)
```

```{r Plot-time-series-1way-individual, message=FALSE, warning=FALSE}
# Same combined figure with individual participant trajectories overlaid
time_series_combine_plot_one_way_individual(
  data_list_num_one_way_final,
  outcome_vars_num_one_way_final,
  output_dir = here("RES/Physiological/Graph")
)
```

---

# Air conditioning subgroup analysis {#ac-analysis}

We investigate whether access to air conditioning (AC) moderates the effect of the heatwave on physiological outcomes.

## Assign AC status and update data lists

```{r define-ac-status}
# Participants with air conditioning access at home
# Derived from self-report at baseline (characteristics_participants.csv, ApartmentAC column)
# Hardcoded for reproducibility -- update if study population changes
participants_with_AC <- c(
  "P01", "P02", "P03", "P04", "P06", "P08", "P09", "P10",
  "P11", "P13", "P16", "P17", "P19", "P21", "P24", "P25", "P26"
)

# Add binary AC indicator (1 = AC, 0 = no AC) to all physiological datasets
# Two-factor outcomes (measured 3x daily)
final_blood_pressure_data$AC <- ifelse(
  final_blood_pressure_data$participant_id %in% participants_with_AC, 1, 0
)
final_core_temp_data$AC <- ifelse(
  final_core_temp_data$participant_id %in% participants_with_AC, 1, 0
)

# One-factor outcomes (measured once daily)
final_body_mass_data$AC <- ifelse(
  final_body_mass_data$participant_id %in% participants_with_AC, 1, 0
)
final_urine_data$AC <- ifelse(
  final_urine_data$participant_id %in% participants_with_AC, 1, 0
)
final_activity_data_day$AC <- ifelse(
  final_activity_data_day$participant_id %in% participants_with_AC, 1, 0
)

# Update all data lists with the AC-enriched dataframes
# This must be done before calling any AC plot or model functions
data_list_num_two_way$bp_data            <- final_blood_pressure_data
data_list_num_two_way$temp_data          <- final_core_temp_data
data_list_num_one_way$mass_data          <- final_body_mass_data
data_list_num_one_way$urine_frequency    <- final_urine_data
data_list_num_one_way$total_movement_day <- final_activity_data_day
```

## Time series by AC status -- one-factor outcomes

```{r plot-ac-one-way}
# Plot one-factor outcomes split by AC status (AC vs no AC)
# Uses outcome_vars_num_one_way_final defined in the plots section above
time_series_heatwave_AC(
  data_list_num_one_way,
  outcome_vars_num_one_way_final,
  group_by   = "heatwave",
  output_dir = here("RES/Physiological/Graph")
)
```

## Time series by AC status -- two-factor outcomes

```{r plot-ac-two-way}
# Redefine outcome_vars for two-factor AC plots
# outcome_vars_num_two_way_final updated here to reflect current variable names
outcome_vars_num_two_way_final <- list(
  "Systolic Blood Pressure (mmHg)"         = list(data = "bp_data",   var = "sbp",            date = "date_converted", day_time = "unix_time_converted"),
  "Diastolic Blood Pressure (mmHg)"        = list(data = "bp_data",   var = "dbp",            date = "date_converted", day_time = "unix_time_converted"),
  "Heart Rate (bpm)"                       = list(data = "bp_data",   var = "hr",             date = "date_converted", day_time = "unix_time_converted"),
  "Daily change\nOral Temperature (deg C)" = list(data = "temp_data", var = "delta_oral_temp", date = "date_converted", day_time = "unix_time_converted")
)

# Plot two-factor outcomes split by AC status
time_series_heatwave_AC(
  data_list_num_two_way,
  outcome_vars_num_two_way_final,
  group_by   = "heatwave",
  output_dir = here("RES/Physiological/Graph")
)
```

---

# Statistical analysis: Linear mixed models {#lmm}


## Two-factor MMRM: heatwave x period of day

```{r lmm-stats-2way, echo=FALSE, message=FALSE, warning=FALSE}
# Fits MMRM with heatwave_status * period as fixed effects and AR(1) covariance
# See PRG/Functions/functions_linear_models.R for full model specification
lmm_two_way(
  data_list_num_two_way,
  outcome_vars_num_two_way,
  posthoc = TRUE
)
```

## One-factor MMRM: heatwave period only

```{r lmm-stats-1way, echo=FALSE, message=FALSE, warning=FALSE}
# Fits MMRM with heatwave_status as the only fixed effect
# Applied to daily outcomes without period-of-day structure
lmm_one_way(
  data_list_num_one_way,
  outcome_vars_num_one_way,
  posthoc = TRUE
)
```

## Effect of air conditioning -- two-factor outcomes

```{r lmm-ac-two-way}
# MMRM including AC status as a moderator for two-factor outcomes
lmm_two_way_AC(
  data_list_num_two_way,
  outcome_vars_num_two_way,
  posthoc = TRUE
)
```

## Effect of air conditioning -- one-factor outcomes

```{r lmm-ac-one-way}
# MMRM including AC status as a moderator for one-factor outcomes
lmm_one_way_AC(
  data_list_num_one_way,
  outcome_vars_num_one_way,
  posthoc = TRUE
)
```

---

# Session information

```{r session-info}
sessionInfo()
```
