---
title: "Oral Temperature HeatSuite 2024"
author: "Souilla Luc"
date: "2026-03-04"
output: 
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
encoding: UTF-8
---
<!--
================================================================================
SCRIPT:  PRG/Physiological/oral_temperature.Rmd
PURPOSE: Full data preparation pipeline for oral temperature collected via oral probe device and smartwatch during the Montreal heatwave 2024.
         This script:
           1. Assigns participant IDs and study periods to raw oral temp data
           2. Applies calibration correction (+2 deg C) to raw values
           3. Flags and removes error and implausible measurements
           4. Computes two delta variables:
                - delta_oral_temp         : day-to-day lag change
                                            used for plots and mixed models
                - delta_oral_temp_baseline: deviation from individual
                                            pre-heatwave mean used for descriptive statistics
           5. Removes duplicate measurements using sleep-based reference times
           6. Exports the clean final dataset and QC plots

INPUT FILES (DAT/):
  - oral_temperature.csv            -> raw thermometer measurements
                                       Note: uses comma separator (sep=",")
                                       unlike blood_pressure.csv (sep=";")
  - baseline_long.csv               -> participant-macID-date lookup
                                       (from Data_preparation.Rmd)
  - sleep_analysis_daily.csv        -> daily sleep start/end times

OUTPUT FILES (RES/Physiological/):
  - oral_temp_processed.csv         -> cleaned dataset with both delta variables

CALIBRATION NOTE:
  A correction factor of +2 deg C is applied to all raw temperature values based on device calibration. Formula: temp_corrected = temp + 2.

IMPLAUSIBLE VALUE THRESHOLDS:
  Error flags      : temp_corrected < 30 or > 43 deg C
  Implausible flags: temp_corrected < 34.5 or > 41 deg C
================================================================================
-->

```{r setup, include=FALSE}
# Global knitr options: show code, suppress messages and warnings
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE
)
```


# Overview

This script prepares the **oral temperature** dataset collected during the Montreal heatwave 2024. It follows the three-step pipeline
defined in `Data_preparation.Rmd`:

1. **`oral_temp_raw`** -- raw upload, calibration correction, participant ID assignment
2. **`oral_temp_cleaned`** -- error flagging, implausible value removal, delta computation
3. **`final_oral_temp_data`** -- deduplication via reference-time selection, export


***

# Environment setup

## Load packages

```{r load-packages, message=FALSE, warning=FALSE, include=FALSE}
required_packages <- c(
  "here", "dplyr", "tidyr", "lubridate",
  "ggplot2", "patchwork", "purrr"
)

install_if_missing <- function(packages) {
  missing <- packages[!(packages %in% installed.packages()[, "Package"])]
  if (length(missing) > 0) install.packages(missing)
}

install_if_missing(required_packages)
invisible(lapply(required_packages, library, character.only = TRUE))
```

## Set project root and directory paths

```{r Set Up environment, message=FALSE, warning=FALSE, include=FALSE}
# Anchor here() to the project root from this script's location
here::i_am("PRG/Physiological/oral_temperature.Rmd")  

# Define paths using 'here' to get project-relative paths
path_dat <- here("DAT")          # Path to the DAT folder within the project
path_prg <- here("PRG")          # Path to the PRG folder within the project
path_res <- here("DAT/Processed")          # Path to Processed folders in DAT
```

## Source custom functions

```{r source-functions}
# Participant ID assignment, period labelling, heatwave status
source(here("PRG/Functions/functions_participant_period.R"))

# Reference-time selection and deduplication (requires purrr)
source(here("PRG/Functions/functions_cleaning_data.R"))

# Night-period reference time adjustment
source(here("PRG/Functions/functions_adjusted_ref_time.R"))
```

***

# Step 1 - Load raw data and assign participant IDs {#step1}

## Load raw oral temperature data

```{r load-ot-raw}

oral_temp_raw <- read.csv(
  file.path(path_dat, "oral_temperature.csv"),
  header = TRUE, sep = ",", quote = "\"'", dec = "."
)
```

## Load participant-macID lookup table

```{r load-baseline-long}
# baseline_long maps each macID + date range to a participant_id
# Created by Data_preparation.Rmd must be run first
baseline_long <- read.csv(
  file.path(path_dat, "baseline_long.csv"),
  header = TRUE, sep = ",", quote = "\"'", dec = "."
)
```

## Assign study period, participant ID, and heatwave status

```{r assign-participant-period}
# Step 1a: Convert unix timestamp and assign period of day (AM / PM / NI)
oral_temp_participant <- assign_study_period(oral_temp_raw, c("unix"))

# Step 1b: Create local_time from unix for participant ID matching
# Applies EDT offset (-14400 s = UTC-4) for Montreal local time
oral_temp_participant <- oral_temp_participant %>%
  mutate(
    local_time = as.POSIXct(unix, origin = "1970-01-01", tz = "UTC") - hours(4)
  )

# Step 1c: Match each row to a participant_id using macID + local_time window
oral_temp_participant <- add_participant_id_watch(
  oral_temp_participant,
  baseline_long,
  mac_col  = "macID",
  time_col = "local_time"
)

# Step 1d: Assign heatwave status (BEF_HW_June / HW_June / AFT_HW_June)
oral_temp_participant <- assign_heatwave(oral_temp_participant)

# Restrict to heatwave monitoring window only
oral_temp_participant <- oral_temp_participant %>%
  filter(heatwave_status %in% c("BEF_HW_June", "HW_June", "AFT_HW_June"))
```

***

# Step 2 - Calibration correction {#step2}

All raw temperature values require a calibration adjustment of +2 deg C,
determined during the device calibration process prior to the study.
Formula: `temp_corrected = temp + 2`

```{r apply-calibration}
# Apply calibration correction to all temperature measurements
oral_temp_participant <- oral_temp_participant %>%
  mutate(temp_corrected = temp + 2)
```

***

# Step 3 - Flag implausible and error measurements {#step3}

Two levels of flagging are applied based on clinical thresholds.

| Flag | Condition |
|---|---|
| `error_measurement` | temp_corrected < 30 or > 43 deg C |
| `implausible_measurement` | temp_corrected < 34.5 or > 41 deg C |


```{r flag-measurements}
oral_temp_participant <- oral_temp_participant %>%
  mutate(
    # Hard error: physiologically impossible values
    error_measurement       = ifelse(temp_corrected < 30 | temp_corrected > 43, 1, 0),
    # Implausible: outside normal human oral temperature range
    implausible_measurement = ifelse(temp_corrected < 34.5 | temp_corrected > 41, 1, 0)
  )
```

## Visualize flagged measurements before removal

```{r plot-before-filtering, echo=FALSE, fig.height=8, fig.width=15}

oral_temp_data_subset <- oral_temp_participant %>%
  mutate(
    measurement_type = case_when(
      error_measurement       == 1 ~ "Error",
      implausible_measurement == 1 ~ "Implausible",
      TRUE                         ~ "Valid"
    ),
    measurement_type = factor(measurement_type,
                               levels = c("Valid", "Implausible", "Error")),
    hour_of_day = hour(unix_time_converted)
  )

flag_colors <- c("Valid" = "#2E86AB", "Implausible" = "#F6AE2D", "Error" = "#E63946")
flag_shapes <- c("Valid" = 16,        "Implausible" = 17,        "Error" = 4)

line_oral_temp_complete_bef <- ggplot(
  oral_temp_data_subset,
  aes(x = date_converted, y = temp_corrected)
) +
  geom_jitter(aes(color = measurement_type, shape = measurement_type),
              size = 0.8, alpha = 0.85, width = 0.2, height = 0) +
  scale_color_manual(values = flag_colors, name = "Type") +
  scale_shape_manual(values = flag_shapes, name = "Type") +
  labs(title = "Before filtering",
       x = "Date of measurement", y = "Oral temperature (deg C)") +
  theme_minimal() +
  theme(plot.title     = element_text(hjust = 0.5),
        legend.position = "right",
        axis.text.x    = element_text(angle = 45, hjust = 1, size = 10))
print(line_oral_temp_complete_bef)
```


## Remove error and implausible measurements

```{r remove-flagged}
oral_temp_cleaned <- oral_temp_participant %>%
  filter(error_measurement == 0, implausible_measurement == 0)

n_removed   <- nrow(oral_temp_participant) - nrow(oral_temp_cleaned)
pct_removed <- round((n_removed / nrow(oral_temp_participant)) * 100, 2)
message("Rows removed (error + implausible): ", n_removed,
        " (", pct_removed, "% of total)")
```


***

# Step 4 - Compute delta variables {#step4}

Two delta variables are computed and retained for different analytical purposes.

## delta_oral_temp: day-to-day lag (for plots and mixed models)


```{r compute-delta-lag}
# Compute day-to-day change in oral temperature
# Used for: time series visualizations and linear mixed models (MMRM)
# Rationale: captures longitudinal fluctuation across the study period
oral_temp_cleaned <- oral_temp_cleaned %>%
  arrange(participant_id, date_converted, period) %>%
  group_by(participant_id) %>%
  mutate(
    delta_oral_temp = temp_corrected - lag(temp_corrected)
  ) %>%
  ungroup() %>%
  # First measurement per participant has no prior value -- set to 0
  mutate(delta_oral_temp = ifelse(is.na(delta_oral_temp), 0, delta_oral_temp))
```

## delta_oral_temp_baseline: deviation from individual pre-heatwave mean (for descriptive tables)
```{r compute-delta-baseline}
# Step 1: Compute each participant's mean temperature during pre-heatwave period
# This serves as the individual thermal baseline reference
baseline_temp <- oral_temp_cleaned %>%
  filter(heatwave_status == "BEF_HW_June") %>%
  group_by(participant_id) %>%
  summarise(
    baseline_mean = mean(temp_corrected, na.rm = TRUE),
    .groups = "drop"
  )

# Step 2: Join baseline mean and compute deviation for every measurement
# Used for: descriptive statistics tables (mean +/- SD per heatwave period)
oral_temp_cleaned <- oral_temp_cleaned %>%
  left_join(baseline_temp, by = "participant_id") %>%
  mutate(delta_oral_temp_baseline = temp_corrected - baseline_mean)

# Verify expected pattern: pre-HW mean delta should be approximately 0
oral_temp_cleaned %>%
  filter(heatwave_status %in% c("BEF_HW_June", "HW_June", "AFT_HW_June")) %>%
  group_by(heatwave_status) %>%
  summarise(
    mean_delta_baseline = round(mean(delta_oral_temp_baseline, na.rm = TRUE), 2),
    sd_delta_baseline   = round(sd(delta_oral_temp_baseline,   na.rm = TRUE), 2),
    mean_delta_lag      = round(mean(delta_oral_temp,          na.rm = TRUE), 2),
    sd_delta_lag        = round(sd(delta_oral_temp,            na.rm = TRUE), 2),
    n                   = n()
  )
```

***

# Step 5 - Check and remove duplicate measurements {#step5}

## Initial duplicate check

```{r check-duplicates-initial, message=FALSE, warning=FALSE}
duplicates_check <- oral_temp_cleaned %>%
  group_by(period, macID, date_converted, participant_id) %>%
  summarise(duplicate_count = n(), .groups = "drop") %>%
  filter(duplicate_count > 1)

if (nrow(duplicates_check) > 0) {
  message("Duplicates found -- will resolve using reference time:")
  print(duplicates_check)
} else {
  message("No duplicates found.")
}
```

## Load and prepare sleep data for reference-time matching

```{r load-sleep}
sleep_analysis_raw <- read.csv(
  file.path(path_dat, "sleep_analysis_daily.csv"),
  header = TRUE, sep = ",", quote = "\"'", dec = "."
)
```

```{r assign-participant-sleep}
# Derive local_time from sleepEnd for participant ID matching
# EDT offset: -14400 seconds (UTC-4, valid for June 2024 Montreal)
sleep_analysis_raw <- sleep_analysis_raw %>%
  mutate(
    local_time = as.Date(
      as.POSIXct(sleepEnd, origin = "1970-01-01", tz = "UTC") - 14400
    )
  )

sleep_analysis_participant <- add_participant_id_watch(
  sleep_analysis_raw,
  baseline_long,
  mac_col  = "macID",
  time_col = "local_time"
)
```

```{r filter-sleep, message=FALSE, warning=FALSE}
# Convert Unix timestamps to local Montreal time and extract date components
sleep_analysis_participant <- sleep_analysis_participant %>%
  mutate(
    sleepStart_time = as.POSIXct(sleepStart, origin = "1970-01-01", tz = "UTC") - 14400,
    sleepEnd_time   = as.POSIXct(sleepEnd,   origin = "1970-01-01", tz = "UTC") - 14400,
    date_bedtime    = as.Date(sleepStart_time),
    date_awakening  = as.Date(sleepEnd_time)
  ) %>%
  # Restrict to heatwave monitoring window (June 9-29, 2024)
  filter(
    (sleepStart_time >= as.POSIXct("2024-06-09 00:00:00") &
     sleepStart_time <= as.POSIXct("2024-06-29 23:59:59")) |
    (sleepEnd_time   >= as.POSIXct("2024-06-09 00:00:00") &
     sleepEnd_time   <= as.POSIXct("2024-06-29 23:59:59"))
  )

# Retain only physiologically plausible sleep episodes
sleep_analysis_filter <- sleep_analysis_participant %>%
  select(participant_id, macID, sleepStart, sleepEnd,
         sleepStart_time, sleepEnd_time, date_bedtime, date_awakening) %>%
  filter(
    format(sleepEnd_time,    "%H:%M") >= "03:00" &
    format(sleepEnd_time,    "%H:%M") <= "11:59" &
    (format(sleepStart_time, "%H:%M") >= "18:00" |
     format(sleepStart_time, "%H:%M") <= "02:59")
  )

n_sleep_removed <- nrow(sleep_analysis_participant) - nrow(sleep_analysis_filter)
message("Sleep records retained: ", nrow(sleep_analysis_filter),
        " | Removed (implausible windows): ", n_sleep_removed)
```

## Merge oral temperature and sleep data

```{r merge-ot-sleep}
# Align date keys: use date_awakening as merge date
sleep_analysis_filter <- sleep_analysis_filter %>%
  mutate(date_converted = date_awakening)

# Left join: every oral temperature record is kept
oral_temp_sleep_merged <- oral_temp_cleaned %>%
  left_join(
    sleep_analysis_filter,
    by = c("macID", "participant_id", "date_converted")
  )
```

## Assign reference times

```{r set-reference-time}
# Assigns reference_time per period:
#   AM -> awakening time (sleepEnd_time)
#   PM -> 15:00 on date_converted
#   NI -> bedtime (sleepStart_time)
oral_temp_sleep_merged <- set_reference_time(oral_temp_sleep_merged)
```

```{r adjust-night-reference-time}
# Adjusts reference_time for NI measurements recorded after midnight
oral_temp_sleep_merged <- set_adjusted_night_time(oral_temp_sleep_merged)
```

## Verify night reference time adjustment

```{r verify-night-reference-time}
# Compute time difference between NI measurements and adjusted bedtime
oral_temp_night_verif <- oral_temp_sleep_merged %>%
  filter(period == "NI") %>%
  mutate(
    time_diff_minutes = as.numeric(
      difftime(unix_time_converted, adjusted_reference_time, units = "mins")
    )
  )

time_diff_summary <- oral_temp_night_verif %>%
  summarise(
    mean_diff = mean(time_diff_minutes, na.rm = TRUE),
    sd_diff   = sd(time_diff_minutes,   na.rm = TRUE),
    min_diff  = min(time_diff_minutes,  na.rm = TRUE),
    max_diff  = max(time_diff_minutes,  na.rm = TRUE)
  )

message(
  "NI period -- time between measurement and adjusted bedtime (minutes):",
  "\n  Mean = ", round(time_diff_summary$mean_diff, 2),
  " | SD = ",   round(time_diff_summary$sd_diff,   2),
  " | Min = ",  round(time_diff_summary$min_diff,  2),
  " | Max = ",  round(time_diff_summary$max_diff,  2)
)
```

## Select one measurement per participant per period


```{r select-reference-measurement}
# Number each measurement per group in descending unix order
# row_num = 1 -> most recent; row_num > 1 -> older measurements to evaluate
oral_temp_sleep_merged <- oral_temp_sleep_merged %>%
  arrange(macID, period, date_converted, unix) %>%
  group_by(macID, period, date_converted, participant_id) %>%
  mutate(row_num = row_number(desc(unix))) %>%
  ungroup()

# Compute absolute time difference from reference_time for each measurement
oral_temp_sleep_merged <- oral_temp_sleep_merged %>%
  group_by(participant_id, date_converted, period) %>%
  mutate(
    abs_time_diff = abs(difftime(unix_time_converted, reference_time, units = "secs"))
  ) %>%
  ungroup()

# Identify rows in groups with more than one measurement
indices_row <- which(oral_temp_sleep_merged$row_num > 1)
message("Rows in groups with multiple measurements: ", length(indices_row))

# Retrieve surrounding indices for comparison within each group
surrounding_indices_list <- get_surrounding_indices(indices_row, oral_temp_sleep_merged)

# Keep only the measurement closest to the reference time per group
final_oral_temp_data <- remove_farthest_measure(
  oral_temp_sleep_merged,
  surrounding_indices_list
)

n_removed_duplicates <- nrow(oral_temp_sleep_merged) - nrow(final_oral_temp_data)
message("Rows removed (farthest from reference time): ", n_removed_duplicates)
```


## Final duplicate verification
```{r check-duplicates-final, message=FALSE, warning=FALSE}
duplicates_final <- final_oral_temp_data %>%
  group_by(period, macID, date_converted) %>%
  summarise(duplicate_count = n(), .groups = "drop") %>%
  filter(duplicate_count > 1)

if (nrow(duplicates_final) > 0) {
  message("WARNING: Duplicates still present:")
  print(duplicates_final)
} else {
  message("Final duplicate check passed: one measurement per participant per period per date.")
}
```

***

# Step 6 - Final dataset and export {#step6}

## Remove intermediate processing columns

```{r clean-final-columns}


final_oral_temp_data <- final_oral_temp_data %>%
  select(
    -id, -tz, -adc, -ta, -rh, -measures, -local_date,
    -error_measurement, -implausible_measurement,
    -matches("^[Ss]leep"),
    -date_bedtime, -date_awakening,
    -hours_discrepancy, -adjusted_reference_time,
    -abs_time_diff, -reference_time,
    -local_time, -row_num,
    -baseline_mean, -time_diff_minutes, -unix, -temp                # Intermediate variable used to compute
                                  # delta_oral_temp_baseline -- not needed in output
  )
```

## Export processed dataset

```{r export-final}
output_file <- file.path(path_res, "oral_temp_processed.csv")
write.csv(final_oral_temp_data, output_file, row.names = FALSE)
message("Exported: oral_temp_processed.csv (",
        nrow(final_oral_temp_data), " rows x ",
        ncol(final_oral_temp_data), " columns)")
```

***

# Step 7 - Quality control plots {#step7}

## Before vs. after filtering: temperature over time

```{r plot-timeseries-after, echo=FALSE}
line_oral_temp_complete_aft <- ggplot(
  final_oral_temp_data,
  aes(x = date_converted, y = temp_corrected)
) +
  geom_jitter(color = "#2E86AB",   # Fixed color outside aes() to avoid legend artefact
              size = 0.8, alpha = 0.85, width = 0.2, height = 0) +
  labs(title = "After filtering",
       x = "Date of measurement", y = "Oral temperature (deg C)") +
  theme_minimal() +
  theme(plot.title      = element_text(hjust = 0.5),
        legend.position = "right",
        axis.text.x     = element_text(angle = 45, hjust = 1, size = 10))

line_oral_temp_complete_bef / line_oral_temp_complete_aft +
  plot_annotation(title = "Effect of data cleaning on oral temperature")
```
## Distribution of delta oral temperature after filtering

```{r plot-histogram-after, echo=FALSE}
histo_oral_tp_aft <- ggplot(final_oral_temp_data, aes(x = delta_oral_temp)) +
  geom_histogram(aes(y = after_stat(density)),
                 fill = "lightblue", color = "grey", bins = 30) +
  labs(title = "After filtering",
       x = "Daily change in oral temperature (deg C)", y = "Density") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))
print(histo_oral_tp_aft)
```


## Boxplots by period of day: before vs. after filtering

```{r plot-boxplots-before-after, echo=FALSE}
bxp_oral_tp_bef <- ggplot(
  oral_temp_cleaned,
  aes(x = period, y = delta_oral_temp, fill = period)
) +
  geom_boxplot(outlier.shape = NA, alpha = 0.8, width = 0.3) +
  geom_jitter(width = 0.2, alpha = 0.4, size = 1, color = "grey2") +
  labs(title = "Before filtering", x = "", y = "Daily change in oral temperature (deg C)") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5), legend.position = "bottom")

bxp_oral_tp_aft <- ggplot(
  final_oral_temp_data,
  aes(x = period, y = delta_oral_temp, fill = period)
) +
  geom_boxplot(outlier.shape = NA, alpha = 0.8, width = 0.3) +
  geom_jitter(width = 0.2, alpha = 0.4, size = 1, color = "grey2") +
  labs(title = "After filtering", x = "", y = "Daily change in oral temperature (deg C)") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5), legend.position = "bottom")

bxp_oral_tp_bef / bxp_oral_tp_aft +
  plot_annotation(title = "Effect of data cleaning on delta oral temperature by period")
```


***

# Session information

```{r session-info}
sessionInfo()
```

