---
title: "Physical Activity HeatSuite 2024"
author: "Souilla Luc"
date: "2026-03-04"
output: 
 html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
encoding: UTF-8
---

<!--
================================================================================
SCRIPT:  PRG/Physiological/urine_frequency.Rmd
PURPOSE: Data preparation pipeline for urine frequency and urine color
         collected via survey on smartwatch during the Montreal Summer
         2024 heatwave study. This script:
           1. Loads daily lifestyle survey data and assigns participant IDs
           2. Converts numeric urine color codes to descriptive labels
           3. Removes invalid entries (urineFreq = 0)
           4. Checks for and reports duplicate entries per participant per day
           5. Exports the clean final dataset and QC plots

INPUT FILES (DAT/):
  - lifestyle_daily.csv             -> daily self-reported lifestyle survey
                                       including urine frequency and color
  - baseline_long.csv               -> participant-macID-date lookup
                                       (from Data_preparation.Rmd)

OUTPUT FILES (RES/Physiological/):
  - urine_data_processed.csv        -> cleaned daily urine dataset

DESIGN NOTE - WHY THIS SCRIPT IS SIMPLER THAN OTHER PHYSIOLOGICAL SCRIPTS:
  Urine data comes from a daily survey (one entry per day per participant), not from continuous smartwatch monitoring. As a result:
    - No unix timestamp exists: local_time is constructed from the date
      column using 15:00 as a fixed arbitrary time (does not affect results)
    - assign_study_period() is NOT called: period of day (AM/PM/NI) is not applicable to daily survey data. Period is set to "AM" by convention
      for consistency with other dataframes in downstream analyses
    - No sleep merge or reference-time deduplication is needed: only one
      entry per participant per day is expected by design

URINE COLOR CODING:
  0 = white
  1 = yellow1 (pale yellow)
  2 = yellow2 (yellow)
  3 = yellow3 (dark yellow)
  4 = yellow4 (amber/brown)
  NA or other = missing / not recorded

================================================================================
-->

```{r setup, include=FALSE}
# Global knitr options: show code, suppress messages and warnings
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE
)
```

# Overview

This script prepares the **urine frequency and urine color** dataset from the daily survey completed by participants during the Montreal heatwave 2024.

Unlike other physiological scripts (blood pressure, body mass, oral temperature), this script does not perform sleep-based reference-time deduplication because the data source is a daily survey with at most one intended entry per participant per day. See the header block above for a full explanation of design differences.

***
# Environment setup

## Load packages

```{r load-packages, message=FALSE, warning=FALSE, include=FALSE}
required_packages <- c(
  "here", "dplyr", "tidyr", "lubridate",
  "ggplot2", "patchwork", "scales"
)

install_if_missing <- function(packages) {
  missing <- packages[!(packages %in% installed.packages()[, "Package"])]
  if (length(missing) > 0) install.packages(missing)
}

install_if_missing(required_packages)
invisible(lapply(required_packages, library, character.only = TRUE))
```


## Set project root and directory paths

```{r set-paths}
# Anchor here() to the project root from this script's location
here::i_am("PRG/Physiological/urine_frequency.Rmd") 

path_dat <- here("DAT")         # Raw data folder
path_prg <- here("PRG")         # Scripts folder
path_res <- here("DAT/Processed") # Path to DAT processed folder within the project
```

## Source custom functions

```{r source-functions}
# Participant ID assignment and heatwave status labelling
# Note: assign_study_period() is intentionally NOT used here -- see design note
source(here("PRG/Functions/functions_participant_period.R"))
```

***

# Step 1 - Load raw data and assign participant IDs {#step1}

## Load raw lifestyle survey data

```{r load-urine-raw}
# Load daily lifestyle survey -- contains urine frequency, urine color,
# and other daily lifestyle variables (steps, sleep, heart rate, etc.)
urine_data_raw <- read.csv(
  file.path(path_dat, "lifestyle_daily.csv"),
  header = TRUE, sep = ",", quote = "\"'", dec = "."
)
```

## Load participant-macID lookup table

```{r load-baseline-long}
# baseline_long maps each macID + date range to a participant_id
# Created by Data_preparation.Rmd -- must be run first
baseline_long <- read.csv(
  file.path(path_dat, "baseline_long.csv"),
  header = TRUE, sep = ",", quote = "\"'", dec = "."
)
```

## Assign participant ID and heatwave status

```{r assign-participant-period}
# This dataset contains daily survey entries with a date column but no unix
# timestamp. We construct local_time at 15:00 as a fixed arbitrary reference
# point solely to satisfy the add_participant_id_watch() function signature.
# The specific time chosen (15:00) does not affect participant ID matching
# since matching is done by date range, not by exact time.
urine_data_participant <- urine_data_raw %>%
  mutate(
    local_time     = as.POSIXct(paste0(date, " 15:00:00"),
                                format = "%Y-%m-%d %H:%M:%S", tz = "UTC"),
    date_converted = as.Date(date, format = "%Y-%m-%d")
  )

# Match each row to a participant_id using macID + local_time window
urine_data_participant <- add_participant_id_watch(
  urine_data_participant,
  baseline_long,
  mac_col  = "macID",
  time_col = "local_time"
)

# Assign heatwave status (BEF_HW_June / HW_June / AFT_HW_June)
urine_data_participant <- assign_heatwave(urine_data_participant)

# Restrict to heatwave monitoring window only
urine_data_participant <- urine_data_participant %>%
  filter(heatwave_status %in% c("BEF_HW_June", "HW_June", "AFT_HW_June"))

# Set period = "AM" by convention for all daily survey entries
# Rationale: urine data is collected once per day with no time-of-day
# granularity. Setting period = "AM" ensures consistency with the period
# column used in all other physiological dataframes for downstream merging.
urine_data_participant <- urine_data_participant %>%
  mutate(period = "AM")
```
***

# Step 2 - Clean urine frequency and color {#step2}

## Convert urine color codes to descriptive labels

Numeric urine color codes (0-4) are converted to descriptive character labels for readability in plots and exported data.
```{r convert-urine-color}
urine_data_participant <- urine_data_participant %>%
  mutate(
    urineColorResp = case_when(
      urineColor == 0 ~ "white",    # Very pale / clear
      urineColor == 1 ~ "yellow1",  # Pale yellow
      urineColor == 2 ~ "yellow2",  # Yellow
      urineColor == 3 ~ "yellow3",  # Dark yellow
      urineColor == 4 ~ "yellow4",  # Amber / brown
      TRUE            ~ NA_character_ # Missing or not recorded
    )
  )
```

## Remove invalid urine frequency entries

A urineFreq value of 0 is physiologically impossible and indicates the
participant did not complete the survey for that day. These rows are removed.

```{r remove-invalid-frequency}
final_urine_data <- urine_data_participant %>%
  filter(urineFreq != 0)

n_removed <- nrow(urine_data_participant) - nrow(final_urine_data)
message("Rows removed (urineFreq = 0, survey not completed): ", n_removed)
```

***

# Step 3 - Check for duplicate entries {#step3}

Each participant should have at most one survey entry per day. Duplicates may occur if a participant submitted the survey twice on the same day.

```{r check-duplicates, message=FALSE, warning=FALSE}
# Check for multiple entries per participant per date
# Note: date_converted already created in Step 1 -- no need to recreate it
duplicates_check <- final_urine_data %>%
  group_by(period, macID, date_converted, participant_id) %>%
  summarise(duplicate_count = n(), .groups = "drop") %>%
  filter(duplicate_count > 1)

if (nrow(duplicates_check) > 0) {
  message("Duplicates found for the following participants and dates:")
  print(duplicates_check)
} else {
  message("No duplicates found: one entry per participant per day confirmed.")
}
```


***

# Step 4 - Final dataset and export {#step4}

## Remove columns not needed for analysis

```{r clean-final-columns}
# Remove lifestyle variables not relevant to urine analysis
# and intermediate processing columns
final_urine_data <- final_urine_data %>%
  select(
    -id, -date,
    -total_min, -charge_time, -steps,
    -sedentary_time,
    -hr_rest_avg, -hr_rest_min, -hr_rest_max, -hr_rest_count,
    -hr_active_avg, -hr_active_min, -hr_active_max, -hr_active_count,
    -error, -local_date
  )
```

## Export processed dataset

```{r export-final}
output_file <- file.path(path_res, "urine_data_processed.csv")
write.csv(final_urine_data, output_file, row.names = FALSE)
message("Exported: urine_data_processed.csv (",
        nrow(final_urine_data), " rows x ",
        ncol(final_urine_data), " columns)")
```


***

# Step 5 - Quality control plots {#step5}

## Urine frequency: before vs. after filtering

```{r plot-frequency-before-after, echo=FALSE}
line_freq_bef <- ggplot(urine_data_participant,
                        aes(x = date_converted, y = urineFreq)) +
  geom_jitter(alpha = 0.8, size = 0.8, width = 0.2) +
  labs(title = "Before filtering",
       x = "", y = "Urine frequency (times/day)") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5), legend.position = "bottom")

line_freq_aft <- ggplot(final_urine_data,
                        aes(x = date_converted, y = urineFreq)) +
  geom_jitter(alpha = 0.8, size = 0.8, width = 0.2) +
  labs(title = "After filtering",
       x = "", y = "Urine frequency (times/day)") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5), legend.position = "bottom")

line_freq_bef / line_freq_aft +
  plot_annotation(title = "Effect of data cleaning on urine frequency")
```

## Urine frequency: distribution before vs. after filtering

```{r plot-frequency-histograms, echo=FALSE}
histo_freq_bef <- ggplot(urine_data_participant, aes(x = urineFreq)) +
  geom_histogram(aes(y = after_stat(density)),
                 fill = "lightblue", color = "grey", bins = 10) +
  labs(title = "Before filtering",
       x = "Urine frequency (times/day)", y = "Density") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))

histo_freq_aft <- ggplot(final_urine_data, aes(x = urineFreq)) +
  geom_histogram(aes(y = after_stat(density)),
                 fill = "lightblue", color = "grey", bins = 10) +
  labs(title = "After filtering",
       x = "Urine frequency (times/day)", y = "Density") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))

histo_freq_bef / histo_freq_aft +
  plot_annotation(title = "Effect of data cleaning on urine frequency distribution")
```

## Urine color: before vs. after filtering

```{r plot-color-before-after, echo=FALSE}
# Color palette matching actual urine color appearance
urine_colors <- c(
  "white"   = "#FFFFF0",
  "yellow1" = "#FFF176",
  "yellow2" = "#FFD600",
  "yellow3" = "#FFA000",
  "yellow4" = "#E65100"
)

line_color_bef <- ggplot(urine_data_participant,
                         aes(x = date_converted, y = urineColorResp,
                             color = urineColorResp)) +
  geom_jitter(alpha = 0.8, size = 0.8, width = 0.2, height = 0.2) +
  scale_color_manual(values = urine_colors, name = "Urine color") +
  scale_y_discrete(limits = c("white", "yellow1", "yellow2",
                               "yellow3", "yellow4")) +
  labs(title = "Before filtering", x = "", y = "Urine color") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5), legend.position = "bottom")

line_color_aft <- ggplot(final_urine_data,
                         aes(x = date_converted, y = urineColorResp,
                             color = urineColorResp)) +
  geom_jitter(alpha = 0.8, size = 0.8, width = 0.2, height = 0.2) +
  scale_color_manual(values = urine_colors, name = "Urine color") +
  scale_y_discrete(limits = c("white", "yellow1", "yellow2",
                               "yellow3", "yellow4")) +
  labs(title = "After filtering", x = "", y = "Urine color") +  # FIX: was "Before filtering"
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5), legend.position = "bottom")

line_color_bef / line_color_aft +
  plot_annotation(title = "Effect of data cleaning on urine color")
```

## Urine color: proportion distribution before vs. after filtering

```{r plot-color-proportions, echo=FALSE}
histo_color_bef <- ggplot(
  urine_data_participant %>% filter(!is.na(urineColorResp)),
  aes(x = urineColorResp, fill = urineColorResp)
) +
  geom_bar(aes(y = after_stat(count / sum(count))), color = "grey") +
  scale_fill_manual(values = urine_colors, name = "Urine color") +
  scale_x_discrete(limits = c("white", "yellow1", "yellow2",
                               "yellow3", "yellow4")) +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "Before filtering", x = "Urine color", y = "Proportion") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5), legend.position = "bottom")

histo_color_aft <- ggplot(
  final_urine_data %>% filter(!is.na(urineColorResp)),
  aes(x = urineColorResp, fill = urineColorResp)
) +
  geom_bar(aes(y = after_stat(count / sum(count))), color = "grey") +
  scale_fill_manual(values = urine_colors, name = "Urine color") +
  scale_x_discrete(limits = c("white", "yellow1", "yellow2",
                               "yellow3", "yellow4")) +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "After filtering", x = "Urine color", y = "Proportion") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5), legend.position = "bottom")

histo_color_bef / histo_color_aft +
  plot_annotation(title = "Effect of data cleaning on urine color distribution")
```

---

# Session information

```{r session-info}
sessionInfo()
```