---
title: "Data Preparation — HeatSuite 2024"
author: "Souilla Luc"
date: "2026-03-04"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
encoding: UTF-8
---

<!--
================================================================================
SCRIPT: Data_preparation.Rmd
PURPOSE: Entry point for all data preparation steps in the HeatSuite 2024 study.
         This script:
           1. Loads and installs all required R packages
           2. Defines project-relative directory paths
           3. Sets up the global color palette
           4. Sources all custom analysis functions
           5. Labels each dataset with a Participant ID (via macID matching)
           6. Exports baseline_long.csv and baseline_nodes.csv for reuse
              across all downstream analysis scripts

INPUT FILES (DAT/):
  - characteristics_participants.csv  → raw participant + smartwatch metadata

OUTPUT FILES (DAT/):
  - baseline_long.csv   → one row per participant per smartwatch (bangle)
  - baseline_nodes.csv  → one row per participant per environmental node

DEPENDENCIES:
  - PRG/Functions/functions_participant_period.R
  - PRG/Functions/functions_cleaning_data.R
  - PRG/Functions/functions_adjusted_ref_time.R
  - PRG/Functions/functions_descriptive_stats.R
  - PRG/Functions/functions_descriptive_plot.R
  - PRG/Functions/functions_linear_models.R
  - PRG/Functions/residual_check_mmrm.R
  - PRG/Functions/functions_check_clmm.R

AUTHOR: Souilla Luc
LAST UPDATED: 2026-03-04
REPOSITORY: https://github.com/Souilla-Luc/Montreal_Summer_2024
================================================================================
-->




```{r setup, include=FALSE}
# Global knitr options: show code in output, suppress messages for cleaner HTML
knitr::opts_chunk$set(
  echo    = TRUE,    # Display code chunks in rendered output
  message = FALSE,   # Suppress package loading messages
  warning = FALSE    # Suppress warnings globally
)
```

# Overview

This script is the **entry point** for all data preparation in the HeatSuite 2024 study. Its objectives are:

- Assign a unique **Study Participant ID** to each data table using smartwatch macIDs and enrollment dates, to facilitate merging across all future analyses.
- Export two baseline lookup tables (`baseline_long.csv`, `baseline_nodes.csv`) that will be reused in every downstream analysis script.

## Project directory structure

The project is organized as follows:

- **`DAT/`** — raw and processed data tables (`.csv`)
- **`PRG/`** — all R scripts and R Markdown analysis files
- **`RES/`** — all output results (figures, model summaries, processed tables)

# Environment setup

## Load required packages

We first define all required packages, then check which ones are missing and install them automatically before loading.


```{r load-packages, message=FALSE, warning=FALSE, include=FALSE}
# Full list of packages required across the entire HeatSuite 2024 analysis
# Note: all packages are listed once — duplicates removed (lubridate, stringr)
required_packages <- c(
  # Data wrangling
  "here", "dplyr", "tidyr", "purrr", "stringr", "lubridate",
  # Reporting and tables
  "knitr", "kableExtra",
  # Visualization
  "ggplot2", "patchwork", "viridis",
  # Statistical modelling
  "carData", "rstatix", "ggeffects", "mlogit",
  "emmeans", "mmrm", "car", "ordinal", "glmmTMB"
)

# Helper: install any packages not yet present in the library
install_if_missing <- function(packages) {
  missing_packages <- packages[!(packages %in% installed.packages()[, "Package"])]
  if (length(missing_packages) > 0) {
    message("Installing missing packages: ", paste(missing_packages, collapse = ", "))
    install.packages(missing_packages)
  }
}

# Helper: load all listed packages silently
load_packages <- function(packages) {
  lapply(packages, library, character.only = TRUE)
}

# Execute: install missing, then load all
install_if_missing(required_packages)
load_packages(required_packages)
```


## Set project root

We use the `here` package to anchor all file paths relative to the project root, making this script portable across machines and operating systems.

```{r set-project-root, message=FALSE, warning=FALSE, include=FALSE}
# Declares this file's location so here() resolves paths from the project root
here::i_am("PRG/Data_preparation.Rmd")  # Set the root of the project
```


## Define directory paths
We define variables for the key directories (`DAT`, `PRG`, `RES`) using `here()` to ensure all file paths are relative to the project root, enhancing reproducibility and portability.
```{r define-paths}
# Project-relative paths to each main folder
# Using here() ensures paths work on any OS and any user's machine
path_dat <- here("DAT")         # Raw and processed data tables
path_prg <- here("PRG")         # R scripts and R Markdown files
path_res <- here("RES")         # Output results (figures, tables, model outputs)
```

## Define global color palette

We override the default ggplot2 discrete color scales globally to use the `viridis` palette, which is color-blind friendly and prints well in greyscale.


```{r set-color-palette, warning=TRUE, include=FALSE}
# Applies viridis to all discrete color and fill scales throughout the analysis
scale_color_discrete <- function(...) scale_color_viridis_d(...)
scale_fill_discrete <- function(...) scale_fill_viridis_d(...)
```

---

# Load custom functions

The following custom functions, stored in `PRG/Functions/`, are sourced here and will be available throughout this and all downstream scripts. Refer to each individual `.R` file for full documentation, parameters, and usage examples.

| File | Purpose |
|---|---|
| `functions_participant_period.R` | Assign participant IDs and day periods (AM/PM/NI) |
| `functions_cleaning_data.R` | Data cleaning and validation helpers |
| `functions_adjusted_ref_time.R` | Adjust timestamps to reference time windows |
| `functions_descriptive_stats.R` | Summary statistics tables |
| `functions_descriptive_plot.R` | Descriptive visualization helpers |
| `functions_linear_models.R` | Wrappers for linear mixed models |
| `residual_check_mmrm.R` | Residual diagnostics for MMRM models |
| `functions_check_clmm.R` | Assumption checks for cumulative link mixed models |

> **Customization notes:**
> - To change heatwave dates or the number of heatwave periods, see the [repository README](https://github.com/Souilla-Luc/Montreal_Summer_2024).
> - To change the time boundaries for AM / PM / NI periods, see the same README.


## Functions used for this analysis
```{r source-functions} 
# Source all custom function scripts from the Functions subfolder
source(here("PRG/Functions/functions_participant_period.R"))
source(here("PRG/Functions/functions_cleaning_data.R"))
source(here("PRG/Functions/functions_adjusted_ref_time.R"))
source(here("PRG/Functions/functions_descriptive_stats.R"))
source(here("PRG/Functions/functions_descriptive_plot.R"))
source(here("PRG/Functions/functions_linear_models.R"))
source(here("PRG/Functions/residual_check_mmrm.R"))
source(here("PRG/Functions/functions_check_clmm.R"))
```
---

# Study design definitions

## Period of day

Measurements were collected three times daily. Time-of-day periods are defined as:

| Period | Label | Time window |
|---|---|---|
| Morning | AM | 03:00 – 11:59 |
| Afternoon | PM | 12:00 – 17:59 |
| Night | NI | 18:00 – 02:59 (next day) |

## Heatwave periods

The 2024 Montreal heatwave study covers three consecutive periods:

| Period | Dates |
|---|---|
| Before heatwave | 2024-06-11 to 2024-06-17 |
| During heatwave | 2024-06-18 to 2024-06-20 |
| After heatwave  | 2024-06-21 to 2024-06-27 |

## Data processing pipeline

Each physiological and perceptual outcome follows a consistent three-step pipeline:

1. **`outcome_data_raw`** — upload, create, and initial preparation from DAT/
2. **`outcome_data_participant`** — processing, cleaning, and filtering
3. **`final_outcome_data`** — ready for statistical analysis and saved in the DAT/ processed folder

---

# Participant ID assignment

## Strategy for smartwatch (bangle) matching

Each participant was assigned a smartwatch identified by a unique **macID**. However, two edge cases complicate the matching:

- A participant may have used **two different smartwatches** (e.g., if the first one broke).
- The **same smartwatch** may have been used by two different participants across the study period.

To resolve this, we match each smartwatch's macID **together with the local timestamp** of its data against each participant's recorded start and end dates of use.

> **Important:** Before running this script, verify that all macIDs in `characteristics_participants.csv` are entered correctly. Typos in macIDs will result in unmatched records.
> Note: Some participants may not have a bangle macID in the baseline because they dropped out after the first familiarization week.

## Load baseline participant data

### Charging baseline data to get macID in order to match mac ID of the smartwatch with the participant ID. 
```{r load-baseline}
# Load the raw participant characteristics table
# This file contains macIDs and date ranges for each smartwatch assigned
baseline_data <- read.csv(file.path(path_dat, "characteristics_participants.csv"))
```


## Reshape to one row per participant per smartwatch

The raw data has one row per participant with up to three bangle columns (Bangle1, Bangle2, Bangle3). We pivot to long format so each participant-smartwatch combination occupies its own row.
```{r}
# Pivot wide bangle columns to long format: one row per participant per bangle
baseline_long <- baseline_data %>%
  pivot_longer(
    cols         = matches("^Bangle[1-3]"),   # Select columns starting with Bangle1/2/3
    names_to     = c("bangle", ".value"),      # Split into bangle number and variable name
    names_pattern = "Bangle(\\d+)(MacAdress|Date_Start|Date_End)"  # Regex to parse column names
  )

# Filter out rows with no macID (participants without a second or third bangle)
# and convert date strings to Date objects using day-month-year format
baseline_long <- baseline_long %>%
  filter(!is.na(MacAdress)) %>%
  mutate(
    date_start = dmy(Date_Start),   # Convert Date_Start string to Date
    date_end   = dmy(Date_End)      # Convert Date_End string to Date
  ) %>%
  select(
    participant_id,
    macID      = MacAdress,          # Rename for consistency across scripts
    date_start,
    date_end
  )
```

## Export baseline_long


```{r export-baseline-long}
# Export to DAT/ for reuse in all downstream analysis scripts
# row.names = FALSE avoids writing an unwanted index column
write.csv(baseline_long, file.path(path_dat, "baseline_long.csv"), row.names = FALSE)
```

---

# Environmental node ID assignment

## Reshape to one row per participant per node

Environmental nodes (indoor sensors) are handled similarly to smartwatches. Each participant has at most one node, identified by `NodeMacAdress`.

```{r reshape-baseline-nodes}
# Filter to participants who were assigned an environmental node
# and convert date strings to Date objects for time-based matching
baseline_nodes <- baseline_data %>%
  filter(!is.na(NodeMacAdress) & NodeMacAdress != "") %>%
  mutate(
    date_start = dmy(Node1Date_Start),   # Start date of node use
    date_end   = dmy(Node1Date_End)      # End date of node use
  ) %>%
  select(
    participant_id,
    macID      = NodeMacAdress,           # Rename for consistency with baseline_long
    date_start,
    date_end
  )
```


## Export baseline_nodes

```{r export-baseline-nodes}
# Export to DAT/ for reuse in environmental analysis scripts
write.csv(baseline_nodes, file.path(path_dat, "baseline_nodes.csv"), row.names = FALSE)
```

---

# Session information

Recording the R session ensures full reproducibility: anyone rerunning this script can verify they are using identical package versions.

```{r session-info}
sessionInfo()
```
